---
title: "Fig 1B — Multi-tissue UMAP"
date: today
date-format: "DD MMM YYYY"
title-block-banner: true
author: 
  - name: Chen Zhan
    orcid: 0000-0002-4794-8339
    email: chen.zhan@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
format: 
  html:
    theme: minty
    df-print: paged
    code-line-numbers: true
    embed-resources: true
    code-fold: true
    code-fold-default: true
    toc: true
    toc-location: left
    toc-title: "Menu"
knitr:
  opts_chunk: 
    message: false
    warning: false
comments:
  hypothesis: 
    theme: clean
editor: source
abstract: |
  Fig 1B. Two multi-tissue UMAPs built on immune-cell pseudobulk expression after
  covariate adjustment. The tissue UMAP preserves tissue and cell-type effects; the
  age UMAP preserves age and cell-type effects. Both use adjusted counts patched with
  zero for genes absent in a given cell type.
---

# Library

```{r setup-libraries}
library(tidySummarizedExperiment)
library(HDF5Array)
library(magrittr)
library(targets)
library(tidyverse)
library(tidybulk)
library(Seurat)
library(tidyseurat)
library(ggrepel)
```

# Cell types

```{r cell-types}
cell_type_list <- c(
  "plasma",
  "nk",
  "cd4 naive",
  "cd8 naive",
  "cd4 th1/th17 em",
  "cd4 th17 em",
  "mait",
  "tgd",
  "cd4 fh em",
  "cd4 th2 em",
  "cd8 tem",
  "cd4 tcm",
  "b memory",
  "b naive",
  "macrophage",
  "cd8 tcm",
  "treg",
  "monocytic"
)
```

# Colour palettes

```{r palettes}
# Cell-type palette: families of related hues
cell_type_colors <- c(
  # B lineage (reds)
  "b naive"   = "#e41a1c",
  "b memory"  = "#fb6a4a",
  "plasma"    = "#fc9272",
  # CD4 lineage (yellows/oranges)
  "cd4 naive"       = "#ffd92f",
  "cd4 tcm"         = "#ffe34d",
  "cd4 th1/th17 em" = "#fdce61",
  "cd4 th17 em"     = "#f4ed43",
  "cd4 fh em"       = "#fee08b",
  "cd4 th2 em"      = "#fdbe61",
  "treg"            = "#e6ab02",
  # CD8 lineage (blues)
  "cd8 naive" = "#377eb8",
  "cd8 tcm"   = "#6baed6",
  "cd8 tem"   = "#08519c",
  # NK / innate-like (greens/teals)
  "nk"   = "#1b9e77",
  "mait" = "#66c2a5",
  "tgd"  = "#41ae76",
  # Myeloid (browns)
  "macrophage" = "#a65628",
  # Monocytic (purples)
  "monocytic"  = "#984ea3"
)

# Tissue-group palette (blood drawn in grey, see plotting chunks)
tissue_group_colors <- c(
  # Immune / lymphoid (reds)
  "bone marrow"      = "#fb6a4a",
  "spleen"           = "#fcbba1",
  "thymus"           = "#fee0d2",
  "lymphatic system" = "#fc9272",
  # Digestive system (oranges)
  "digestive system (general)" = "#fd8d3c",
  "large intestine"            = "#fdae6b",
  "small intestine"            = "#fdd0a2",
  "oesophagus"                 = "#f16913",
  "stomach"                    = "#d94801",
  "pancreas"                   = "#f6c18a",
  "liver"                      = "#e6550d",
  # Respiratory system (turquoise / cyan)
  "respiratory system" = "#1f9bcf",
  "trachea"            = "#41b6c4",
  # Nervous system (blues)
  "brainstem and cerebellar structures" = "#2171b5",
  "cerebral lobes and cortical areas"   = "#6baed6",
  "general brain and major structures"  = "#9ecae1",
  "limbic and basal systems"            = "#c6dbef",
  "renal system"                        = "#08306b",
  # Reproductive (pinks / purples)
  "female reproductive system" = "#ce1256",
  "ovary"                      = "#e7298a",
  "prostate"                   = "#f768a1",
  # Endocrine & glands (purples)
  "endocrine system"     = "#756bb1",
  "miscellaneous glands" = "#9e9ac8",
  # Musculoskeletal & integumentary (greens)
  "muscular system (skeletal muscles)" = "#31a354",
  "integumentary system (skin)"        = "#74c476",
  "cardiovascular system"              = "#a1d99b",
  "vasculature"                        = "#006d2c",
  # Sensory / special (lavender / violet)
  "sensory-related structures"          = "#756bb1",
  "nasal, oral, and pharyngeal regions" = "#9ebcda",
  # Other
  "adipose tissue"                             = "#ff7f00",
  "peritoneal and abdominal cavity structures" = "grey40",
  "breast"                                      = "#f781bf"
)
```

# Adjusted counts

The adjusted counts are produced by the per-cell-type `targets` pipeline
(`NEW_CELL_TYPE/V1_<cell type>`). Two adjustments are extracted for each cell type:

* `adjust_tissue` — expression adjusted to preserve **tissue + cell-type** effects.
* `adjust_age` — expression adjusted to preserve **age + cell-type** effects.

## Compute from the targets pipeline

This chunk is disabled by default (`eval: false`); it documents how the adjusted-count
files read below were generated. Enable it to recompute from the `targets` stores.

```{r compute-adjusted-counts}
#| eval: false

data_dir  <- "~/lab/chen/HPC_posterior/Tissue_umap"
root_path <- "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/HPC_posterior/NEW_CELL_TYPE"

# Read one adjusted-count target per cell type, cap extreme values and floor at 0.
read_adjusted <- function(target_contains, estimate_nested) {
  map(cell_type_list, .f = ~ {
    message(.x)

    target_name <- tar_meta(
      contains(target_contains),
      store = glue::glue("{root_path}/V1_{.x %>% make.names}/_targets/")
    ) %>% filter(type == "pattern") %>% pull(name) %>% unique()

    target_value <- tar_read_raw(
      target_name,
      store = glue::glue("{root_path}/V1_{.x %>% make.names}/_targets/")
    )

    m <-
      target_value %>%
      select(.feature, {{ estimate_nested }}) %>%
      unnest({{ estimate_nested }}) %>%
      pivot_wider(names_from = sample_id, values_from = adjusted___Estimate) |>
      tidybulk:::as_matrix(rownames = ".feature") |>
      as("sparseMatrix") |>
      Matrix::Matrix(sparse = TRUE)

    # Cap infinite
    max_rm_infinite =
      m |>
      _[!m |> is.infinite()] |>
      quantile(0.999)

    m |>
      _[m > max_rm_infinite] =
      max_rm_infinite

    m |>
      _[m < 0] =
      0

    return(m)
  }) %>% set_names(cell_type_list)
}

adjust_tissue_list <- read_adjusted("adjust_tissue", brms_fit_adjusted_tissue_estimate)
saveRDS(adjust_tissue_list, file = file.path(data_dir, "adjust_tissue_list.rds"))

adjust_age_list <- read_adjusted("adjust_age", brms_fit_adjusted_age_estimate)
saveRDS(adjust_age_list, file = file.path(data_dir, "adjust_age_list.rds"))
```

## Read pre-computed adjusted counts

```{r read-adjusted-counts}
data_dir <- "~/lab/chen/HPC_posterior/Tissue_umap"

adjust_tissue_list <- readRDS(file.path(data_dir, "adjust_tissue_list.rds"))
adjust_age_list    <- readRDS(file.path(data_dir, "adjust_age_list.rds"))

names(adjust_tissue_list) <- cell_type_list
names(adjust_age_list)    <- cell_type_list
```

## Shared helpers and sample metadata

```{r shared-helpers}
# Update these paths to your data locations
hdf5_path            <- "~/lab/Mangiola_ImmuneAtlas/taskforce_shared_folder/zenodo/pseudobulk_se.h5ad"
edit_covariates_path <- "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/immuneHealthyBodyMap/rebuttal_CellPress/edit_covariates.R"

se  <- zellkonverter::readH5AD(hdf5_path, use_hdf5 = TRUE, reader = "R")
source(edit_covariates_path)

sample_meta <-
  se %>%
  colData() %>%
  as_tibble() %>%
  mutate(cell_type = cell_type_unified_ensemble) %>%
  edit_covariates() %>%
  distinct(sample_id, tissue_groups, age_years, age_decade)

rm(se)
gc()

# Union of genes across cell types (genes absent in a cell type become NA)
build_union_matrix <- function(adjust_list) {
  adjust_list %>%
    .[names(adjust_list) %>% setdiff("cytotoxic")] %>%
    reduce(.f = ~ full_join(
      .x %>% as.matrix() %>% as.data.frame() %>% rownames_to_column("rowname"),
      .y %>% as.matrix() %>% as.data.frame() %>% rownames_to_column("rowname"),
      by = "rowname"
    ) %>% column_to_rownames("rowname"))
}

# Build sample x cell-type metadata for a patched matrix
build_meta <- function(mat) {
  tibble(name = colnames(mat)) %>%
    tidyr::extract(
      name,
      into   = c("sample_id", "cell_type"),
      regex  = "^(.*)_{3,}(.+)$",
      remove = FALSE
    ) %>%
    left_join(sample_meta, by = "sample_id") %>%
    column_to_rownames("name")
}

# Cell-type label positions (median UMAP coordinate per cell type)
cell_type_labels <- function(umap_df) {
  umap_df %>%
    group_by(cell_type) %>%
    summarise(
      umap_1 = median(umap_1, na.rm = TRUE),
      umap_2 = median(umap_2, na.rm = TRUE),
      .groups = "drop"
    )
}
```

# Tissue UMAP

Preserves **tissue + cell-type** effects. Union set of genes, adjusted counts patched
with zero.

## Patch with zero + metadata

```{r tissue-prepare}
adjust_tissue_matrix         <- build_union_matrix(adjust_tissue_list)
adjust_tissue_matrix[is.na(adjust_tissue_matrix)] <- 0

meta_tissue <- build_meta(adjust_tissue_matrix)
```

## Seurat pipeline + UMAP

```{r tissue-seurat}
options(future.globals.maxSize = 20 * 1024^3)  # 20 GB

sobj_tissue <-
  CreateSeuratObject(counts = adjust_tissue_matrix, meta.data = meta_tissue) %>%
  SCTransform(assay = "RNA", vst.flavor = "v2", verbose = FALSE) %>%
  RunPCA() %>%
  RunUMAP(reduction = "pca", dims = 1:30, reduction.name = "umap")
```

## Plot

```{r tissue-plot, fig.width=9, fig.height=9}
umap_df_tissue <- FetchData(
  sobj_tissue,
  vars = c("umap_1", "umap_2", "tissue_groups", "cell_type")
)

label_df_tissue <- cell_type_labels(umap_df_tissue)

Plot_umap_tissue <-
  ggplot(umap_df_tissue, aes(x = umap_1, y = umap_2)) +
  ggrastr::rasterise(
    geom_point(
      data  = subset(umap_df_tissue, tissue_groups == "blood"),
      color = "grey", size = 0.05, alpha = 0.5, stroke = 0
    ),
    dpi = 500, scale = 1
  ) +
  ggrastr::rasterise(
    geom_point(
      data = subset(umap_df_tissue, tissue_groups != "blood"),
      aes(color = tissue_groups), size = 0.1, stroke = 0
    ),
    dpi = 500, scale = 1
  ) +
  geom_text_repel(
    data = label_df_tissue,
    aes(x = umap_1, y = umap_2, label = cell_type),
    inherit.aes = FALSE,
    size = 3, fontface = "bold", color = "black",
    box.padding = 0.4, point.padding = 0.2,
    max.overlaps = Inf, seed = 123
  ) +
  scale_color_manual(values = tissue_group_colors) +
  labs(x = "UMAP 1", y = "UMAP 2") +
  theme_void() +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank())

print(Plot_umap_tissue)

saveRDS(Plot_umap_tissue, file.path(data_dir, "Plot___umap_tissue_fig1.rds"))
```

```{r tissue-cleanup}
rm(sobj_tissue, adjust_tissue_matrix)
gc()
```

# Age UMAP

Preserves **age + cell-type** effects. Union set of genes, adjusted counts patched
with zero.

## Patch with zero + metadata

```{r age-prepare}
adjust_age_matrix         <- build_union_matrix(adjust_age_list)
adjust_age_matrix[is.na(adjust_age_matrix)] <- 0

meta_age <- build_meta(adjust_age_matrix)
```

## Seurat pipeline + UMAP

```{r age-seurat}
options(future.globals.maxSize = 20 * 1024^3)  # 20 GB

sobj_age <-
  CreateSeuratObject(counts = adjust_age_matrix, meta.data = meta_age) %>%
  SCTransform(assay = "RNA", vst.flavor = "v2", verbose = FALSE) %>%
  RunPCA() %>%
  RunUMAP(reduction = "pca", dims = 1:20, n.neighbors = 80, min.dist = 0.5)
```

## Plot

```{r age-plot, fig.width=9, fig.height=9}
umap_df_age <- FetchData(
  sobj_age,
  vars = c("umap_1", "umap_2", "age_decade", "cell_type")
) %>%
  mutate(age_decade = as.numeric(age_decade))

label_df_age <- cell_type_labels(umap_df_age)

Plot_umap_age <-
  ggplot(umap_df_age, aes(x = umap_1, y = umap_2)) +
  ggrastr::rasterise(
    geom_point(aes(color = age_decade), size = 0.05, shape = ".", alpha = 0.8, stroke = 0),
    dpi = 500, scale = 1
  ) +
  geom_text_repel(
    data = label_df_age,
    aes(x = umap_1, y = umap_2, label = cell_type),
    inherit.aes = FALSE,
    size = 3, fontface = "bold", color = "black",
    box.padding = 0.4, point.padding = 0.2,
    max.overlaps = Inf, seed = 123
  ) +
  scale_color_distiller(palette = "Spectral") +
  labs(x = "UMAP 1", y = "UMAP 2") +
  theme_void() +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank())

print(Plot_umap_age)

saveRDS(Plot_umap_age, file.path(data_dir, "Plot___umap_age_fig1.rds"))
```

```{r age-cleanup}
rm(sobj_age, adjust_age_matrix)
gc()
```

# Supplementary tables

Export UMAP coordinates and metadata backing Fig 1B panels.

```{r export-supplementary-tables}
proj_root <- Sys.getenv(
  "AGE_CLOCK_ROOT",
  unset = normalizePath("~/lab/chen/HPC_posterior/Age_Clock", mustWork = FALSE)
)
source(file.path(proj_root, "report", "R", "export_supplementary_tables.R"))

export_supplementary_workbook(
  plots = list(
    umap_tissue = Plot_umap_tissue,
    umap_age = Plot_umap_age
  ),
  output_path = file.path(proj_root, "Supplementary_files", "SupplData_AgeClock_Fig1.xlsx"),
  prefix = "Fig1_"
)
```
