---
title: "Figure 2 supplementary plots"
date: today
date-format: "DD MMM YYYY"
format:
  html:
    theme: minty
    toc: true
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
---

# Purpose

This report uses the finalized Figure 2 source-data workbook. It adds two
inferential views that are not visible in full in the main figure:

1. all body-level age effects on immune-cell composition with posterior intervals; and
2. the complete tissue-level myeloid-minus-lymphoid contrast underlying the
   anatogram.

The plasma-cell trajectory panel is not repeated because the main Figure 2
already displays every tissue meeting its age-coverage criterion.

```{r setup}
library(readxl)
library(dplyr)
library(ggplot2)
library(scales)
library(here)

source(file.path("..", "R", "paths.R"))
workbook <- path_source_tables("SupplData_AgeClock_Fig2.xlsx")
if (!file.exists(workbook)) stop("Figure 2 source-data workbook not found: ", workbook)

plot_object_dir <- file.path(here::here(), "vignettes", "supplementary_plot_objects")
dir.create(plot_object_dir, recursive = TRUE, showWarnings = FALSE)

read_panel_sheet <- function(sheet) {
  if (!sheet %in% excel_sheets(workbook)) stop("Missing worksheet: ", sheet)
  read_xlsx(workbook, sheet = sheet, guess_max = 1000000L)
}

save_plot_object <- function(plot, filename) {
  path <- file.path(plot_object_dir, filename)
  saveRDS(plot, path, compress = "xz")
  message("Saved ggplot object: ", path)
  invisible(path)
}

theme_supplementary <- theme_bw(base_size = 9) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    strip.background = element_blank(),
    legend.position = "bottom",
    plot.title.position = "plot"
  )

format_cell_type <- function(x) {
  out <- tools::toTitleCase(gsub("_", " ", x, fixed = TRUE))
  out <- gsub("^Cd", "CD", out)
  out <- gsub(" Mono$", " Mono", out)
  out <- gsub("^Nk$", "NK", out)
  out <- gsub("^Nkt$", "NKT", out)
  out <- gsub("^Ilc$", "ILC", out)
  out <- gsub("^Mait$", "MAIT", out)
  out <- gsub("^Tgd$", "TGD", out)
  out
}
```

```{r read-source-data}
composition_tests <- read_panel_sheet("Fig2_B_age_composition_tests")
lineage_tests <- read_panel_sheet("Fig2_C_lineage_and_contrast_tes")
```

# All body-level compositional age effects

The main boxplots show selected high-ranking enriched and depleted cell types.
This forest plot retains every tested cell type and therefore separates the
selection used for display from the complete inferential result.

```{r fig2-all-composition-forest, fig.width=8, fig.height=7}
composition_forest_data <- composition_tests |>
  mutate(
    cell_label = format_cell_type(L3),
    supported = !is.na(c_FDR) & c_FDR < 0.05
  ) |>
  arrange(c_effect) |>
  mutate(cell_label = factor(cell_label, levels = unique(cell_label)))

Plot_SFig2_all_composition_forest <-
  ggplot(composition_forest_data, aes(y = cell_label)) +
  geom_vline(xintercept = 0, colour = "grey55", linewidth = 0.35) +
  geom_segment(
    aes(x = c_lower, xend = c_upper, yend = cell_label, colour = supported),
    linewidth = 0.6
  ) +
  geom_point(aes(x = c_effect, fill = supported), shape = 21, size = 2.2) +
  scale_colour_manual(values = c(`TRUE` = "#B24A3B", `FALSE` = "grey65"), guide = "none") +
  scale_fill_manual(
    values = c(`TRUE` = "#B24A3B", `FALSE` = "white"),
    name = "FDR < 0.05"
  ) +
  labs(
    x = "Age-associated composition effect",
    y = NULL,
    title = "Complete body-level age effects on immune-cell composition",
    subtitle = "Points are posterior estimates; horizontal lines are posterior intervals"
  ) +
  theme_supplementary +
  theme(legend.position = "right")

Plot_SFig2_all_composition_forest
save_plot_object(
  Plot_SFig2_all_composition_forest,
  "Fig2_supp_all_composition_forest.rds"
)
```

# Tissue-level myeloid-minus-lymphoid contrast

This plot exposes the complete contrast table behind the anatomical summary.
Point size encodes contrast FDR. At present no tissue passes FDR < 0.05, so the
main anatogram should be interpreted as a descriptive spatial summary.

```{r fig2-lineage-contrast, fig.width=9, fig.height=8}
lineage_contrast_data <- lineage_tests |>
  distinct(tissue_groups, c_delta, contrast_FDR, significant) |>
  filter(!is.na(tissue_groups), is.finite(c_delta), is.finite(contrast_FDR)) |>
  mutate(
    direction = ifelse(c_delta >= 0, "Myeloid relative increase", "Lymphoid relative increase"),
    neg_log10_fdr = -log10(pmax(contrast_FDR, .Machine$double.xmin))
  ) |>
  arrange(c_delta) |>
  mutate(tissue_groups = factor(tissue_groups, levels = unique(tissue_groups)))

minimum_contrast_fdr <- min(lineage_contrast_data$contrast_FDR, na.rm = TRUE)

Plot_SFig2_lineage_contrast <-
  ggplot(lineage_contrast_data, aes(x = c_delta, y = tissue_groups)) +
  geom_vline(xintercept = 0, colour = "grey55", linewidth = 0.35) +
  geom_segment(
    aes(x = 0, xend = c_delta, yend = tissue_groups, colour = direction),
    linewidth = 0.45
  ) +
  geom_point(aes(size = neg_log10_fdr, fill = significant), shape = 21, colour = "black") +
  scale_colour_manual(
    values = c(
      "Myeloid relative increase" = "#B24A3B",
      "Lymphoid relative increase" = "#3B6FB6"
    ),
    name = "Contrast direction"
  ) +
  scale_fill_manual(values = c(`TRUE` = "black", `FALSE` = "white"), name = "FDR < 0.05") +
  scale_size_continuous(name = expression(-log[10](FDR)), range = c(1.5, 4.5)) +
  labs(
    x = "Myeloid-minus-lymphoid age contrast",
    y = NULL,
    title = "Tissue-level myeloid–lymphoid ageing contrast",
    subtitle = sprintf("No contrast passes FDR < 0.05; minimum FDR = %.3f", minimum_contrast_fdr)
  ) +
  theme_supplementary +
  theme(legend.position = "right")

Plot_SFig2_lineage_contrast
save_plot_object(
  Plot_SFig2_lineage_contrast,
  "Fig2_supp_lineage_contrast.rds"
)
```

# Output objects

```{r output-manifest}
data.frame(
  object = c(
    "Plot_SFig2_all_composition_forest",
    "Plot_SFig2_lineage_contrast"
  ),
  rds_file = c(
    "Fig2_supp_all_composition_forest.rds",
    "Fig2_supp_lineage_contrast.rds"
  )
)
```
