---
title: "Figure 2 supplementary plots"
date: today
date-format: "DD MMM YYYY"
format:
  html:
    theme: minty
    toc: true
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
---

# Purpose

This report expands panels that the main figure truncates for display:

1. every body-level compositional age effect (forest);
2. every tissue-level myeloid-minus-lymphoid contrast;
3. age-decade boxplots for **all** tested immune cell types (not only the ten
   highlighted in Fig. 2e); and
4. plasma-cell trajectories for **every** tissue with sccomp plasma
   predictions (main Fig. 2f keeps only tissues with >3 age decades).

On render, each ggplot object is written to
`vignettes/supplementary_plot_objects/`.

```{r setup}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(here)
library(forcats)
library(stringr)
library(ggh4x)
library(patchwork)

source(file.path("..", "R", "paths.R"))
workbook <- path_source_tables("Supplementary_File_Figure_2.xlsx")
if (!file.exists(workbook)) stop("Figure 2 supplementary file not found: ", workbook)

# Dropped machine-readable L3 tests (not in the journal file) for joining caches.
workbook_report <- path_source_tables("Supplementary_File_Figure_2_report_inputs.xlsx")

plot_object_dir <- file.path(publication_vignette_root(), "supplementary_plot_objects")
dir.create(plot_object_dir, recursive = TRUE, showWarnings = FALSE)

read_panel_sheet <- function(sheet, path = workbook) {
  if (!sheet %in% excel_sheets(path)) stop("Missing worksheet: ", sheet, " in ", path)
  # Submission / report-input workbooks store a caption in row 1 and headers in row 2.
  read_xlsx(path, sheet = sheet, skip = 1, guess_max = 1000000L)
}

save_plot_object <- function(plot, filename) {
  path <- file.path(plot_object_dir, filename)
  saveRDS(plot, path, compress = "xz")
  message("Saved ggplot object: ", path)
  invisible(path)
}

# Same asin-sqrt transform used in Fig2.qmd boxplots / plasma line chart.
asin_sqrt_trans <- scales::trans_new(
  name = "asin_sqrt",
  transform = function(x) asin(sqrt(x)),
  inverse = function(y) sin(y)^2
)

# Match Fig2.qmd theme_multipanel exactly for boxplots and line charts.
theme_multipanel <-
  theme_bw() +
  theme(
    panel.border = element_blank(),
    axis.line = element_line(size = 0.1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    strip.background = element_blank(),
    axis.title.y = element_text(margin = margin(t = 0, r = 0, b = 0, l = 0), size = 7),
    axis.title.x = element_text(margin = margin(t = 0, r = 0, b = 0, l = 0), size = 7),
    panel.spacing.x = unit(0.1, "lines"),
    axis.text.x = element_text(size = 6),
    axis.text.y = element_text(size = 6),
    strip.text.x = element_text(size = 7),
    strip.text.y = element_text(size = 7),
    legend.key.size = unit(5, "mm"),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 6),
    strip.clip = "off",
    plot.title = element_text(size = 7),
    axis.line.x = element_line(size = 0.2),
    axis.line.y = element_line(size = 0.2),
    axis.ticks.x = element_line(size = 0.2),
    axis.ticks.y = element_line(size = 0.2)
  )

# Lighter theme retained only for forest / contrast summary panels.
theme_supplementary <- theme_bw(base_size = 9) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    strip.background = element_blank(),
    legend.position = "bottom",
    plot.title.position = "plot"
  )

# Same abbreviations as Fig2.qmd boxplot strips.
cell_type_abbrev_dict <- c(
  "macrophage" = "Mac",
  "cd4 th1 em" = "CD4\nTh1\nEM",
  "cd4 th1/th17 em" = "CD4\nTh1/17 EM",
  "cd4 th17 em" = "CD4\nTh17 EM",
  "cd4 th2 em" = "CD4\nTh2 EM",
  "cd8 naive" = "CD8\nNaive",
  "plasma" = "Plasma",
  "granulocyte" = "Gran",
  "cd4 naive" = "CD4\nNaive",
  "b naive" = "B\nNaive",
  "nk" = "NK",
  "nkt" = "NKT",
  "cdc" = "CDC",
  "pdc" = "PDC",
  "b memory" = "B\nMemory",
  "cd14 mono" = "CD14\nMono",
  "cd16 mono" = "CD16\nMono",
  "cd4 fh em" = "CD4\nFH EM",
  "cd4 tcm" = "CD4\nTCM",
  "cd8 tcm" = "CD8\nTCM",
  "cd8 tem" = "CD8\nTEM",
  "ilc" = "ILC",
  "mait" = "MAIT",
  "mast" = "Mast",
  "tgd" = "TGD",
  "treg" = "Treg"
)

tissue_name_dict <- c(
  "blood" = "Blood",
  "sensory-related structures" = "Sensory",
  "cerebral lobes and cortical areas" = "Cerebral cortex",
  "limbic and basal systems" = "Limbic/basal",
  "liver" = "Liver",
  "bone marrow" = "Bone marrow",
  "large intestine" = "Large int",
  "lymphatic system" = "Lymphatic",
  "small intestine" = "Small int",
  "respiratory system" = "Respiratory",
  "renal system" = "Renal",
  "brainstem and cerebellar structures" = "Hindbrain",
  "cardiovascular system" = "Cardiovascular",
  "general brain and major structures" = "Brain",
  "vasculature" = "Vasculature",
  "ovary" = "Ovary",
  "integumentary system (skin)" = "Skin",
  "nasal, oral, and pharyngeal regions" = "Pharyngeal",
  "muscular system (skeletal muscles)" = "Skeletal muscle",
  "prostate" = "Prostate",
  "breast" = "Breast",
  "digestive system (general)" = "Digestive",
  "oesophagus" = "Oesophagus",
  "female reproductive system" = "Female repro",
  "spleen" = "Spleen",
  "peritoneal and abdominal cavity structures" = "Peritoneal",
  "thymus" = "Thymus",
  "adipose tissue" = "Adipose",
  "trachea" = "Trachea",
  "endocrine system" = "Endocrine",
  "pancreas" = "Pancreas",
  "miscellaneous glands" = "Glands",
  "stomach" = "Stomach",
  "gallbladder" = "Gallbladder"
)

resolve_fig2_cache <- function(filename) {
  candidates <- c(
    path_intermediate("fig2", filename),
    {
      arch <- path_age_clock_archive("data", "age_fig", filename)
      if (is.null(arch)) character() else arch
    }
  )
  hit <- candidates[nzchar(candidates) & file.exists(candidates)][1]
  if (is.na(hit)) {
    stop(
      "Missing Fig2 cache '", filename, "'. Expected under path_intermediate('fig2') ",
      "or AGE_CLOCK_ARCHIVE/data/age_fig/.",
      call. = FALSE
    )
  }
  hit
}
```

```{r read-source-data}
composition_tests <- read_panel_sheet("Fig2_B_composition_tests")
lineage_tests <- read_panel_sheet("Fig2_C_lineage_tests")
plasma_predictions <- read_panel_sheet("Fig2_F_plasma_predictions")
plasma_tests <- read_panel_sheet("Fig2_F_plasma_tests")
# Full tissue × cell-type age effects (includes plasma tissues excluded from
# Fig. 2f by the >3 age-decade coverage rule).
tissue_effects <- read_panel_sheet("Fig2_D_tissue_effects")

# L3-coded composition tests for joining prediction caches.
if (file.exists(workbook_report) &&
    "Fig2_B_composition_tests_L3" %in% excel_sheets(workbook_report)) {
  composition_tests_l3 <- read_panel_sheet(
    "Fig2_B_composition_tests_L3",
    path = workbook_report
  ) |>
    filter(L3 != "cytotoxic") |>
    transmute(
      L3,
      c_effect,
      c_FDR,
      significant = !is.na(c_FDR) & c_FDR < 0.05
    )
} else {
  stop(
    "Need machine-readable composition tests at ", workbook_report,
    " (sheet Fig2_B_composition_tests_L3) to label all-cell-type boxplots."
  )
}
```

# All body-level compositional age effects

The main boxplots show selected high-ranking enriched and depleted cell types.
This forest plot retains every tested cell type.

```{r fig2-all-composition-forest, fig.width=8, fig.height=7}
composition_forest_data <- composition_tests |>
  transmute(
    cell_label = .data[["Immune cell type"]],
    c_effect = .data[["Posterior mean effect (unconstrained sccomp composition scale)"]],
    c_lower = .data[["95% CrI lower"]],
    c_upper = .data[["95% CrI upper"]],
    supported = .data[["FDR < 0.05"]] == "Yes",
    displayed_main = .data[["Displayed in Fig. 2e"]] == "Yes"
  ) |>
  arrange(c_effect) |>
  mutate(cell_label = factor(cell_label, levels = unique(cell_label)))

Plot_SFig2_all_composition_forest <-
  ggplot(composition_forest_data, aes(y = cell_label)) +
  geom_vline(xintercept = 0, colour = "grey55", linewidth = 0.35) +
  geom_segment(
    aes(x = c_lower, xend = c_upper, yend = cell_label, colour = supported),
    linewidth = 0.6
  ) +
  geom_point(aes(x = c_effect, fill = supported, shape = displayed_main), size = 2.2) +
  scale_colour_manual(values = c(`TRUE` = "#B24A3B", `FALSE` = "grey65"), guide = "none") +
  scale_fill_manual(
    values = c(`TRUE` = "#B24A3B", `FALSE` = "white"),
    name = "FDR < 0.05"
  ) +
  scale_shape_manual(
    values = c(`TRUE` = 21, `FALSE` = 22),
    name = "Shown in Fig. 2e"
  ) +
  labs(
    x = "Age-associated composition effect",
    y = NULL,
    title = "Complete body-level age effects on immune-cell composition",
    subtitle = "Points are posterior estimates; horizontal lines are posterior intervals"
  ) +
  theme_supplementary +
  theme(legend.position = "right")

Plot_SFig2_all_composition_forest
save_plot_object(
  Plot_SFig2_all_composition_forest,
  "Fig2_supp_all_composition_forest.rds"
)
```

# Tissue-level myeloid-minus-lymphoid contrast

```{r fig2-lineage-contrast, fig.width=9, fig.height=8}
lineage_contrast_data <- lineage_tests |>
  distinct(
    tissue_groups = .data[["Tissue group"]],
    c_delta = .data[["Myeloid-minus-lymphoid change"]],
    contrast_FDR = .data[["Bayesian FDR (lineage contrast)"]],
    significant = .data[["Lineage contrast FDR < 0.05"]] == "Yes"
  ) |>
  filter(!is.na(tissue_groups), is.finite(c_delta), is.finite(contrast_FDR)) |>
  mutate(
    direction = ifelse(
      c_delta >= 0, "Myeloid relative increase", "Lymphoid relative increase"
    ),
    neg_log10_fdr = -log10(pmax(contrast_FDR, .Machine$double.xmin))
  ) |>
  arrange(c_delta) |>
  mutate(tissue_groups = factor(tissue_groups, levels = unique(tissue_groups)))

minimum_contrast_fdr <- min(lineage_contrast_data$contrast_FDR, na.rm = TRUE)

Plot_SFig2_lineage_contrast <-
  ggplot(lineage_contrast_data, aes(x = c_delta, y = tissue_groups)) +
  geom_vline(xintercept = 0, colour = "grey55", linewidth = 0.35) +
  geom_segment(
    aes(x = 0, xend = c_delta, yend = tissue_groups, colour = direction),
    linewidth = 0.45
  ) +
  geom_point(aes(size = neg_log10_fdr, fill = significant), shape = 21, colour = "black") +
  scale_colour_manual(
    values = c(
      "Myeloid relative increase" = "#B24A3B",
      "Lymphoid relative increase" = "#3B6FB6"
    ),
    name = "Contrast direction"
  ) +
  scale_fill_manual(values = c(`TRUE` = "black", `FALSE` = "white"), name = "FDR < 0.05") +
  scale_size_continuous(name = expression(-log[10](FDR)), range = c(1.5, 4.5)) +
  labs(
    x = "Myeloid-minus-lymphoid age contrast",
    y = NULL,
    title = "Tissue-level myeloid–lymphoid ageing contrast",
    subtitle = sprintf(
      "No contrast passes FDR < 0.05; minimum FDR = %.3f", minimum_contrast_fdr
    )
  ) +
  theme_supplementary +
  theme(legend.position = "right")

Plot_SFig2_lineage_contrast
save_plot_object(
  Plot_SFig2_lineage_contrast,
  "Fig2_supp_lineage_contrast.rds"
)
```

# Age-decade boxplots for every tested cell type

Main Fig. 2e keeps five enrichment and five depletion cell types. Below, every
non-cytotoxic cell type with a body-level composition test is shown, split by
effect direction and ordered by FDR within direction.

```{r fig2-all-boxplots-data}
pred_draws <- readRDS(resolve_fig2_cache("sccomp_age_pred_draws.rds"))
age_adjusted_path <- tryCatch(
  resolve_fig2_cache("sccomp_age_adjusted.rds"),
  error = function(e) NULL
)
age_adjusted <- NULL
if (!is.null(age_adjusted_path) && file.exists(age_adjusted_path)) {
  message("Loading age_adjusted cache: ", age_adjusted_path)
  age_adjusted <- readRDS(age_adjusted_path)
} else {
  message("age_adjusted cache missing; plotting prediction boxplots only.")
}

format_fdr <- function(x) {
  case_when(
    is.na(x) ~ "NA",
    x == 0 ~ "0",
    x < 0.001 ~ gsub("^0\\.", ".", formatC(x, format = "e", digits = 2)),
    TRUE ~ gsub("^0\\.", ".", formatC(x, format = "f", digits = 3))
  )
}

build_age_boxplot_data <- function(direction_sign) {
  selected <- composition_tests_l3 |>
    filter(sign(c_effect) == direction_sign) |>
    arrange(c_FDR, desc(abs(c_effect)))

  cell_types <- selected$L3
  if (length(cell_types) == 0) {
    stop("No cell types with effect sign ", direction_sign)
  }

  plot_df <- pred_draws |>
    filter(L3 %in% cell_types) |>
    transmute(
      sample_id, L3,
      age_decade = as.character(age_decade),
      proportion_mean,
      version = "prediction"
    )

  if (!is.null(age_adjusted)) {
    prop_col <- if ("proportion_mean" %in% names(age_adjusted)) {
      "proportion_mean"
    } else if ("proportion" %in% names(age_adjusted)) {
      "proportion"
    } else {
      stop("age_adjusted cache lacks proportion columns")
    }
    adj_part <- age_adjusted |>
      filter(L3 %in% cell_types) |>
      group_by(sample_id, L3, age_decade) |>
      summarise(proportion_mean = mean(.data[[prop_col]], na.rm = TRUE), .groups = "drop") |>
      mutate(
        age_decade = as.character(age_decade),
        version = "adjusted"
      )
    plot_df <- bind_rows(plot_df, adj_part)
  }

  plot_df |>
    left_join(selected, by = "L3") |>
    mutate(
      L3_abbrev_base = recode(L3, !!!cell_type_abbrev_dict),
      L3_abbrev_base = if_else(
        L3_abbrev_base == L3,
        tools::toTitleCase(gsub("_", " ", L3, fixed = TRUE)),
        L3_abbrev_base
      ),
      L3_abbrev = paste0(L3_abbrev_base, "\n(FDR: ", format_fdr(c_FDR), ")"),
      age_decade = factor(age_decade, levels = as.character(1:10)),
      L3 = factor(L3, levels = cell_types)
    ) |>
    arrange(L3) |>
    mutate(L3_abbrev = factor(L3_abbrev, levels = unique(as.character(L3_abbrev))))
}

build_age_boxplot <- function(df, fill_prediction = NULL) {
  # Mirror Fig2.qmd Plot__boxplot_up/down_cell_type styling, with fixed 5 columns.
  df <- mutate(df, age_decade = factor(age_decade, levels = as.character(1:10)))
  has_adjusted <- any(df$version == "adjusted")

  p <- ggplot(df, aes(x = age_decade, y = proportion_mean))
  if (has_adjusted) {
    p <- p +
      ggbeeswarm::geom_quasirandom(
        data = df %>% dplyr::filter(version == "adjusted"),
        width = 0.3,
        colour = "grey60",
        shape = ".",
        size = 0.15,
        alpha = 0.6
      )
  }
  p <- p +
    geom_boxplot(
      aes(fill = version),
      data = df %>% dplyr::filter(version == "prediction"),
      outlier.shape = NA,
      alpha = 0.4,
      linewidth = 0.15,
      width = 0.7
    ) +
    scale_y_continuous(
      trans = asin_sqrt_trans,
      expand = expansion(mult = c(0, 0.05)),
      labels = function(x) gsub("^0\\.", ".", format(x, nsmall = 1))
    ) +
    scale_x_discrete(labels = function(x) x) +
    guides(fill = "none") +
    facet_wrap(~ L3_abbrev, ncol = 5, scales = "free_y") +
    labs(title = "", x = "Age Decade", y = "Proportion Mean (arcsin-sqrt scale)") +
    theme_multipanel +
    theme(
      strip.text.x.top = element_text(margin = margin(0, 0, 0, 0, "mm")),
      strip.placement = "inside",
      panel.spacing.x = unit(0, "mm"),
      panel.spacing.y = unit(0, "mm"),
      legend.position = "none"
    )

  if (!is.null(fill_prediction)) {
    p <- p + scale_fill_manual(values = c("prediction" = fill_prediction))
  }
  p
}

df_up_all <- build_age_boxplot_data(1)
df_down_all <- build_age_boxplot_data(-1)
# Free large cache before plotting.
rm(age_adjusted)
invisible(gc())
```

```{r fig2-all-boxplots-up, fig.width=14, fig.height=8}
Plot_SFig2_all_boxplot_enrichment <- build_age_boxplot(df_up_all)
Plot_SFig2_all_boxplot_enrichment
save_plot_object(
  Plot_SFig2_all_boxplot_enrichment,
  "Fig2_supp_all_boxplot_enrichment.rds"
)
```

```{r fig2-all-boxplots-down, fig.width=14, fig.height=8}
# Same light-blue fill used for the downregulated panel in Fig2.qmd.
Plot_SFig2_all_boxplot_depletion <- build_age_boxplot(
  df_down_all,
  fill_prediction = "#ADD8E6"
)
Plot_SFig2_all_boxplot_depletion
save_plot_object(
  Plot_SFig2_all_boxplot_depletion,
  "Fig2_supp_all_boxplot_depletion.rds"
)
```

```{r fig2-all-boxplots-combined, fig.width=14, fig.height=16}
# Single SI figure: A = age enrichment, B = age depletion.
Plot_SFig2_all_boxplot <-
  (Plot_SFig2_all_boxplot_enrichment / Plot_SFig2_all_boxplot_depletion) +
  plot_annotation(tag_levels = "A") &
  theme(
    plot.tag = element_text(size = 9, face = "bold"),
    plot.margin = margin(2, 2, 2, 2, "mm")
  )
Plot_SFig2_all_boxplot
save_plot_object(Plot_SFig2_all_boxplot, "Fig2_supp_all_boxplot.rds")
ggsave(
  file.path(plot_object_dir, "Fig2_supp_all_boxplot.png"),
  Plot_SFig2_all_boxplot,
  width = 183,
  height = 240,
  units = "mm",
  dpi = 500,
  bg = "white"
)
```

# Plasma trajectories for every tissue with predictions

Main Fig. 2f keeps tissues with plasma predictions in **more than three** age
decades. This panel drops that coverage filter and plots every tissue present in
the sccomp plasma prediction cache. Significance colouring (red = Bayesian
FDR < 0.05) uses the tissue-level plasma tests from `Fig2_D_tissue_effects`.
Styling matches `Plot__plasma_line_chart` in Fig2.qmd.

```{r fig2-all-plasma-line, fig.width=15, fig.height=14}
# Full prediction cache (main figure applies tissue_groups_of_interest: n > 3).
sccomp_prediction_path <- resolve_fig2_cache(
  "sccomp_prediction____age_decade___L3___disease_FALSE___immune_only_TRUE.rds"
)
sccomp_df <- readRDS(sccomp_prediction_path)

# Display-name → model-key helpers (workbook omits the Oxford comma / uses Skin).
normalise_tissue_key <- function(x) {
  key <- str_to_lower(as.character(x))
  key <- if_else(
    key == "nasal, oral and pharyngeal regions",
    "nasal, oral, and pharyngeal regions",
    key
  )
  if_else(key == "skin", "integumentary system (skin)", key)
}

plasma_test_lookup <- tissue_effects |>
  filter(.data[["Immune cell type"]] == "Plasma cells") |>
  transmute(
    tissue_key = normalise_tissue_key(.data[["Tissue group"]]),
    c_FDR = .data[["Bayesian FDR"]],
    significant = .data[["FDR < 0.05"]] == "Yes"
  )

# Main Fig. 2f coverage rule (for annotation only).
main_fig_tissues <- sccomp_df |>
  filter(cell_type_unified_ensemble == "plasma") |>
  distinct(tissue_groups, age_decade) |>
  count(tissue_groups, name = "n_decades") |>
  filter(n_decades > 3) |>
  pull(tissue_groups)

plasma_line_data <- sccomp_df |>
  filter(cell_type_unified_ensemble == "plasma") |>
  mutate(age_decade = as.character(age_decade)) |>
  group_by(tissue_groups, age_decade) |>
  summarise(
    n_source_rows = n(),
    n_samples = n_distinct(sample_id),
    proportion_mean = first(proportion_mean),
    proportion_lower = first(proportion_lower),
    proportion_upper = first(proportion_upper),
    .groups = "drop"
  ) |>
  mutate(
    tissue_key = normalise_tissue_key(tissue_groups),
    in_main_fig2f = tissue_groups %in% main_fig_tissues
  ) |>
  left_join(plasma_test_lookup, by = "tissue_key") |>
  mutate(
    age_decade = factor(age_decade, levels = as.character(1:10)),
    significant = replace_na(significant, FALSE),
    tissue_short = unname(tissue_name_dict[tissue_key]),
    tissue_short = if_else(is.na(tissue_short), as.character(tissue_groups), tissue_short),
    tissue_short = factor(
      tissue_short,
      levels = c(setdiff(unique(tissue_short), "Large int"), "Large int")
    )
  )

message(
  "Plasma supplementary tissues: ",
  n_distinct(plasma_line_data$tissue_groups),
  " (main Fig. 2f coverage filter keeps ",
  length(main_fig_tissues),
  "). Added: ",
  paste(
    setdiff(unique(plasma_line_data$tissue_groups), main_fig_tissues),
    collapse = ", "
  )
)

Plot_SFig2_all_plasma_line <-
  plasma_line_data |>
  mutate(
    significant = as.character(significant),
    tissue_groups_short = tissue_short
  ) |>
  ggplot() +
  # Match Fig2.qmd Plot__plasma_line_chart layer order and aesthetics.
  geom_point(aes(x = age_decade, y = proportion_mean), shape = NA) +
  geom_line(
    aes(x = as.numeric(age_decade), y = proportion_mean, colour = significant),
    show.legend = FALSE,
    linewidth = 0.15
  ) +
  geom_ribbon(
    aes(
      x = as.numeric(age_decade),
      ymin = proportion_lower,
      ymax = proportion_upper
    ),
    colour = NA,
    alpha = 0.1,
    fill = "grey80"
  ) +
  scale_size_continuous(range = c(0.5, 2), guide = guide_legend(title = "-log10(FDR)")) +
  scale_y_continuous(
    trans = asin_sqrt_trans,
    expand = expansion(mult = c(0, 0.05)),
    labels = function(x) gsub("^0\\.", ".", format(x, nsmall = 1))
  ) +
  scale_colour_manual(
    values = c("TRUE" = "#B24A3B", "FALSE" = "black"),
    guide = guide_legend(override.aes = list(alpha = 1))
  ) +
  labs(x = "Age decade", y = "Proportion Mean (arcsin-sqrt scale)") +
  ggh4x::facet_wrap2(
    ~ tissue_groups_short,
    ncol = 5,
    scales = "free_y",
    axes = "margins"
  ) +
  theme_multipanel +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 0, hjust = 1)
  )

Plot_SFig2_all_plasma_line
save_plot_object(
  Plot_SFig2_all_plasma_line,
  "Fig2_supp_all_plasma_line.rds"
)
ggsave(
  file.path(plot_object_dir, "Fig2_supp_all_plasma_line.png"),
  Plot_SFig2_all_plasma_line,
  width = 183,
  height = 220,
  units = "mm",
  dpi = 500,
  bg = "white"
)
```

# Output objects

```{r output-manifest}
data.frame(
  object = c(
    "Plot_SFig2_all_composition_forest",
    "Plot_SFig2_lineage_contrast",
    "Plot_SFig2_all_boxplot_enrichment",
    "Plot_SFig2_all_boxplot_depletion",
    "Plot_SFig2_all_boxplot",
    "Plot_SFig2_all_plasma_line"
  ),
  rds_file = c(
    "Fig2_supp_all_composition_forest.rds",
    "Fig2_supp_lineage_contrast.rds",
    "Fig2_supp_all_boxplot_enrichment.rds",
    "Fig2_supp_all_boxplot_depletion.rds",
    "Fig2_supp_all_boxplot.rds",
    "Fig2_supp_all_plasma_line.rds"
  )
)
```
