---
title: "Figure 2 — Asynchronous ageing of immune-cell composition across tissues"
date: today
date-format: "DD MMM YYYY"
title-block-banner: true
author: 
  - name: Chen Zhan
    orcid: 0000-0002-4794-8339
    email: chen.zhan@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
format: 
  html:
    theme: minty
    df-print: paged
    code-line-numbers: true
    embed-resources: true
    code-fold: true
    code-fold-default: true
    toc: true
    toc-location: left
    toc-title: "Menu"
knitr:
  opts_chunk: 
    message: false
    warning: false
comments:
  hypothesis: 
    theme: clean
editor: source
abstract: |
  Immune ageing is asynchronous across tissues.
---

# Library


```{r publication-paths}
# Portable publication data roots (no HPC absolute paths).
.path_candidates <- c(
  file.path("..", "R", "paths.R"),
  if (requireNamespace("here", quietly = TRUE)) {
    file.path(here::here(), "vignettes", "R", "paths.R")
  },
  file.path("R", "paths.R")
)
.path_candidates <- .path_candidates[!is.na(.path_candidates) & nzchar(.path_candidates)]
.path_file <- .path_candidates[file.exists(.path_candidates)][1]
if (is.na(.path_file)) {
  stop(
    "Cannot find vignettes/R/paths.R. Render from the immuneHealthyBodyMap ",
    "repository root or from vignettes/reports/, or set IMMUNE_HEALTHY_BODY_MAP_ROOT."
  )
}
source(.path_file)
fig_cache_dir <- path_figure_ready()
dir.create(path_intermediate("fig2"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_intermediate("fig4"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_intermediate("fig5"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_figure_ready("fig1"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_figure_ready("fig2"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_figure_ready("fig4"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_figure_ready("fig5"), recursive = TRUE, showWarnings = FALSE)
```

```{r setup-libraries}
# Set your R library path here if needed
# .libPaths removed for portability; use renv/conda or site library
# Ensure required directories exist
required_dirs <- c(
  path_figure_ready("fig2"),
  path_intermediate("fig2")
)
sapply(required_dirs, function(d) {
  if (!dir.exists(d)) {
    dir.create(d, recursive = TRUE, showWarnings = FALSE)
    message(paste("Created directory:", d))
  }
})

library(tidySummarizedExperiment)
library(tidySingleCellExperiment)
library(HDF5Array)
library(magrittr)
library(sccomp)
library(targets)
library(tidyverse)
library(tidybulk)
library(brms)
library(rstan)
library(cmdstanr)
library(here)  # For path management
library(plotly)
library(ggplot2)
library(ggforce)  # for geom_bspline
library(grid)     # for unit()
library(gganatogram)
library(zellkonverter)
library(cellNexus)
library(Seurat)
library(tidyseurat)
library(schex)
library(patchwork)
library(ComplexHeatmap)
library(tidyHeatmap)
library(ggnewscale)  # For multiple color scales in one plot

```

# Age range across tissue

```{r load-data}
# Resolve the configured pseudobulk SummarizedExperiment.
hdf5_path = path_pseudobulk_h5ad()

if (!file.exists(hdf5_path)) {
  warning(paste("HDF5 file not found at:", hdf5_path, "\nPlease update the path."))
}

# Resolve the vendored covariate-mapping helper.
edit_covariates_path = path_edit_covariates()

if (file.exists(edit_covariates_path)) {
  source(edit_covariates_path)
} else {
  warning(paste("edit_covariates.R not found at:", edit_covariates_path, "\nPlease update the path."))
}

cell_metadata <-
  get_publication_metadata() %>% filter(disease=="normal") |> filter(age_days > 365) 

samples = cell_metadata |>
  distinct(sample_id,
           self_reported_ethnicity,
           assay,
           age_days,
           sex,
           tissue,
           disease) |> as_tibble() %>%
  edit_covariates() %>%
  filter(sex %in% c("female", "male")) %>%
  mutate(
    age_groups = case_when(
      age_decade  %in% c('1', ' 2', '3') ~ 'Young',
      age_decade  %in% c('4', ' 5', '6') ~ 'Mid',
      age_decade  %in% c('7', ' 8', '9') ~ 'Old',
      .default = 'Old',
    )
  ) %>% distinct(sample_id, tissue_groups, age_years)

# Create dictionary of shortened tissue names (shared across plots)
tissue_name_dict <- c(
  "blood" = "Blood",
  "sensory-related structures" = "Sensory",
  "cerebral lobes and cortical areas" = "Cerebral cortex",
  "limbic and basal systems" = "Limbic/basal",
  "liver" = "Liver",
  "bone marrow" = "Bone marrow",
  "large intestine" = "Large int",
  "lymphatic system" = "Lymphatic",
  "small intestine" = "Small int",
  "respiratory system" = "Respiratory",
  "renal system" = "Renal",
  "brainstem and cerebellar structures" = "Hindbrain",
  "cardiovascular system" = "Cardiovascular",
  "general brain and major structures" = "Brain",
  "vasculature" = "Vasculature",
  "ovary" = "Ovary",
  "integumentary system (skin)" = "Skin",
  "nasal, oral, and pharyngeal regions" = "Pharyngeal",
  "muscular system (skeletal muscles)" = "Skeletal muscle",
  "prostate" = "Prostate",
  "breast" = "Breast",
  "digestive system (general)" = "Digestive",
  "digestive tract junctions and connections" = "Digestive junction",
  "oesophagus" = "Oesophagus",
  "female reproductive system" = "Female repro",
  "spleen" = "Spleen",
  "peritoneal and abdominal cavity structures" = "Peritoneal",
  "thymus" = "Thymus",
  "adipose tissue" = "Adipose",
  "trachea" = "Trachea",
  "endocrine system" = "Endocrine",
  "pancreas" = "Pancreas",
  "miscellaneous glands" = "Glands",
  "stomach" = "Stomach",
  "gallbladder" = "Gallbladder",
  "connective tissue" = "Connective"
)

```

```{r define-theme}
theme_multipanel =
  theme_bw() +
  theme(
    panel.border = element_blank(),
    axis.line = element_line(size = 0.1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    strip.background = element_blank(),
    axis.title.y = element_text(margin = margin(
      t = 0,
      r = 0,
      b = 0,
      l = 0
    ), size = 7),
    axis.title.x = element_text(margin = margin(
      t = 0,
      r = 0,
      b = 0,
      l = 0
    ), size = 7),
    panel.spacing.x = unit(0.1, "lines"),
    axis.text.x = element_text(size = 6),
    axis.text.y = element_text(size = 6),
    strip.text.x = element_text(size = 7),
    strip.text.y = element_text(size = 7),
    
    # legend
    legend.key.size = unit(5, 'mm'),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 6),
    
    # Avoid text clipping for facets. Currently not merged remotes::install_github("tidyverse/ggplot2#4223")
    strip.clip = "off",
    
    # Title
    plot.title = element_text(size = 7),
    
    axis.line.x = element_line(size = 0.2),
    axis.line.y = element_line(size = 0.2),
    axis.ticks.x = element_line(size = 0.2),
    axis.ticks.y = element_line(size = 0.2)
  )
```

```{r fig.width=6, fig.height=5, label="age-range-plot"}

# 1. summarize
Plot__sample_summary <-
  samples %>%
  group_by(tissue_groups) %>%
  summarise(
    n          = n(),
    min_age    = min(age_years, na.rm = TRUE),
    max_age    = max(age_years, na.rm = TRUE),
    median_age = median(age_years, na.rm = TRUE)
  ) %>%
  # Apply shortened names
  mutate(tissue_groups_short = recode(tissue_groups, !!!tissue_name_dict)) %>%
  # optional: order tissues by descending median age
  mutate(tissue_groups_short = fct_reorder(tissue_groups_short, -median_age)) %>%
  # 2. plot
  ggplot(aes(x = tissue_groups_short)) +
  # errorbar for [min, max]
  geom_errorbar(aes(ymin = min_age, ymax = max_age),
                width = 0,
                colour = "grey50") +
  
  # dot at median, sized by sample count
  geom_point(aes(y    = median_age, size = n), colour = "black") +
  
  scale_size_area(max_size = 3, name     = "Sample size") +
  
  labs(x = "Tissue", y = "Age (years)", title = "Age distribution by tissue") +
  theme_multipanel +
  theme(
    axis.text.x = element_text(
      angle = 90,
      vjust = 0.5,
      hjust = 1
    )
  )

Plot__sample_summary

```

# Volcano plot

```{r volcano-theme}

# Signed square-root transform for the x axis
signed_sqrt_trans <- scales::trans_new(
  name = "signed_sqrt",
  transform = function(x) sign(x) * sqrt(abs(x)),
  inverse = function(x) sign(x) * (x^2)
)

path_binary_test = file.path(path_intermediate("fig2"), "sccomp_binary_test_v2.rds")

if (file.exists(path_binary_test)) {
  binary_test = readRDS(path_binary_test)
} else {
  # Load the configured L3 sccomp estimates.
  est <- readRDS(
    path_sccomp_estimates("estimates_age_decade___L3___disease_FALSE___immune_only_TRUE.rds")
  )
  # Old (young: decades 1-4) vs late (decades 6-9), age_decade5 as reference
  binary_test = est |> sccomp_test(contrasts = '(0 + `age_decade6` + `age_decade7` + `age_decade8` + `age_decade9`)/5 - (`age_decade1` + `age_decade2` + `age_decade3` + `age_decade4`)/4')
  saveRDS(binary_test, file = path_binary_test)
}

# Dictionary of cell type abbreviations (consistent with Fig4.qmd)
cell_type_abbrev_dict <- c(
  # Red cell types
  "macrophage" = "Mac",
  "cd4 th1 em" = "CD4\nTh1\nEM",
  "cd4 th1/th17 em" = "CD4\nTh1/17 EM",
  "cd4 th17 em" = "CD4\nTh17 EM",
  "cd4 th2 em" = "CD4\nTh2 EM",
  "cd8 naive" = "CD8\nNaive",
  "plasma" = "Plasma",
  # Blue cell types
  "granulocyte" = "Gran",
  "cd4 naive" = "CD4\nNaive",
  "b naive" = "B\nNaive",
  "nk" = "NK",
  "nkt" = "NKT",
  "cdc" = "CDC",
  "pdc" = "PDC",
  # Additional cell types
  "b memory" = "B\nMemory",
  "cd14 mono" = "CD14\nMono",
  "cd16 mono" = "CD16\nMono",
  "cd4 fh em" = "CD4\nFH EM",
  "cd4 tcm" = "CD4\nTCM",
  "cd8 tcm" = "CD8\nTCM",
  "cd8 tem" = "CD8\nTEM",
  "ilc" = "ILC",
  "mait" = "MAIT",
  "mast" = "Mast",
  "tgd" = "TGD",
  "treg" = "Treg"
)


Plot__volcano <-
  binary_test |> # use sccomp test
  filter(L3 != "cytotoxic") %>% 
  mutate(
    naive_experienced = case_when(
      L3 |> str_detect("naive") ~ "Antigen-naive lymphocytes",
      L3 |> str_detect("cd4|cd8|memory") ~ "Antigen-experienced lymphocytes"
    )
  ) |>
  mutate(significant = c_FDR < 0.05) |>
  mutate(
    L3_abbrev = recode(L3, !!!cell_type_abbrev_dict),
    L3_abbrev = case_when(c_FDR < 0.05 ~ L3_abbrev)
  ) |>
  ggplot(aes(c_effect, c_FDR)) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "black", alpha = 0.5, linewidth = 0.3) +
  geom_point(aes(color = naive_experienced, size = significant), alpha = 0.6) +
  ggrepel::geom_text_repel(aes(label = L3_abbrev), size = 2, point.padding = 1.5) +
  ylab("False discovery rate
(inverse log scale)") +
  xlab("Composition change") +
  scale_y_continuous(trans = tidybulk::log10_reverse_trans()) +
  scale_x_continuous(trans = signed_sqrt_trans) +
  scale_color_brewer(palette = "Set1", na.value = "grey50") +
  scale_size_manual(values = c(`FALSE` = 0.5, `TRUE` = 1.5), guide = "none") +
  theme_multipanel


```

```{r fig.width = 7, fig.height = 6, label="volcano-plot"}
print(Plot__volcano)
```

# Anatogram

```{r load-anatogram-libs}

library(gganatogram)

# Path to the configured L0 (lineage-level) sccomp estimates.
est_l0_path = path_sccomp_estimates("estimates_age_decade___L0___disease_FALSE___immune_only_TRUE.rds")
if (file.exists(est_l0_path)) {
  est_l0 = readRDS(est_l0_path)
} else {
  stop(paste("Estimates file not found at:", est_l0_path, "\nPlease update the path."))
}

new_df <-
  est_l0 |>
  attr("count_data") |>
  sccomp:::.subset(sample_id) |>
  distinct(age_decade, tissue_groups) |>
  arrange(age_decade) |>
  mutate(sex = NA, ethnicity_groups_imputed = NA) |>
  mutate(sample_id = glue::glue("my_sample_{1:n()}")) |>
  mutate(sample_id = if_else(
    sample_id |> is.na(),
    glue::glue("my_sample_imputed_{row_number()}") ,
    sample_id
  )) 


pred =
    est_l0 |>
    sccomp_predict(
      formula_composition = ~ age_decade + sex + age_decade:sex + ethnicity_groups_imputed + (age_decade + sex + age_decade:sex | tissue_groups),
      new_data = new_df,
      summary_instead_of_draws = F,
      robust = TRUE
    )

myeloid_types  <- c("dc", "granulocyte", "macrophage", "mast", "monocytic")
lymphoid_types <- c("b", "t", "nk", "nkt", "ilc", "cytotoxic")

pred <- 
  pred %>% 
  mutate(lineage = case_when(
    L0 %in% myeloid_types  ~ "myeloid",
    L0 %in% lymphoid_types ~ "lymphoid",
    TRUE                   ~ NA_character_
  )) %>% 
  mutate(
    age_group = case_when(
      as.numeric(as.character(age_decade)) < 5 ~ 'young',
      as.numeric(as.character(age_decade)) >= 5 ~ 'old'
    )
  )



# 1. Define your mapping: your organ names → gganatogram keys
organ_map <- c(
  "adipose tissue"                       = "adipose_tissue",
  "blood"                                 = "leukocyte",
  "bone marrow"                           = "bone_marrow",
  "brainstem and cerebellar structures"   = "cerebellum",
  # choose one; adjust if needed
  "breast"                                = "breast",
  "cerebral lobes and cortical areas"     = "brain",
  "endocrine system"                      = "pituitary_gland",
  # representative; adjust if you prefer thyroid_gland, etc.
  "female reproductive system"            = NA_character_,
  # no male equivalent in hgMale_key
  "integumentary system (skin)"           = "skin",
  "large intestine"                       = "colon",
  "liver"                                 = "liver",
  "lymphatic system"                      = "lymph_node",
  "nasal, oral, and pharyngeal regions"   = "nasal_pharynx",
  "oesophagus"                            = "esophagus",
  "pancreas"                              = "pancreas",
  "prostate"                              = "prostate",
  "renal system"                          = "kidney",
  # or "renal_cortex"
  "respiratory system"                    = "lung",
  "sensory-related structures"            = NA_character_,
  # map manually if you pick eye, ear, etc.
  "small intestine"                       = "small_intestine",
  "spleen"                                = "spleen",
  "stomach"                               = "stomach",
  "thymus"                                = "thymus",
  "trachea"                               = "trachea"
)

# ---------------------------------------------------------------------------
# Draw-level statistical test of the age-related lineage shift per tissue.
#
# Instead of averaging `proportion` across draws first (which discards all
# uncertainty and forces every tissue to be plotted), we keep the posterior
# draws and, for each tissue, estimate:
#   * the fold change / effect of the old-vs-young shift per lineage, and
#   * a posterior probability + sccomp-style FDR for that shift,
# so that only tissues with a credible myeloid-vs-lymphoid shift are shown.
# ---------------------------------------------------------------------------

# Tissues are kept for plotting when their myeloid-vs-lymphoid contrast FDR is
# below this threshold.
fdr_threshold <- 0.05

# 1. Aggregate the L0-level proportions up to the lineage level for every
#    posterior draw (sum the cell types that belong to each lineage), then
#    average across the age decades that make up each age group.
draw_lineage <-
  pred %>%
  filter(!is.na(lineage)) %>%
  group_by(.draw, tissue_groups, age_decade, age_group, lineage) %>%
  summarise(proportion = sum(proportion, na.rm = TRUE), .groups = "drop") %>%
  group_by(.draw, tissue_groups, lineage, age_group) %>%
  summarise(proportion = mean(proportion, na.rm = TRUE), .groups = "drop")

# 2. Per draw: old-vs-young effect (difference) and log2 fold change per lineage.
draw_effect <-
  draw_lineage %>%
  pivot_wider(names_from = age_group, values_from = proportion) %>%
  filter(!is.na(old), !is.na(young)) %>%
  mutate(
    effect = old - young,
    lfc    = log2((old + 1e-6) / (young + 1e-6))
  )

# 3. Summarise the posterior per tissue x lineage:
#    fold change / effect (median across draws) and the posterior probability
#    of the shift together with a sccomp-style Bayesian FDR (sorted cumulative
#    mean of the probability that the effect lies on the opposite side of 0).
lineage_stats <-
  draw_effect %>%
  group_by(tissue_groups, lineage) %>%
  summarise(
    median_young = median(young),
    median_old   = median(old),
    c_effect     = median(effect),
    c_lfc        = median(lfc),
    fc           = 2 ^ median(lfc),
    p_increase   = mean(effect > 0),
    p_decrease   = mean(effect < 0),
    .groups      = "drop"
  ) %>%
  mutate(pH0 = pmin(p_increase, p_decrease)) %>%
  group_by(lineage) %>%
  arrange(pH0, .by_group = TRUE) %>%
  mutate(FDR = cummean(pH0)) %>%
  ungroup()

# 4. Directly contrast the two lineages within each tissue for each draw
#    (delta > 0 => myeloid increases relative to lymphoid, delta < 0 => the
#    reverse). This is the quantity that drives the anatogram colouring, so its
#    FDR is what decides whether a tissue is "significant".
draw_contrast <-
  draw_effect %>%
  dplyr::select(.draw, tissue_groups, lineage, effect) %>%
  pivot_wider(names_from = lineage, values_from = effect) %>%
  filter(!is.na(myeloid), !is.na(lymphoid)) %>%
  mutate(delta = myeloid - lymphoid)

contrast_stats <-
  draw_contrast %>%
  group_by(tissue_groups) %>%
  summarise(
    c_delta    = median(delta),
    p_myeloid  = mean(delta > 0),   # P(myeloid increase relative to lymphoid)
    p_lymphoid = mean(delta < 0),   # P(lymphoid increase relative to myeloid)
    .groups    = "drop"
  ) %>%
  mutate(pH0 = pmin(p_myeloid, p_lymphoid)) %>%
  arrange(pH0) %>%
  mutate(
    FDR         = cummean(pH0),
    significant = FDR < fdr_threshold
  ) %>%
  ungroup()

# 5. Assemble the plot data. Keep the per-lineage effect (`value`) that the
#    downstream anatogram code expects and attach the contrast statistics.
df <-
  lineage_stats %>%
  left_join(
    contrast_stats %>%
      dplyr::select(tissue_groups, c_delta, contrast_FDR = FDR, significant),
    by = "tissue_groups"
  ) %>%
  mutate(organ = recode(tissue_groups, !!!organ_map)) %>%
  rename(value = c_effect)
  


data("hgMale_key", package = "gganatogram")

# 3. Merge your values & lineage onto the key
df_merged <- hgMale_key %>%
  dplyr::select(-value) %>% 
  left_join(df, by = "organ") %>% 
  filter(!is.na(value))

```

## myeloid vs lymphoid

```{r fig.width=8, fig.height=6, label="myeloid-lymphoid-plot"}

# 1. define your palette
colour_scheme <- c(
  myeloid_pos_lymphoid_neg = "#d73027", # red
  lymphoid_pos_myeloid_neg = "#4575b4", # blue
  both_positive            = "#1a9850", # green
  both_negative            = "#762a83"   # purple
)

# 2. pivot + classify + assign colour
df_coloured <- df_merged %>%
  dplyr::select(organ, lineage, value) %>%
  pivot_wider(
    names_from = lineage,
    values_from = value
  ) %>%
  mutate(
    effect_category = case_when(
      lymphoid >  0 & myeloid >  0 ~ "both_positive",
      lymphoid <  0 & myeloid <  0 ~ "both_negative",
      myeloid >  0 & lymphoid <  0 ~ "myeloid increase\nlymphoid decrease",
      lymphoid >  0 & myeloid <  0 ~ "lymphoid increase\nmyeloid decrease",
      TRUE                          ~ NA_character_
    ),
    colour = colour_scheme[effect_category]
  ) %>% 
  mutate(value = abs(myeloid - lymphoid)) 

df_coloured %<>%
  mutate(
    value = ifelse(effect_category == 'lymphoid increase\nmyeloid decrease', -value, value)
  )

Plot__df_coloured <- 
  df_coloured %>% 
  rename(c_delta  = value) %>% 
  dplyr::select(-colour) %>% 
  left_join(df_merged) %>% 
  mutate(
    lineage_to_keep = case_when(
      lymphoid > myeloid ~ "lymphoid",
      myeloid > lymphoid ~ "myeloid",
      TRUE ~ lineage   # keeps both if exactly tied
    )
  ) %>%
  filter(lineage == lineage_to_keep) %>%
  select(-lineage_to_keep)

Plot__anatogram <-
  gganatogram(
    data = Plot__df_coloured  %>% 
      mutate(
        type = effect_category,
        value = c_delta 
      )
    %>% arrange(desc(organ == "skin"))
    ,
    fillOutline = 'lightgray', 
    organism = 'human',
    sex = 'male',
    fill = "value", ggplot2_only = T
  ) +
  facet_wrap( ~ type) +
  scale_fill_gradientn(
    colors = rev(RColorBrewer::brewer.pal(11, "RdBu")),
    # low=blue, high=red
    values = scales::rescale(c(
      min(df_coloured$value, na.rm = TRUE),
      0,
      max(df_coloured$value, na.rm = TRUE)
    )),
    limits = c(
      min(df_coloured$value, na.rm = TRUE),
      max(df_coloured$value, na.rm = TRUE)
    ),
    name   = NULL,
    # remove default "value"
    labels = c("Lymphoid increase", "Myeloid increase"),
    # bottom, top
    breaks = c(
      min(df_coloured$value, na.rm = TRUE),
      max(df_coloured$value, na.rm = TRUE)
    )              # only show two ticks
  ) +
  guides(
    fill = guide_colorbar(
      title = NULL,
      direction = "horizontal", 
      position = 'bottom',
      title.position = "bottom",
      title.hjust = 0.5,
      barwidth = 8,
      barheight = 0.8,
      # custom labels at ends
      label.position = "bottom"
    )
  ) +
  theme_void() 

```

## Add organ labels to the anatogram

Extract each organ's centroid from the built plot (using the panel where the
organ is actually coloured) and draw a leader line + text label just outside
the body outline. Ported from `report/Fig_files/Fig2.R`.

```{r fig.width=8, fig.height=6, label="anatogram-labels"}

# Thin borders for ALL polygon and path layers (body outline + organs)
Plot__anatogram$layers <- lapply(Plot__anatogram$layers, function(layer) {
  if (inherits(layer$geom, "GeomPolygon") || inherits(layer$geom, "GeomPath")) {
    layer$aes_params$linewidth <- 0.1
    layer$aes_params$colour <- "grey40"
  }
  layer
})

# Use ggplot_build to extract organ centroids WITH correct panel/facet assignments
built <- ggplot_build(Plot__anatogram)
facet_layout <- built$layout$layout
facet_layout$PANEL <- as.integer(as.character(facet_layout$PANEL))

# For each organ layer, find the panel where it has a visible (non-grey) fill and
# use that panel's polygon centroid as the label anchor. Layers 1-2 are the body
# outline + path, so organ layers start at index 3.
organ_centroids <- do.call(rbind, lapply(3:length(built$data), function(i) {
  d <- built$data[[i]]
  if (is.null(d) || nrow(d) == 0) return(NULL)
  if (!"x" %in% names(d) || !"y" %in% names(d)) return(NULL)

  raw_d <- Plot__anatogram$layers[[i]]$data
  organ_name <- if (is.data.frame(raw_d) && "id" %in% names(raw_d)) {
    unique(na.omit(raw_d$id))[1]
  } else {
    NA_character_
  }
  if (is.na(organ_name) || organ_name == "human_male_outline") return(NULL)

  d$PANEL_int <- as.integer(as.character(d$PANEL))
  panel_fills <- d %>%
    group_by(PANEL_int) %>%
    summarise(fill_val = unique(fill)[1], .groups = "drop")

  active_panel <- panel_fills %>%
    filter(fill_val != "lightgray" & fill_val != "#D3D3D3" & fill_val != "grey") %>%
    pull(PANEL_int)

  if (length(active_panel) == 0) active_panel <- panel_fills$PANEL_int[1]
  active_panel <- active_panel[1]

  panel_d <- d %>% filter(PANEL_int == active_panel)
  tibble(
    organ = organ_name,
    x = mean(panel_d$x, na.rm = TRUE),
    y = mean(panel_d$y, na.rm = TRUE),
    PANEL = active_panel
  )
}))

# Manually anchor "skin" (for label only, otherwise it will cover all other tissues on plot).
organ_centroids <-
  bind_rows(
    tibble(organ = "skin", x = 54.1, y = -103, PANEL = 1L),
    organ_centroids
  )

if (!is.null(organ_centroids) && nrow(organ_centroids) > 0) {
  organ_centroids <- organ_centroids %>%
    left_join(facet_layout %>% dplyr::select(PANEL, type), by = "PANEL") %>%
    mutate(label = gsub("_", " ", organ)) %>%
    filter(!is.na(type)) %>%
    distinct(organ, .keep_all = TRUE)

  # Panel 1 = left body, Panel 2 = right body
  left_facet  <- as.character(facet_layout$type[facet_layout$PANEL == 1])
  right_facet <- as.character(facet_layout$type[facet_layout$PANEL == 2])
  organ_centroids$type <- as.character(organ_centroids$type)

  left_labels  <- organ_centroids %>% filter(type == left_facet)
  right_labels <- organ_centroids %>% filter(type == right_facet)

  # Position labels just outside body edges, spread evenly in y
  if (nrow(left_labels) > 0) {
    left_labels <- left_labels %>%
      arrange(desc(y)) %>%
      mutate(
        label_x = 15,
        label_y = seq(from = max(y) + 5, to = min(y) - 5, length.out = n())
      )
  }

  if (nrow(right_labels) > 0) {
    right_labels <- right_labels %>%
      arrange(desc(y)) %>%
      mutate(
        label_x = 102,
        label_y = seq(from = max(y) + 5, to = min(y) - 5, length.out = n())
      )
  }

  Plot__anatogram <- Plot__anatogram +
    geom_segment(
      data = left_labels,
      aes(x = x, y = y, xend = label_x, yend = label_y),
      linewidth = 0.15, colour = "grey60", inherit.aes = FALSE
    ) +
    geom_text(
      data = left_labels,
      aes(x = label_x, y = label_y, label = label),
      size = 5 / .pt, hjust = 1, colour = "grey20", inherit.aes = FALSE
    ) +
    geom_segment(
      data = right_labels,
      aes(x = x, y = y, xend = label_x, yend = label_y),
      linewidth = 0.15, colour = "grey60", inherit.aes = FALSE
    ) +
    geom_text(
      data = right_labels,
      aes(x = label_x, y = label_y, label = label),
      size = 5 / .pt, hjust = 0, colour = "grey20", inherit.aes = FALSE
    ) +
    coord_cartesian(clip = "off") +
    theme(panel.spacing.x = unit(0, "pt"))
}

print(Plot__anatogram)

```

# Heatmap: Binarised age (early vs late)

Age effects (fixed and random) across different tissue

```{r}
#| label: heatmap load data
#| eval: false

# This chunk documents how `effect_by_tissue_fixed_plus_random.rds`
# (loaded in the plotting chunk below) was generated. Update the path to your
# sccomp estimates file to reproduce.
est <- readRDS(
  path_sccomp_estimates("estimates_age_decade___L3___disease_FALSE___immune_only_TRUE.rds")
)

cell_type_col = 'L3'

combination_fn_diff <- function(terms, age_split) {
  uq_split = unique(na.omit(age_split))
  if (length(uq_split) > 2) {
    stop("only binary splits supported")
  }
  if (!all(age_split %in% c("Early", "Late"))) {
    stop("age split should be either 'Early', 'Late' or NA")
  }

  # remove NAs
  terms = terms[!is.na(age_split)]
  age_split = na.omit(age_split)

  terms_split = split(terms, age_split)
  terms = lapply(names(terms_split), function(group_name) {
    x = terms_split[[group_name]]
    denom = if (group_name == "Late") length(x) + 1 else length(x)
    sprintf("((%s) / %d)", paste(x, collapse = " + "), denom)
  })
  names(terms) = names(terms_split)
  
  if (length(terms) == 1) {
    contr = terms[[1]]
    if (names(terms) == "Early") {
      contr = paste0("-", contr)
    }
  } else {
    contr = paste(terms$Late, terms$Early, sep = " - ")
  }
  
  return(contr)
}

age_bins_all = c("1", "2", "3", "4", "5", "6", "7", "8", '9', '10')
age_bins = c("1", "2", "3", "4", "6", "7", "8", '9', '10') # bins without intercept - Senior_50
age_binary = setNames(c("Early", "Early", "Early", "Early", "Late", "Late", "Late", "Late", "Late"), age_bins)

load(file.path(path_intermediate("fig2"), "df_cell_statistics.RData"))

param_df = est |>
  filter(grepl("age_decade", parameter) &
           !grepl(":", parameter)) |>
  mutate(params = str_split(parameter, "___", simplify = TRUE),
         params = (\(x) {
           colnames(x) = c("Age", "Tissue")
           as.data.frame(x)
         })(params)) |>
  dplyr::select(params, parameter) |>
  tidyr::unpack(params) %>%
  mutate(
    Age = str_replace(Age, "age_decade", ""),
    Age = str_replace(Age, ":.+", ""),
    Tissue = if_else(Tissue == "", "global", Tissue),
    AgeBinary = age_binary[Age]
  ) |>
  distinct()

param_df

contrasts = list()

# Age_only effect
contrasts = param_df %>% 
  drop_na() |>
  mutate(
    Tissue = sprintf("Age_only effect (%s)", Tissue),
    term = sprintf("`%s`", parameter)
  ) |>
  group_by(Tissue) |>
  reframe(
    contr = combination_fn_diff(term, AgeBinary)
  ) |>
  drop_na() |>
  deframe() |>
  c(contrasts)

contrasts2 = contrasts[setdiff(names(contrasts), "Age_only effect (global)")]
contrasts2_names = names(contrasts2)

contrasts2 = contrasts2 %>% paste('+', contrasts["Age_only effect (global)"]) %>% as.list()
names(contrasts2) = contrasts2_names

effect_by_tissue_fixed_plus_random = est |>
  sccomp_test(contrasts = contrasts2) |>
  mutate(
    test_type = gsub("^(.+effect) \\((.+)\\)$", "\\1", parameter),
    tissue = gsub("^(.+effect) \\((.+)\\)$", "\\2", parameter)
  ) |>
  dplyr::rename(cell_type_unified_ensemble = !!sym(cell_type_col))
```

## Plot__Heatmap_Binarised_age

```{r fig.height=9, fig.width=10}
#| label: Plot__Heatmap_Binarised_age

load(file.path(path_intermediate("fig2"), "df_cell_statistics.RData"))

effect_by_tissue_fixed_plus_random <- readRDS(file.path(path_intermediate("fig2"), "effect_by_tissue_fixed_plus_random.rds"))

fdr = effect_by_tissue_fixed_plus_random$c_FDR
min_fdr = fdr[fdr > 0] %>% min

effect_by_tissue_fixed_plus_random$c_FDR[fdr == 0] = min_fdr

# Cell type abbreviations for the heatmap (single line, no line breaks)
cell_type_abbrev_dict_heatmap <- c(
  "macrophage" = "Mac",
  "cd4 th1 em" = "CD4 Th1 EM",
  "cd4 th1/th17 em" = "CD4 Th1/17 EM",
  "cd4 th17 em" = "CD4 Th17 EM",
  "cd4 th2 em" = "CD4 Th2 EM",
  "cd8 naive" = "CD8 Naive",
  "plasma" = "Plasma",
  "granulocyte" = "Gran",
  "cd4 naive" = "CD4 Naive",
  "b naive" = "B Naive",
  "nk" = "NK",
  "nkt" = "NKT",
  "cdc" = "CDC",
  "pdc" = "PDC",
  "b memory" = "B Memory",
  "cd14 mono" = "CD14 Mono",
  "cd16 mono" = "CD16 Mono",
  "cd4 fh em" = "CD4 FH EM",
  "cd4 tcm" = "CD4 TCM",
  "cd8 tcm" = "CD8 TCM",
  "cd8 tem" = "CD8 TEM",
  "ilc" = "ILC",
  "mait" = "MAIT",
  "mast" = "Mast",
  "tgd" = "TGD",
  "treg" = "Treg"
)

library(grid)

bar_size_top  <- grid::unit(3, "mm")
bar_size_side <- grid::unit(5, "mm")

anno_axis_none <- list(
  labels = FALSE,
  at = NULL
)

anno_name_gp <- grid::gpar(
  fontsize = 7,
  fontface = "plain"
)

Plot__Heatmap_Binarised_age <-
  effect_by_tissue_fixed_plus_random |>
  as_tibble() |>
  mutate(size = -log10(c_FDR)) |>
  left_join(df_cells_ct, by = "cell_type_unified_ensemble") |>
  left_join(df_cells_tissue, by = "tissue") |>
  mutate(
    tissue_short = recode(tissue, !!!tissue_name_dict),
    cell_type_abbrev = recode(
      cell_type_unified_ensemble,
      !!!cell_type_abbrev_dict_heatmap
    ),
    cell_type_abbrev = factor(
      cell_type_abbrev,
      levels = unname(cell_type_abbrev_dict_heatmap)
    )
  ) |>
  heatmap(
    .row = tissue_short,
    .column = cell_type_abbrev,
    .value = c_effect,
    name = "Effect\nsize",
    palette_value = circlize::colorRamp2(
      seq(-1, 1, length.out = 7),
      scico::scico(7, palette = "vik")
    ),
    row_title = NULL,
    column_title = NULL,

    row_dend_width = grid::unit(3, "mm"),
    column_dend_height = grid::unit(4, "mm"),

    row_names_gp = grid::gpar(fontsize = 7, fontface = "plain"),
    column_names_gp = grid::gpar(fontsize = 7, fontface = "plain"),
    row_title_gp = grid::gpar(fontsize = 8, fontface = "plain"),
    column_title_gp = grid::gpar(fontsize = 8, fontface = "plain")
  ) |>
  layer_point(c_FDR < 0.05, .size = size) |>
  annotation_bar(
    NCellsCellType,
    annotation_label = "log(Cells)",
    size = bar_size_top,
    annotation_name_gp = anno_name_gp,
    axis_param = anno_axis_none,
    axis = FALSE
  ) |>
  annotation_bar(
    NSamplesCellType,
    annotation_label = "log(Samples)",
    size = bar_size_top,
    annotation_name_gp = anno_name_gp,
    axis_param = anno_axis_none,
    axis = FALSE
  ) |>
  annotation_bar(
    NCellsTissue,
    annotation_label = "log(Cells)",
    annotation_name_rot = 90,
    width = bar_size_side,
    annotation_name_gp = anno_name_gp,
    axis_param = anno_axis_none,
    axis = FALSE
  ) |>
  annotation_bar(
    NSamplesTissue,
    annotation_label = "log(Samples)",
    annotation_name_rot = 90,
    width = bar_size_side,
    annotation_name_gp = anno_name_gp,
    axis_param = anno_axis_none,
    axis = FALSE
  )

Plot__Heatmap_Binarised_age
```


# Boxplot of Up and Down cell type

```{r load-predictions}

path_sccomp_age_pred_draws = file.path(path_intermediate("fig2"), "sccomp_age_pred_draws.rds")

if (file.exists(path_sccomp_age_pred_draws)) {
  pred_draws = readRDS(path_sccomp_age_pred_draws)
} else {
  # Load the configured L3 sccomp estimates.
  est <- readRDS(
    path_sccomp_estimates("estimates_age_decade___L3___disease_FALSE___immune_only_TRUE.rds")
  )

  cell_type_col = 'L3'
  
  pred_draws =
    est |>
    sccomp_predict(
      formula_composition = ~ age_decade * sex + ethnicity_groups_imputed,
      new_data =
        est |>
        attr("count_data") |>
        sccomp:::.subset(sample_id) |>
        distinct(age_decade) |>
        arrange(age_decade) |>
        mutate(sex = NA, ethnicity_groups_imputed = NA) |>
        mutate(sample_id = glue("my_sample_{1:n()}")),
      number_of_draws = 100,
      summary_instead_of_draws = FALSE
    ) |>
    rename(proportion_mean = proportion) |>
    mutate(sample_id_original = sample_id) |>
    unite("sample_id", c(sample_id, .draw), remove = FALSE)
  
  saveRDS(pred_draws, file = path_sccomp_age_pred_draws)
}

path_sccomp_age_adjusted = file.path(path_intermediate("fig2"), "sccomp_age_adjusted.rds")

if (file.exists(path_sccomp_age_adjusted)) {
  age_adjusted = readRDS(path_sccomp_age_adjusted)
} else {
  age_adjusted =
    est |>
    sccomp_predict(
      formula_composition = ~ age_decade * sex + ethnicity_groups_imputed,
      new_data =
        est |>
        attr("count_data") |>
        sccomp:::.subset(sample_id),
      number_of_draws = 100,
      summary_instead_of_draws = FALSE
    ) |>
    rename(proportion_mean = proportion) |>
    mutate(sample_id_original = sample_id) |>
    unite("sample_id", c(sample_id, .draw), remove = FALSE)
  
  saveRDS(age_adjusted, file = path_sccomp_age_adjusted)
}

```

## upregulated

```{r define-upregulated}

path_binary_test = file.path(path_intermediate("fig2"), "sccomp_binary_test_v2.rds")

binary_test = readRDS(path_binary_test)

prop_upregulated_cell_type_df <- 
  binary_test %>% 
  as_tibble() %>% 
  filter(c_FDR < 0.05 & c_effect > 0) %>% 
  filter(L3 != 'cytotoxic') %>% 
  arrange(c_FDR) %>% 
  head(5) 

prop_upregulated_cell_type <- prop_upregulated_cell_type_df$L3


df_up = bind_rows(
    pred_draws %>%
      filter(L3 %in% prop_upregulated_cell_type) %>%
      dplyr::select(sample_id, L3, age_decade, proportion_mean) %>%
      mutate(version = 'prediction'), 
    age_adjusted %>% 
      filter(L3 %in% prop_upregulated_cell_type) %>% 
      group_by(sample_id, L3, age_decade) %>% 
      summarise(proportion_mean = mean(proportion, na.rm =T)) %>%
      ungroup %>% 
      dplyr::select(sample_id, L3, age_decade, proportion_mean) %>%
      mutate(version = 'adjusted')
  ) %>% 
  mutate(
    L3_abbrev_base = recode(L3, !!!cell_type_abbrev_dict)
  ) %>%
  left_join(
    prop_upregulated_cell_type_df %>% 
      dplyr::select(L3, c_FDR),
    by = "L3"
  ) %>%
  mutate(
    c_FDR = replace_na(c_FDR, 0),
    # Format FDR for strip title
    # Use better formatting: scientific notation for very small values, decimal for larger ones
    # Remove leading zero before decimal point
    fdr_formatted = case_when(
      c_FDR == 0 ~ "0",
      c_FDR < 0.001 ~ gsub("^0\\.", ".", formatC(c_FDR, format = "e", digits = 2)),
      TRUE ~ gsub("^0\\.", ".", formatC(c_FDR, format = "f", digits = 3))
    ),
    # Format strip title: cell type, then FDR (2 lines)
    L3_abbrev = paste0(L3_abbrev_base, "\n(FDR: ", fdr_formatted, ')')
  ) %>%
  # Create factor levels in the specified order
  {
    # Get unique L3_abbrev values in the desired order: cd4 th1/th17 em, nkt, granulocyte, cd16 mono, nk
    desired_order <- c("cd4 th1/th17 em", "nkt", "granulocyte", "cd16 mono", "nk")
    df_temp <- .
    unique_levels <- df_temp %>%
      distinct(L3, L3_abbrev) %>%
      arrange(match(L3, desired_order)) %>%
      pull(L3_abbrev)
    
    df_temp %>%
      mutate(L3_abbrev = factor(L3_abbrev, levels = unique_levels)) %>%
      dplyr::select(-L3_abbrev_base, -fdr_formatted)
  }

Plot__boxplot_up_cell_type <-
  df_up %>% 
  mutate(age_decade = factor(age_decade, levels = as.character(1:10))) %>%
  ggplot(aes(x = age_decade, y = proportion_mean)) + 
  
  # dots for the other version
  ggbeeswarm::geom_quasirandom(
    data = . %>% filter(version == "adjusted"),
    width = 0.3,
    colour = "grey60",
    shape  = '.',
    size   = 0.15,
    alpha  = 0.6
  ) +
  
  # boxplot for one version only
  geom_boxplot(
    aes(fill = version),
    data = . %>% filter(version == "prediction"),  
    outlier.shape = NA,
    alpha = 0.4,
    linewidth = 0.15,
    width = 0.7
  ) +
  scale_y_continuous(
    trans = scales::trans_new(
      name = "asin_sqrt",
      transform = function(x) asin(sqrt(x)),
      inverse = function(y) sin(y)^2
    ),
    expand = expansion(mult = c(0, 0.05)),
    labels = function(x) {
      # Remove leading zero before decimal point
      gsub("^0\\.", ".", format(x, nsmall = 1))
    }
  ) +
  scale_x_discrete(labels = function(x) x) +
  guides(fill = 'none') +
  facet_wrap(
    ~ L3_abbrev,
    nrow    = 1,
    scales  = "free_y"
  ) +
  labs(title = '', x = "Age Decade", y = "Proportion Mean (arcsin-sqrt scale)") +
  theme_multipanel +
  theme(
    strip.text.x.top = element_text(margin = margin(0, 0, 0, 0, "mm")),
    strip.placement = "inside",
    axis.ticks.x = element_blank(),
    panel.spacing.x = unit(0, "mm"),
    panel.spacing.y = unit(0, "mm"),
    legend.position = "none"
  ) 
  

```

```{r fig.width = 10, fig.height = 4, label="boxplot-upregulated"}

print(Plot__boxplot_up_cell_type)

```

## downregulated

```{r define-downregulated}
prop_downregulated_cell_type_df <- 
  binary_test %>% 
  as_tibble() %>% 
  filter(c_FDR < 0.05 & c_effect < 0) %>% 
  filter(L3 != 'cytotoxic') %>% 
  arrange(c_FDR) %>% 
  head(5) 

prop_downregulated_cell_type <- prop_downregulated_cell_type_df$L3 
  

df_down = bind_rows(
    pred_draws %>%
      filter(L3 %in% prop_downregulated_cell_type) %>%
      dplyr::select(sample_id, L3, age_decade, proportion_mean) %>%
      mutate(version = 'prediction'), 
    age_adjusted %>% 
      filter(L3 %in% prop_downregulated_cell_type) %>% 
      group_by(sample_id, L3, age_decade) %>% 
      summarise(proportion_mean = mean(proportion, na.rm =T)) %>%
      ungroup %>%  
      dplyr::select(sample_id, L3, age_decade, proportion_mean) %>%
      mutate(version = 'adjusted')
  ) %>% 
  mutate(
    L3_abbrev_base = recode(L3, !!!cell_type_abbrev_dict)
  ) %>%
  left_join(
    prop_downregulated_cell_type_df %>% 
      dplyr::select(L3, c_FDR),
    by = "L3"
  ) %>%
  mutate(
    c_FDR = replace_na(c_FDR, 0),
    # Format FDR for strip title
    # Use better formatting: scientific notation for very small values, decimal for larger ones
    # Remove leading zero before decimal point
    fdr_formatted = case_when(
      c_FDR == 0 ~ "0",
      c_FDR < 0.001 ~ gsub("^0\\.", ".", formatC(c_FDR, format = "e", digits = 0)),
      TRUE ~ gsub("^0\\.", ".", formatC(c_FDR, format = "f", digits = 3))
    ),
    # Format strip title: cell type, then FDR (2 lines)
    L3_abbrev = paste0(L3_abbrev_base, "\n(FDR: ", fdr_formatted, ')')
  ) %>%
  # Create factor levels in the specified order
  {
    # Get unique L3_abbrev values in the desired order
    desired_order <- c("cd8 naive", "cd4 naive", "b naive", "plasma", "b memory")
    df_temp <- .
    unique_levels <- df_temp %>%
      distinct(L3, L3_abbrev) %>%
      arrange(match(L3, desired_order)) %>%
      pull(L3_abbrev)
    
    df_temp %>%
      mutate(L3_abbrev = factor(L3_abbrev, levels = unique_levels)) %>%
      dplyr::select(-L3_abbrev_base, -fdr_formatted)
  }

Plot__boxplot_down_cell_type <-
  df_down %>% 
  mutate(age_decade = factor(age_decade, levels = as.character(1:10))) %>%
  ggplot(aes(x = age_decade, y = proportion_mean)) + 
  
  # dots for the other version
  ggbeeswarm::geom_quasirandom(
    data = . %>% filter(version == "adjusted"),
    width = 0.3,
    colour = "grey60",
    shape  = '.',
    size   = 0.15,
    alpha  = 0.6
  ) +
  
  # boxplot for one version only
  geom_boxplot(
    aes(fill = version),
    data = . %>% filter(version == "prediction"),  
    outlier.shape = NA,
    alpha = 0.4,
    linewidth = 0.15,
    width = 0.7
  ) +
  scale_fill_manual(values = c("prediction" = "#ADD8E6")) +
  scale_y_continuous(
    trans = scales::trans_new(
      name = "asin_sqrt",
      transform = function(x) asin(sqrt(x)),
      inverse = function(y) sin(y)^2
    ),
    expand = expansion(mult = c(0, 0.05)),
    labels = function(x) {
      # Remove leading zero before decimal point
      gsub("^0\\.", ".", format(x, nsmall = 1))
    }
  ) +
  scale_x_discrete(labels = function(x) x) +
  guides(fill = 'none') +
  facet_wrap(
    ~ L3_abbrev,
    nrow    = 1,
    scales  = "free_y"
  ) +
  labs(title = '', x = "Age Decade", y = "Proportion Mean (arcsin-sqrt scale)") +
  theme_multipanel +
  theme(
    strip.text.x.top = element_text(margin = margin(0, 0, 0, 0, "mm")),
    strip.placement = "inside",
    axis.ticks.x = element_blank(),
    panel.spacing.x = unit(0, "mm"),
    panel.spacing.y = unit(0, "mm"),
    legend.position = "none"
  ) 
  
```

```{r fig.width = 10, fig.height = 4, label="boxplot-downregulated"}
print(Plot__boxplot_down_cell_type)
```

# Line chart

```{r}
#| label: sccomp_prediction

path_sccomp_prediction = file.path(path_intermediate("fig2"), "sccomp_prediction____age_decade___L3___disease_FALSE___immune_only_TRUE.rds")

if (file.exists(path_sccomp_prediction)) {
  sccomp_df = readRDS(path_sccomp_prediction)
} else {
  # Load the configured L3 sccomp estimates.
  est <- readRDS(
    path_sccomp_estimates("estimates_age_decade___L3___disease_FALSE___immune_only_TRUE.rds")
  )

  cell_type_col = 'L3'
  
  new_df = est %>%  attr("count_data") 
  new_df$disease_groups___altered = NA
  new_df$ethnicity_groups_imputed = NA
  new_df$assay_groups___altered = NA
  new_df$dataset_id___altered = NA
  new_df$sex = NA
  new_df = new_df %>% dplyr::select(-!!cell_type_col) %>% dplyr::select(-n) %>% distinct()
  
  sccomp_df = sccomp_predict(
    est, 
    new_data = new_df,
    formula_composition = ~ 1 + age_decade + sex + age_decade:sex + ethnicity_groups_imputed + assay_groups___altered  + ( 1 + age_decade + sex + age_decade:sex + ethnicity_groups_imputed | tissue_groups),
    number_of_draws = 100,
    summary_instead_of_draws = F
  ) |>
    dplyr::rename(cell_type_unified_ensemble = !!sym(cell_type_col))
  
  saveRDS(sccomp_df, file = path_sccomp_prediction)
}
```

```{r}
#| label: create plasma line chart

transform_fn = \(x) x

x = 'plasma'
age_bins_all = c("1", "2", "3", "4", "5", "6", "7", "8", '9', '10')

tissue_groups_of_interest <-
  sccomp_df %>%
  filter(cell_type_unified_ensemble == x) %>%
  distinct(tissue_groups, age_decade) %>% group_by(tissue_groups) %>% count(tissue_groups) %>% filter(n > 3)  %>% pull(tissue_groups)

# Prepare c_FDR data for age_decade 5
c_fdr_data <- effect_by_tissue_fixed_plus_random |>
  as_tibble() |>
  filter(cell_type_unified_ensemble == x) |>
  dplyr::select(tissue, cell_type_unified_ensemble, c_FDR, c_effect) |>
  mutate(
    tissue_groups = tissue,
    size = -log10(c_FDR),
    significant = c_FDR < 0.05,
    tissue_groups_short = recode(tissue_groups, !!!tissue_name_dict),
    age_decade = "5"
  ) 

# Prepare data for the plot
plot_data <- sccomp_df |>
  filter(
    tissue_groups %in% tissue_groups_of_interest,
    cell_type_unified_ensemble == x
  ) |>
  mutate(
    age_decade = factor(age_decade, age_bins_all),
    proportion_mean = transform_fn(proportion_mean),
    proportion_lower = transform_fn(proportion_lower),
    proportion_upper = transform_fn(proportion_upper),
    # Apply shortened tissue names using the same dictionary
    tissue_groups_short = recode(tissue_groups, !!!tissue_name_dict),
    tissue_groups_short = factor(
      tissue_groups_short,
      levels = c(setdiff(unique(tissue_groups_short), "Large intestine"), "Large intestine")
    )
  )

# Prepare points data for age_decade 5 (only significant)
# Start with c_fdr_data for all significant tissues, then get proportion_mean
# For tissues without data at age_decade 5, interpolate from nearby age_decades
points_data <- c_fdr_data |>
  filter(
    tissue_groups %in% tissue_groups_of_interest,
    cell_type_unified_ensemble == x
  ) %>% 
  filter(significant == TRUE) |>
  left_join(
    plot_data |>
      filter(age_decade == "5") |>
      dplyr::select(tissue_groups, cell_type_unified_ensemble, proportion_mean),
    by = c("tissue_groups", "cell_type_unified_ensemble")
  ) %>% distinct()

# Calculate interpolation values for tissues missing data at age_decade 5
interp_values <- points_data |>
  filter(is.na(proportion_mean)) |>
  distinct(tissue_groups, cell_type_unified_ensemble) |>
   rowwise() |>
   mutate(
     row_tissue = tissue_groups,
     row_cell = cell_type_unified_ensemble,
     interp_value = {
       tissue_data <- plot_data |>
         filter(
           tissue_groups == row_tissue,
           cell_type_unified_ensemble == row_cell
         ) |>
         arrange(as.numeric(age_decade))
      
      if (nrow(tissue_data) > 0) {
        ages <- as.numeric(tissue_data$age_decade)
        values <- tissue_data$proportion_mean
        
        # Find closest younger age (age < 5)
        younger_idx <- which(ages < 5)
        younger_age <- if (length(younger_idx) > 0) max(ages[younger_idx]) else NULL
        younger_value <- if (!is.null(younger_age)) values[ages == younger_age][1] else NULL
        
        # Find closest older age (age > 5)
        older_idx <- which(ages > 5)
        older_age <- if (length(older_idx) > 0) min(ages[older_idx]) else NULL
        older_value <- if (!is.null(older_age)) values[ages == older_age][1] else NULL
        
        # Linear interpolation between the two closest points
        if (!is.null(younger_age) && !is.null(older_age)) {
          # Linear interpolation: y = y1 + (y2 - y1) * (x - x1) / (x2 - x1)
          younger_value + (older_value - younger_value) * (5 - younger_age) / (older_age - younger_age)
        } else {
          NA_real_
        }
      } else {
        NA_real_
      }
     }
   ) |>
   ungroup() |>
   select(-row_tissue, -row_cell)

# Join interpolation values back and update proportion_mean
points_data <- points_data |>
  left_join(
    interp_values |>
      dplyr::select(tissue_groups, cell_type_unified_ensemble, interp_value),
    by = c("tissue_groups", "cell_type_unified_ensemble")
  ) |>
  mutate(
    proportion_mean = if_else(is.na(proportion_mean), interp_value, proportion_mean)
  ) |>
  dplyr::select(-interp_value) |>
  filter(!is.na(proportion_mean))

plot_data <-
  plot_data %>% 
  mutate(significant = tissue_groups %in% (points_data %>% filter(significant) %>% pull(tissue_groups))) %>%
  # sccomp returns one identical prediction row per source sample. Collapse to
  # the unique tissue-by-age estimate actually represented by each trajectory,
  # while retaining the cohort size for the supplementary table.
  group_by(
    tissue_groups, tissue_groups_short, cell_type_unified_ensemble,
    age_decade, significant
  ) %>%
  summarise(
    n_source_rows = n(),
    n_samples = n_distinct(sample_id),
    proportion_mean = first(proportion_mean),
    proportion_lower = first(proportion_lower),
    proportion_upper = first(proportion_upper),
    .groups = "drop"
  )

Plot__plasma_line_chart <-
  plot_data |>
  ggplot() +
  geom_point(aes(x = age_decade, y = proportion_mean), shape = NA) +
  geom_line(
    aes(x = as.numeric(age_decade), y = proportion_mean, colour = significant), 
    show.legend = FALSE, linewidth = 0.15
  ) +
  geom_ribbon(aes(x = as.numeric(age_decade), ymin = proportion_lower, ymax = proportion_upper), colour = NA, alpha = 0.1, fill = "grey80") +
  scale_size_continuous(range = c(0.5, 2), guide = guide_legend(title = "-log10(FDR)")) +
  scale_y_continuous(
    trans = scales::trans_new(
      name = "asin_sqrt",
      transform = function(x) asin(sqrt(x)),
      inverse = function(y) sin(y)^2
    ),
    expand = expansion(mult = c(0, 0.05)),
    labels = function(x) {
      # Remove leading zero before decimal point
      gsub("^0\\.", ".", format(x, nsmall = 1))
    }
  ) + 
  scale_colour_manual(values = c(
      "TRUE" = "#B24A3B",
      "FALSE" = "black"
    ), guide = guide_legend(override.aes = list(alpha = 1))) +
  labs(x = "Age decade", y = "Proportion Mean (arcsin-sqrt scale)") +
  ggh4x::facet_wrap2(~ tissue_groups_short, ncol = 6, scales = 'free_y', axes = "margins") +
  theme_multipanel +
  theme(
    legend.position = "bottom",
    axis.ticks.x.bottom = element_blank(),
    axis.text.x = element_text(angle = 0, hjust = 1)
  )

```

```{r fig.width=15, fig.height=12}
#| label: plot plasma line chart

Plot__plasma_line_chart
```

# Supplementary tables

Export plot statistics from ggplot / heatmap `@data` slots for all Fig 2 panels.

```{r export-supplementary-tables}
message(
  "Panel caches updated. Run `Rscript report/build_supplementary_workbooks.R` ",
  "to rebuild the split-tab workbooks with separate plot-data, summary and test sheets."
)
```

# Save panels

```{r save-all-plots, eval=FALSE}
# Save all Plot__ objects for figure assembly
save(
  list = ls(pattern = "Plot__"),
  file = file.path(path_figure_ready("fig2"), "Fig2_plot_data_all.RData")
)

sessionInfo()
```
