---
title: "Figure 4 supplementary plots"
date: today
date-format: "DD MMM YYYY"
format:
  html:
    theme: minty
    toc: true
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
---

# Purpose

This report uses the finalized Figure 4 source-data workbook to provide:

1. individual prediction points and regression diagnostics for every tissue;
2. the full tissue distribution of blood-clock prediction error; and
3. a donor-by-tissue error map for every donor represented in at least three
   tissues, rather than only the nine donors shown in the main figure.

The ageing-distance trajectories are not duplicated because the main Figure 4
already shows all eligible tissues. A feature-count sensitivity plot should be
added only when feature counts per tissue are exported to source data.

```{r setup}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(here)
library(knitr)

source(file.path("..", "R", "paths.R"))
workbook <- path_source_tables("Supplementary_File_Figure_4.xlsx")
if (!file.exists(workbook)) stop("Figure 4 supplementary file not found: ", workbook)

plot_object_dir <- file.path(publication_vignette_root(), "supplementary_plot_objects")
dir.create(plot_object_dir, recursive = TRUE, showWarnings = FALSE)

# Submission workbooks store a caption in row 1 and headers in row 2.
read_panel_sheet <- function(sheet) {
  if (!sheet %in% excel_sheets(workbook)) stop("Missing worksheet: ", sheet)
  read_xlsx(workbook, sheet = sheet, skip = 1, guess_max = 1000000L)
}

save_plot_object <- function(plot, filename) {
  path <- file.path(plot_object_dir, filename)
  saveRDS(plot, path, compress = "xz")
  message("Saved ggplot object: ", path)
  invisible(path)
}

# Same short tissue labels used for Fig2 / Fig5 supplementary line charts.
tissue_name_dict <- c(
  "blood" = "Blood",
  "sensory-related structures" = "Sensory",
  "cerebral lobes and cortical areas" = "Cerebral cortex",
  "limbic and basal systems" = "Limbic/basal",
  "liver" = "Liver",
  "bone marrow" = "Bone marrow",
  "large intestine" = "Large int",
  "lymphatic system" = "Lymphatic",
  "small intestine" = "Small int",
  "respiratory system" = "Respiratory",
  "renal system" = "Renal",
  "brainstem and cerebellar structures" = "Hindbrain",
  "cardiovascular system" = "Cardiovascular",
  "general brain and major structures" = "Brain",
  "vasculature" = "Vasculature",
  "ovary" = "Ovary",
  "integumentary system (skin)" = "Skin",
  "nasal, oral, and pharyngeal regions" = "Pharyngeal",
  "muscular system (skeletal muscles)" = "Skeletal muscle",
  "prostate" = "Prostate",
  "breast" = "Breast",
  "digestive system (general)" = "Digestive",
  "oesophagus" = "Oesophagus",
  "female reproductive system" = "Female repro",
  "spleen" = "Spleen",
  "peritoneal and abdominal cavity structures" = "Peritoneal",
  "thymus" = "Thymus",
  "adipose tissue" = "Adipose",
  "trachea" = "Trachea",
  "endocrine system" = "Endocrine",
  "pancreas" = "Pancreas",
  "miscellaneous glands" = "Glands",
  "stomach" = "Stomach",
  "gallbladder" = "Gallbladder"
)

short_tissue_label <- function(x) {
  key <- tolower(as.character(x))
  key <- ifelse(
    key == "nasal, oral and pharyngeal regions",
    "nasal, oral, and pharyngeal regions",
    key
  )
  key <- ifelse(key == "skin", "integumentary system (skin)", key)
  out <- unname(tissue_name_dict[key])
  ifelse(is.na(out), tools::toTitleCase(gsub("_", " ", key, fixed = TRUE)), out)
}

theme_supplementary <- theme_bw(base_size = 8) +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_blank(),
    legend.position = "bottom",
    plot.title.position = "plot"
  )
```

```{r read-source-data}
clock_predictions <- read_panel_sheet("Fig4_E_predictions") |>
  transmute(
    donor_id = .data[["Donor ID"]],
    tissue_groups = .data[["Tissue group"]],
    tissue_label = short_tissue_label(tissue_groups),
    age_years = .data[["Chronological age (years)"]],
    predicted_age_scaled = .data[["Predicted age (years)"]],
    prediction_error_years = .data[["Prediction error (years)"]]
  ) |>
  filter(
    is.finite(age_years),
    is.finite(predicted_age_scaled),
    !is.na(tissue_groups)
  )

clock_models <- read_panel_sheet("Fig4_E_performance") |>
  transmute(
    tissue_groups = .data[["Tissue group"]],
    tissue_label = short_tissue_label(tissue_groups),
    n = .data[["Number of prediction rows"]],
    correlation = .data[["Pearson correlation"]],
    r_squared = .data[["R²"]],
    rmse_identity = .data[["Root-mean-square error (years)"]],
    mae_identity = .data[["Mean absolute error (years)"]]
  ) |>
  filter(is.finite(n), n >= 2, is.finite(correlation), is.finite(rmse_identity)) |>
  mutate(
    metric_label = sprintf(
      "n=%s\nr=%.2f; R²=%.2f\nRMSE=%.1f; MAE=%.1f",
      format(n, big.mark = ","), correlation, r_squared, rmse_identity, mae_identity
    )
  )

clock_predictions <- clock_predictions |>
  semi_join(clock_models, by = "tissue_groups")
```

# Per-tissue clock performance

The main panel shows individual points for blood but mainly fitted lines for
non-blood tissues. These small multiples restore the individual observations
and report the same model and identity-error statistics exported with the
figure.

```{r fig4-clock-by-tissue, fig.width=12, fig.height=13}
age_limits <- range(
  c(clock_predictions$age_years, clock_predictions$predicted_age_scaled),
  na.rm = TRUE
)

Plot_SFig4_clock_by_tissue <-
  ggplot(clock_predictions, aes(x = age_years, y = predicted_age_scaled)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey55", linewidth = 0.3) +
  geom_point(alpha = 0.10, size = 0.25, colour = "#315A7D") +
  geom_smooth(method = "lm", se = FALSE, colour = "#B24A3B", linewidth = 0.55) +
  geom_text(
    data = clock_models,
    aes(x = -Inf, y = Inf, label = metric_label),
    inherit.aes = FALSE,
    hjust = -0.05,
    vjust = 1.05,
    size = 2.15,
    lineheight = 0.9
  ) +
  facet_wrap(~ tissue_label, ncol = 5) +
  coord_cartesian(xlim = age_limits, ylim = age_limits) +
  labs(
    x = "Chronological age (years)",
    y = "Blood-clock predicted age (years)",
    title = "Blood-derived immune-age clock performance in each tissue"
  ) +
  theme_supplementary

Plot_SFig4_clock_by_tissue
save_plot_object(Plot_SFig4_clock_by_tissue, "Fig4_supp_clock_by_tissue.rds")
rm(Plot_SFig4_clock_by_tissue)
invisible(gc())
```

# Prediction-error distributions

```{r fig4-clock-errors, fig.width=9, fig.height=8}
tissue_order <- clock_predictions |>
  group_by(tissue_label) |>
  summarise(median_error = median(prediction_error_years, na.rm = TRUE), .groups = "drop") |>
  arrange(median_error) |>
  pull(tissue_label)

clock_error_data <- clock_predictions |>
  mutate(
    tissue_label = factor(tissue_label, levels = tissue_order),
    is_blood = tissue_label == "Blood"
  )

# Highlight blood on the y-axis (factor levels run bottom → top).
y_axis_colours <- ifelse(as.character(tissue_order) == "Blood", "#B24A3B", "black")

Plot_SFig4_clock_error_distribution <-
  ggplot(clock_error_data, aes(x = prediction_error_years, y = tissue_label)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50", linewidth = 0.35) +
  geom_boxplot(
    aes(fill = is_blood, colour = is_blood),
    width = 0.65,
    outlier.shape = NA,
    linewidth = 0.35
  ) +
  scale_fill_manual(values = c(`FALSE` = "#C9D9E8", `TRUE` = "#E8B4AD"), guide = "none") +
  scale_colour_manual(values = c(`FALSE` = "#315A7D", `TRUE` = "#B24A3B"), guide = "none") +
  coord_cartesian(xlim = quantile(clock_error_data$prediction_error_years, c(0.01, 0.99), na.rm = TRUE)) +
  labs(
    x = "Prediction error (predicted − chronological age; years)",
    y = NULL,
    title = "Blood-clock prediction error across tissues (1st–99th percentiles)"
  ) +
  theme_supplementary +
  theme(axis.text.y = element_text(colour = y_axis_colours))

Plot_SFig4_clock_error_distribution
save_plot_object(
  Plot_SFig4_clock_error_distribution,
  "Fig4_supp_clock_error_distribution.rds"
)
rm(Plot_SFig4_clock_error_distribution, clock_error_data)
invisible(gc())
```

# Multi-tissue donors represented in at least three tissues

The source-data workbook permits a broader donor audit than panel F. The heatmap
below includes all donors with predictions in at least three tissues.

```{r fig4-multitissue-donor-errors, fig.width=13, fig.height=8}
donor_tissue_predictions <- clock_predictions |>
  filter(!is.na(donor_id)) |>
  group_by(donor_id, tissue_groups, tissue_label) |>
  summarise(
    chronological_age = median(age_years, na.rm = TRUE),
    predicted_age = mean(predicted_age_scaled, na.rm = TRUE),
    prediction_error = predicted_age - chronological_age,
    n_profiles = n(),
    .groups = "drop"
  )

eligible_donors <- donor_tissue_predictions |>
  count(donor_id, name = "n_tissues") |>
  filter(n_tissues >= 3)

donor_key <- donor_tissue_predictions |>
  semi_join(eligible_donors, by = "donor_id") |>
  group_by(donor_id) |>
  summarise(chronological_age = median(chronological_age, na.rm = TRUE), .groups = "drop") |>
  left_join(eligible_donors, by = "donor_id") |>
  arrange(chronological_age, donor_id) |>
  mutate(donor_label = sprintf("D%02d", row_number()))

multitissue_error_data <- donor_tissue_predictions |>
  semi_join(eligible_donors, by = "donor_id") |>
  left_join(donor_key, by = "donor_id", suffix = c("", "_donor")) |>
  mutate(
    donor_label = factor(donor_label, levels = donor_key$donor_label),
    tissue_label = factor(tissue_label, levels = rev(sort(unique(tissue_label))))
  )

error_limit <- max(abs(multitissue_error_data$prediction_error), na.rm = TRUE)

Plot_SFig4_multitissue_donor_errors <-
  ggplot(multitissue_error_data, aes(x = donor_label, y = tissue_label, fill = prediction_error)) +
  geom_tile(colour = "white", linewidth = 0.2) +
  scale_fill_gradient2(
    low = "#3B6FB6", mid = "white", high = "#B24A3B", midpoint = 0,
    limits = c(-error_limit, error_limit),
    name = "Prediction\nerror (years)"
  ) +
  labs(
    x = sprintf("Donor (ordered by chronological age; n = %d)", nrow(eligible_donors)),
    y = NULL,
    title = "Blood-clock errors in donors represented across at least three tissues"
  ) +
  theme_supplementary +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 5))

Plot_SFig4_multitissue_donor_errors
save_plot_object(
  Plot_SFig4_multitissue_donor_errors,
  "Fig4_supp_multitissue_donor_errors.rds"
)
rm(Plot_SFig4_multitissue_donor_errors)
invisible(gc())
```

```{r fig4-multitissue-donor-key}
donor_key |>
  select(donor_label, donor_id, chronological_age, n_tissues) |>
  kable(
    digits = 1,
    caption = "Donor labels used in the multi-tissue clock-error heatmap"
  )
```

# Output objects

```{r output-manifest}
data.frame(
  object = c(
    "Plot_SFig4_clock_by_tissue",
    "Plot_SFig4_clock_error_distribution",
    "Plot_SFig4_multitissue_donor_errors"
  ),
  rds_file = c(
    "Fig4_supp_clock_by_tissue.rds",
    "Fig4_supp_clock_error_distribution.rds",
    "Fig4_supp_multitissue_donor_errors.rds"
  )
)
```
