---
title: "Figure 4 — Immune-ageing trajectories and clocks across tissues"
date: today
date-format: "DD MMM YYYY"
title-block-banner: true
author: 
  - name: Chen Zhan
    orcid: 0000-0002-4794-8339
    email: chen.zhan@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
format: 
  html:
    theme: minty
    df-print: paged
    code-line-numbers: true
    embed-resources: true
    code-fold: true
    code-fold-default: true
    toc: true
    toc-location: left
    toc-title: "Menu"
knitr:
  opts_chunk: 
    message: false
    warning: false
comments:
  hypothesis: 
    theme: clean
editor: source
abstract: |
  Immune ageing is asynchronous across tissues.
---

# Overview

This report reproduces **Figure 4** (immunological ageing clocks across tissues).
The final figure is a patchwork of the following panels:

- **A** – Linear immunological-ageing distance for composition and gene expression.
- **B** – sccomp 1D credible intervals of the ageing slope per cell type.
- **C** – UpSet plot of differentially expressed (DE) genes shared across cell types.
- **D** – Smoothed ageing velocity across tissues (composition vs. gene expression).
- **E** – Predicted vs. actual age (blood LASSO clock applied to all tissues).
- **F** – Per-donor predicted age across tissues.

All expensive intermediate results are cached to `vignettes/data/figure_ready/` and
`vignettes/data/intermediate/`. Each analysis chunk **reads the saved object by default** and
only recomputes it when the file is absent, so the report renders quickly while
remaining fully reproducible.

The cache filenames retain their historical `Fig3_` prefix for compatibility;
all manuscript-facing figures and supplementary exports use Figure 4.

# Library


```{r publication-paths}
# Portable publication data roots (no HPC absolute paths).
.path_candidates <- c(
  file.path("..", "R", "paths.R"),
  if (requireNamespace("here", quietly = TRUE)) {
    file.path(here::here(), "vignettes", "R", "paths.R")
  },
  file.path("R", "paths.R")
)
.path_candidates <- .path_candidates[!is.na(.path_candidates) & nzchar(.path_candidates)]
.path_file <- .path_candidates[file.exists(.path_candidates)][1]
if (is.na(.path_file)) {
  stop(
    "Cannot find vignettes/R/paths.R. Render from the immuneHealthyBodyMap ",
    "repository root or from vignettes/reports/, or set IMMUNE_HEALTHY_BODY_MAP_ROOT."
  )
}
source(.path_file)
fig_cache_dir <- path_figure_ready()
dir.create(path_intermediate("fig2"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_intermediate("fig4"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_intermediate("fig5"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_figure_ready("fig1"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_figure_ready("fig2"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_figure_ready("fig4"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_figure_ready("fig5"), recursive = TRUE, showWarnings = FALSE)
```

```{r setup-libraries}
# Set your R library path here if needed
# .libPaths removed for portability; use renv/conda or site library
library(here)  # project-root aware paths

# Ensure required output directories exist
required_dirs <- c(
  path_figure_ready("fig4"),
  path_intermediate("fig4")
)
invisible(lapply(required_dirs, function(d) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE, showWarnings = FALSE)
}))

library(tidySummarizedExperiment)
library(HDF5Array)          # loadHDF5SummarizedExperiment()
library(magrittr)
library(tidyverse)
library(tidybulk)           # as_SummarizedExperiment(), keep_abundant(), scale_abundance()
library(sccomp)             # sccomp_predict(), sccomp_test()
library(glue)
library(compositions)       # clr()
library(patchwork)          # figure assembly
library(ggplotify)          # as.ggplot() for the UpSet panel
library(ggupset)            # scale_x_upset()
library(ggnewscale)         # multiple colour scales in one plot
library(RColorBrewer)
library(grid)               # unit()
library(ggplot2)
library(org.Hs.eg.db)       # gene symbol mapping
library(cellNexus)          # get_metadata()
library(dittoSeq)           # colour-blind friendly palette
# NOTE: `glmnet` and `scImmuAging` are loaded on-demand inside the LASSO clock
# chunk (panels E & F) so the report still renders from cache without them.
```

# Setup: theme, palette and helpers

```{r define-theme}
theme_multipanel <-
  theme_bw() +
  theme(
    panel.border = element_blank(),
    axis.line = element_line(size = 0.1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    strip.background = element_blank(),
    axis.title.y = element_text(margin = margin(0, 0, 0, 0), size = 7),
    axis.title.x = element_text(margin = margin(0, 0, 0, 0), size = 7),
    panel.spacing.x = unit(0.1, "lines"),
    axis.text.x = element_text(size = 6),
    axis.text.y = element_text(size = 6),
    strip.text.x = element_text(size = 7),
    strip.text.y = element_text(size = 7),
    legend.key.size = unit(5, 'mm'),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 6),
    strip.clip = "off",
    plot.title = element_text(size = 7),
    axis.line.x = element_line(size = 0.2),
    axis.line.y = element_line(size = 0.2),
    axis.ticks.x = element_line(size = 0.2),
    axis.ticks.y = element_line(size = 0.2)
  )
```

```{r define-palette}
# Fixed colours for the six most-represented tissues; remaining tissues get
# colour-blind friendly dittoSeq colours.
top6 <- c(
  "Blood"         = "#91AD9E",
  "Respiratory"   = "#979771",
  "Cerebral Lobes" = "#B0B1B6",
  "Breast"        = "#C2CEDC",
  "Skin"          = "#D4BAAD",
  "Renal"         = "#E1CCB1"
)

build_tissue_palette_ditto <- function(tissues, top6_map = top6) {
  rest <- setdiff(tissues, names(top6_map))
  pal <- dittoSeq::dittoColors()
  pal <- pal[!pal %in% unname(top6_map)]
  if (length(pal) < length(rest)) pal <- rep_len(pal, length(rest))
  rest_map <- stats::setNames(pal[seq_along(rest)], rest)
  out <- c(rest_map, top6_map)
  out[unique(tissues)]
}

cell_metadata <- get_publication_metadata()
all_tissues <- cell_metadata |> distinct(tissue_groups) |> pull()
tissue_colors <- build_tissue_palette_ditto(all_tissues)
```

```{r define-helpers}
# Convert per-decade predictions into a cumulative "immunological ageing"
# distance along the age axis.
pred_to_age_distance <- function(pred,
                                 scale = TRUE,
                                 center = TRUE,
                                 normalise_by_n_feature = TRUE,
                                 trans_fx = identity,
                                 distance_method = "euclidean",
                                 .abundance,
                                 ...) {

  se <-
    pred |>
    tidybulk::as_SummarizedExperiment(
      .sample = sample_id,
      .transcript = L3,
      .abundance = !!enquo(.abundance)
    ) |>
    mutate(age_decade_continuous = age_decade |> as.character() |> as.integer()) |>
    mutate(age_decade = age_decade |> fct_relevel(as.character(1:9)))

  se_matrix <- se %>% assay() %>% trans_fx() %>% as.matrix() %>% t
  se_matrix[se_matrix %>% is.na()] <- 0  # patch NA with 0

  if (scale || center) {
    se_matrix <- scale(se_matrix, center = center, scale = scale)
  }

  if (distance_method == "euclidean")
    dist_obj <- dist(se_matrix, method = "euclidean")

  if (normalise_by_n_feature)
    dist_obj <- dist_obj / sqrt(ncol(se_matrix))

  se_sample <- se %>% pivot_sample() %>% arrange(age_decade_continuous)
  samps <- se_sample$sample_id
  dm <- dist_obj %>% as.matrix() %>% as.data.frame %>% .[samps, samps]

  # successive pairs along the age axis
  prev <- head(samps, -1)
  nex  <- tail(samps, -1)
  step_dists <- mapply(function(a, b) dm[b, a], prev, nex)
  cum_dists  <- cumsum(step_dists)

  se_sample$step_dist <- c(0, step_dists %>% unname())
  se_sample$cum_dist  <- c(0, cum_dists  %>% unname())

  se_sample <- se_sample %>%
    mutate(
      min_age_decade = min(se$age_decade_continuous),
      max_age_decade = max(se$age_decade_continuous)
    )

  se_sample %>% filter(step_dist != 0 & cum_dist != 0)
}

# Detect local maxima of a numeric vector (used to find ageing "velocity" peaks).
find_peaks <- function(x) {
  n <- length(x)
  if (n == 1) return(TRUE)
  peaks <- rep(FALSE, n)
  if (n >= 3) {
    peaks[2:(n - 1)] <- (x[2:(n - 1)] > x[1:(n - 2)] & x[2:(n - 1)] > x[3:n])
  }
  peaks[1] <- x[1] > x[2]
  peaks[n] <- x[n] > x[n - 1]
  if (all(x == x[1])) peaks[] <- FALSE
  peaks
}

# Interpolate + annotate ageing velocity peaks on a fine age grid.
smooth_ageing_velocity <- function(linear_distance_tissue, fine_grid) {
  linear_distance_tissue %>%
    dplyr::select(-data) %>%
    unnest(distance_df) %>%
    group_by(tissue_groups) %>%
    nest() %>%
    mutate(
      peaks = map(data, ~ {
        d <- .
        d$local_max <- find_peaks(d$step_dist)
        d %>% filter(local_max)
      }),
      main_peak_age = map_dbl(peaks, ~ .x$age_decade_continuous[which.max(.x$step_dist)] %||% NA_real_),
      secondary_peak_age = map2_dbl(
        peaks, main_peak_age,
        ~ {
          if (nrow(.x) < 2) return(NA_real_)
          .x2 <- .x %>% filter(abs(age_decade_continuous - .y) >= 3)
          if (nrow(.x2) == 0) return(NA_real_)
          .x2$age_decade_continuous[which.max(.x2$cum_dist)]
        }
      ),
      interp = map(data, ~ tibble(
        age_decade_continuous = fine_grid,
        step_dist = approx(.x$age_decade_continuous, .x$step_dist, xout = fine_grid, rule = 2)$y,
        cum_dist  = approx(.x$age_decade_continuous, .x$cum_dist,  xout = fine_grid, rule = 2)$y,
        min_age_decade = unique(.x$min_age_decade),
        max_age_decade = unique(.x$max_age_decade)
      ))
    ) |>
    unnest(interp) %>%
    ungroup() %>%
    arrange(main_peak_age) %>%
    mutate(
      tissue_groups = forcats::fct_relevel(
        tissue_groups,
        unique(tissue_groups[order(main_peak_age, decreasing = TRUE)])
      )
    )
}

# Root directory holding the pre-computed brms gene-expression predictions.
# Rebuild with: Rscript vignettes/scripts/rebuild_fig4_gene_predictions.R
# (requires NEW_CELL_TYPE_ROOT; see rebuild_fig4_gene_predictions.md).
new_pred_root <- Sys.getenv("NEW_PRED_DIR", unset = path_intermediate("fig4"))
```

# Gene expression ageing distance

Per-decade gene-expression predictions (brms fixed / total effects) are
computed from NEW_CELL_TYPE targets fits via
`vignettes/scripts/rebuild_fig4_gene_predictions.R` and read from disk here.
They are **not** sccomp outputs. Stomach / missing-tissue extensions are
documented in `rebuild_fig4_gene_predictions.md`.

```{r gene-load-hyp}
# Gene-level monotonic-ageing hypotheses (full table, all cell types).
gene_hyp_df <- readRDS(file.path(path_intermediate("fig4"), "new_gene_hyp_df.rds"))

# Fixed-effect short list (10 strongest genes per cell type; monocytic excluded
# to match the pre-computed prediction object below).
short_list_fixed <-
  gene_hyp_df %>%
  filter(cell_type != "monocytic") %>%
  filter(component == "fixed") %>%
  group_by(cell_type, .feature) %>%
  summarise(n = n(), max_abs_est = max(abs_est, na.rm = TRUE), .groups = "drop") %>%
  group_by(cell_type) %>%
  arrange(desc(n), desc(max_abs_est), .by_group = TRUE) %>%
  slice_head(n = 10) %>%
  ungroup()
```

```{r gene-fixed-effect}
pred_fixed <- readRDS(file.path(new_pred_root, "pred_fixed.rds"))

if (!all(pred_fixed$.feature == short_list_fixed$.feature)) {
  stop("gene list unmatch!")
}

pred_fixed <-
  pred_fixed %>%
  mutate(cell_type = short_list_fixed$cell_type) %>%
  unnest(pred_df) %>%
  group_by(cell_type, .feature, age_decade) %>%
  reframe(
    count_mean  = mean(count_avg, na.rm = TRUE),
    count_lower = unname(quantile(count_avg, prob = 0.025, na.rm = TRUE)),
    count_upper = unname(quantile(count_avg, prob = 0.975, na.rm = TRUE))
  ) %>%
  mutate(
    sample_id = paste0('my_sample_', age_decade),
    L3 = paste0(cell_type, '_', .feature),
    age_decade = factor(age_decade, levels = as.character(1:10))
  )

linear_distance_gene_expr <-
  pred_fixed |>
  pred_to_age_distance(scale = TRUE, trans_fx = log1p, .abundance = count_mean)
```

```{r gene-tissue-total}
# Tissue-level (total effect) short list.
short_list_total <-
  gene_hyp_df %>%
  filter(cell_type != "monocytic") %>%
  filter(component == "total") %>%
  group_by(tissue, cell_type, .feature) %>%
  summarise(n = n(), max_abs_est = max(abs_est, na.rm = TRUE), .groups = "drop") %>%
  group_by(tissue, cell_type) %>%
  arrange(desc(n), desc(max_abs_est), .by_group = TRUE) %>%
  slice_head(n = 10) %>%
  ungroup()

pred_total         <- readRDS(file.path(new_pred_root, "pred_total.rds"))
pred_total_stomach <- readRDS(file.path(new_pred_root, "pred_total_stomach.rds"))
pred_total_missing <- readRDS(file.path(new_pred_root, "pred_total_missing.rds"))

if (!all(pred_total$.feature == short_list_total$.feature)) {
  stop("gene list unmatch!")
}

pred_tissue_gene <-
  pred_total %>%
  mutate(cell_type = short_list_total$cell_type) %>%
  dplyr::select(-tar_group) %>%
  bind_rows(pred_total_stomach %>% dplyr::select(.feature, pred_df, cell_type)) %>%
  unnest(pred_df) %>%
  filter(!tissue_groups %in% unique(pred_total_missing$tissue_groups)) %>%
  bind_rows(
    pred_total_missing %>%
      dplyr::select(-tissue_groups) %>%
      unnest(pred_df) %>%
      dplyr::select(.feature, age_decade, tissue_groups, draws, count_avg, cell_type)
  ) %>%
  group_by(tissue_groups, cell_type, .feature, age_decade) %>%
  reframe(
    count_mean  = mean(count_avg, na.rm = TRUE),
    count_lower = unname(quantile(count_avg, prob = 0.025, na.rm = TRUE)),
    count_upper = unname(quantile(count_avg, prob = 0.975, na.rm = TRUE))
  ) %>%
  mutate(
    sample_id = paste0('my_sample_', age_decade),
    L3 = paste0(cell_type, '_', .feature),
    age_decade = factor(age_decade, levels = as.character(1:9))
  )

linear_distance_tissue_gene <-
  pred_tissue_gene |>
  filter(age_decade %in% as.character(1:9)) %>%
  nest(data = -tissue_groups) |>
  filter(map_int(data, ~ .x |> distinct(age_decade) |> nrow()) > 2) |>
  filter(tissue_groups |> str_detect("trachea|sensory|endocrine", negate = TRUE)) |>
  mutate(distance_df = map(
    data, pred_to_age_distance, trans_fx = log1p, .abundance = count_mean
  ))

smoothed_df_2D_gene_expr <-
  smooth_ageing_velocity(linear_distance_tissue_gene, fine_grid = seq(1, 9, by = 0.05))
```

# Composition ageing distance

Cell-type composition estimates (`sccomp`) are loaded lazily: the raw estimate
object is only read from disk when a downstream cache is missing.

```{r comp-est-helper}
# Lazily load (and memoise) the sccomp estimate object only when needed.
.est_cache <- NULL
get_est <- function() {
  if (is.null(.est_cache)) {
    .est_cache <<- readRDS(
      path_sccomp_estimates("estimates_age_decade___L3___disease_FALSE___immune_only_TRUE.rds")
    )
  }
  .est_cache
}
```

```{r comp-significant-cells}
all_decades <- 1:9

make_contrast_string <- function(k) {
  left  <- paste0("age_decade", all_decades[1:k], collapse = " + ")
  right <- paste0("age_decade", all_decades[(k + 1):length(all_decades)], collapse = " + ")
  sprintf("(%s)/%d - (%s)/%d", right, length(all_decades) - k, left, k)
}

contrast_strings <- sapply(1:(length(all_decades) - 1), make_contrast_string)
contrast_strings <- map_chr(contrast_strings,
                            ~ str_replace_all(.x, "\\b(age_decade\\d+|0)\\b", "`\\1`"))
contrast_strings <- map_chr(contrast_strings, ~ str_replace_all(.x, "`age_decade5`", "0"))
names(contrast_strings) <- paste0("split_at_", all_decades[2:length(all_decades)])

path_my_test <- file.path(path_intermediate("fig4"), "test_contrast_age.rds")

if (file.exists(path_my_test)) {
  my_test <- readRDS(path_my_test)
} else {
  my_test <- get_est() |> sccomp_test(contrasts = contrast_strings)
  saveRDS(my_test, file = path_my_test)
}

cells_of_interest <-
  my_test |> filter(c_pH0 < 0.05) |> dplyr::count(L3) |> arrange(desc(n)) |> pull(L3)
```

```{r comp-linear-distance}
path_sccomp_age_pred <- file.path(path_intermediate("fig4"), "sccomp_age_pred.rds")

if (file.exists(path_sccomp_age_pred)) {
  pred <- readRDS(path_sccomp_age_pred)
} else {
  est <- get_est()
  pred <-
    est |>
    sccomp_predict(
      formula_composition = ~ age_decade * sex + ethnicity_groups_imputed,
      new_data =
        est |>
        attr("count_data") |>
        sccomp:::.subset(sample_id) |>
        distinct(age_decade) |>
        mutate(sex = NA, ethnicity_groups_imputed = NA) |>
        arrange(age_decade) |>
        mutate(sample_id = glue("my_sample_{1:n()}")),
      summary_instead_of_draws = TRUE,
      robust = TRUE
    )
  saveRDS(pred, file = path_sccomp_age_pred)
}

linear_distance_composition <-
  pred |>
  filter(L3 %in% cells_of_interest) |>
  pred_to_age_distance(scale = TRUE, trans_fx = compositions::clr, .abundance = proportion_mean)
```

```{r comp-tissue-distance}
path_sccomp_age_tissue_pred <-
  file.path(path_intermediate("fig4"), "sccomp_age_tissue_pred.rds")

if (file.exists(path_sccomp_age_tissue_pred)) {
  pred_tissue <- readRDS(path_sccomp_age_tissue_pred)
} else {
  est <- get_est()
  pred_tissue <-
    est |>
    sccomp_predict(
      formula_composition = ~ age_decade * sex + ethnicity_groups_imputed +
        (age_decade * sex + ethnicity_groups_imputed | tissue_groups),
      new_data =
        est |>
        attr("count_data") |>
        sccomp:::.subset(sample_id) |>
        distinct(age_decade, tissue_groups) |>
        arrange(age_decade) |>
        mutate(sex = NA, ethnicity_groups_imputed = NA) |>
        mutate(sample_id = glue("my_sample_{1:n()}")) |>
        mutate(sample_id = if_else(
          is.na(sample_id), glue("my_sample_imputed_{row_number()}"), sample_id
        )),
      summary_instead_of_draws = TRUE,
      robust = TRUE
    )
  saveRDS(pred_tissue, file = path_sccomp_age_tissue_pred)
}

linear_distance_tissue_comp <-
  pred_tissue |>
  filter(age_decade %in% as.character(1:9)) %>%
  nest(data = -tissue_groups) |>
  filter(map_int(data, ~ .x |> distinct(age_decade) |> nrow()) > 2) |>
  filter(tissue_groups |> str_detect("trachea|sensory|endocrine", negate = TRUE)) |>
  mutate(distance_df = map(
    data, pred_to_age_distance, scale = TRUE,
    trans_fx = compositions::clr, .abundance = proportion_mean
  ))

smoothed_df_2D_composition <-
  smooth_ageing_velocity(linear_distance_tissue_comp, fine_grid = seq(1, 10, by = 0.05))
```

# Panel A – Linear immunological ageing distance

```{r fig.width=6, fig.height=3, label="panel-A-linear-distance"}
linear_distance_scaled <-
  bind_rows(
    linear_distance_composition %>%
      dplyr::select(age_decade_continuous, cum_dist) %>% mutate(version = 'composition'),
    linear_distance_gene_expr %>%
      dplyr::select(age_decade_continuous, cum_dist) %>% mutate(version = 'gene_expr')
  ) %>%
  group_by(version) %>%
  mutate(cum_dist_scaled = scales::rescale(cum_dist, to = c(0, 10))) %>%
  ungroup()

Plot__linear_distance_all <-
  ggplot(linear_distance_scaled, aes(x = cum_dist_scaled, y = 0, group = version)) +
  geom_line() +
  geom_point(color = "red", size = 3) +
  geom_text(aes(label = age_decade_continuous * 10 - 10), vjust = -1, size = 3) +
  facet_grid(rows = vars(version), switch = "y") +
  theme_minimal() +
  theme(
    strip.placement = "outside",
    strip.text.y.left = element_text(angle = 0, hjust = 1, face = "bold"),
    axis.title.x = element_blank(), axis.text.x = element_blank(), axis.ticks.x = element_blank(),
    axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(),
    panel.grid.major = element_blank(), panel.grid.minor = element_blank()
  )

saveRDS(
  Plot__linear_distance_all,
  file.path(path_figure_ready("fig4"), "Fig3_Plot__linear_distance_all.rds")
)

Plot__linear_distance_all
```

# Panel D – Smoothed ageing velocity across tissues

```{r fig.width=12, fig.height=9, label="panel-D-smoothed-velocity"}
smoothed_df_2D_all <-
  bind_rows(
    smoothed_df_2D_composition %>% mutate(version = 'composition'),
    smoothed_df_2D_gene_expr   %>% mutate(version = 'gene_expr')
  ) %>%
  filter(!tissue_groups %in% c('digestive system (general)')) %>%
  mutate(tissue_version = glue::glue('{tissue_groups} ({version})')) %>%
  group_by(version) %>%
  mutate(step_dist_scaled = scales::rescale(step_dist, to = c(0, 10))) %>%
  ungroup() %>%
  group_by(tissue_groups) %>%
  mutate(min_main_peak = min(main_peak_age, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(-min_main_peak, tissue_groups)

version_order <- c("gene_expr", "composition")
pin_first     <- c("thymus", "blood")   # pinned to the bottom of the axis

tissue_order <-
  smoothed_df_2D_all %>%
  group_by(tissue_groups) %>%
  summarise(min_main_peak = min(main_peak_age, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(min_main_peak), tissue_groups) %>%
  pull(tissue_groups) %>%
  as.character()

tissue_order <- c(
  tissue_order[!tissue_order %in% pin_first],
  pin_first[pin_first %in% tissue_order]
)

levels_interleaved <-
  as.vector(unlist(map(tissue_order, ~ paste0(.x, " (", version_order, ")"))))

smoothed_df_2D_all <-
  smoothed_df_2D_all %>%
  mutate(tissue_version = factor(tissue_version, levels = levels_interleaved))

Plot__smoothed_age_dist_all <-
  smoothed_df_2D_all %>%
  filter(between(age_decade_continuous, min_age_decade, max_age_decade)) |>
  ggplot(aes(x = age_decade_continuous * 10, y = tissue_version)) +
  geom_line(
    data = ~ filter(.x, version == "composition"),
    aes(color = step_dist_scaled, linewidth = step_dist_scaled)
  ) +
  scale_color_gradientn(colors = rev(RColorBrewer::brewer.pal(11, "RdYlBu")), name = "composition") +
  ggnewscale::new_scale_color() +
  geom_line(
    data = ~ filter(.x, version == "gene_expr"),
    aes(color = step_dist_scaled, linewidth = step_dist_scaled)
  ) +
  scale_color_gradientn(colors = rev(RColorBrewer::brewer.pal(11, "RdYlBu")), name = "gene_expr") +
  scale_linewidth(range = c(0.5, 4), guide = "none") +
  geom_point(aes(x = main_peak_age * 10), color = "black", size = 0.9) +
  geom_point(aes(x = secondary_peak_age * 10), shape = 12, size = 0.6, color = "black") +
  scale_y_discrete(limits = levels_interleaved) +
  theme_minimal() +
  labs(x = "Age years", y = "Tissue (version)") +
  theme(legend.position = "right")

saveRDS(
  Plot__smoothed_age_dist_all,
  file.path(path_figure_ready("fig4"), "Fig3_Plot__smoothed_age_dist_all.rds")
)

Plot__smoothed_age_dist_all
```

# Panel B – sccomp 1D credible intervals

```{r fig.width=7, fig.height=7, label="panel-B-sccomp-1d"}
path_sccomp_1d_colored <-
  file.path(path_figure_ready("fig4"), "Fig3_Plot__sccomp_1d_colored.rds")

if (file.exists(path_sccomp_1d_colored)) {
  Plot__sccomp_1d_colored <- readRDS(path_sccomp_1d_colored)
} else {
  significance_threshold <- 0.05

  plot_df <-
    my_test |>
    filter(L3 != "cytotoxic") |>
    mutate(
      abs_c_effect = abs(c_effect),
      parameter = str_replace_all(parameter, "_", " ")
    ) |>
    group_by(L3) |>
    slice_max(abs_c_effect, n = 1, with_ties = FALSE) |>
    ungroup() |>
    mutate(
      sig = c_pH0 < significance_threshold,
      L3 = fct_reorder(L3, c_effect)
    )

  split_cols <- colorRampPalette(brewer.pal(11, "Spectral")[6:11])(8)
  names(split_cols) <- paste0("split at ", 2:9)

  Plot__sccomp_1d_colored <-
    ggplot(plot_df, aes(x = c_effect, y = L3, colour = parameter)) +
    geom_vline(xintercept = 0, colour = "grey55", linewidth = 0.4) +
    geom_errorbarh(aes(xmin = c_lower, xmax = c_upper), height = 0.25, linewidth = 0.8) +
    scale_colour_manual(values = split_cols, labels = paste0("split_at_", 2:9)) +
    geom_point(size = 2.2) +
    labs(x = "Credible interval of the slope", y = "Cell group", colour = "Split decade") +
    theme_multipanel

  saveRDS(Plot__sccomp_1d_colored, path_sccomp_1d_colored)
}

Plot__sccomp_1d_colored
```

# Panel C – UpSet plot of shared DE genes

```{r fig.width=8, fig.height=5, label="panel-C-upset"}
path_upset <- file.path(path_figure_ready("fig4"), "Fig3_Plot__upset.rds")

if (file.exists(path_upset)) {
  Plot__upset <- readRDS(path_upset)
} else {
  # DE gene sets per cell type (fixed component; drop cytotoxic and NA cell types)
  gene_sets <-
    gene_hyp_df %>%
    filter(component == "fixed", !is.na(cell_type), cell_type != 'cytotoxic') %>%
    group_by(.feature) %>%
    summarise(ct_set = list(sort(unique(cell_type))), .groups = "drop") %>%
    mutate(n_ct = lengths(ct_set))

  Plot__upset <-
    ggplot(gene_sets, aes(x = ct_set)) +
    geom_bar() +
    scale_x_upset(n_intersections = 50) +
    labs(y = "Number of DE genes", x = NULL) +
    theme_minimal() +
    theme(panel.grid = element_blank())

  saveRDS(Plot__upset, path_upset)
}

Plot__upset
```

# Panels E & F – Blood ageing clock applied to all tissues

Pre-trained blood LASSO clocks (`scImmuAging`) are applied to pseudobulk
profiles of the matched cell-type families across all tissues. The predicted-age
plot and the per-donor summary are cached; the full LASSO pipeline only runs
when the cache is absent.

```{r define-clock-inputs}
cell_type_list <- list(
  CD4T = c("cd4 naive", "cd4 tcm", "cd4 th17 em", "cd4 th1/th17 em",
           "cd4 th2 em", "cd4 fh em", "cd4 tem", "cd4 th1 em", "t cd4"),
  CD8T = c("cd8 naive", "cd8 tcm", "cd8 tem", "t cd8"),
  MONO = c("cd14 mono", "cd16 mono", "monocytic"),
  NK   = c("nk", "nkt"),
  B    = c("b naive", "b memory", "b")
)

path_pred_whole_plot <-
  file.path(path_figure_ready("fig4"), "Fig3_Plot__pred_whole.rds")
path_total_pred_summary <-
  file.path(path_intermediate("fig4"), "Fig3_total_pred_summary.rds")
```

```{r run-age-clock}
if (file.exists(path_pred_whole_plot) && file.exists(path_total_pred_summary)) {

  Plot__pred_whole   <- readRDS(path_pred_whole_plot)
  total_pred_summary <- readRDS(path_total_pred_summary)

} else {

  library(glmnet)       # cv.glmnet() LASSO age clock
  library(scImmuAging)  # pre-trained blood ageing clocks

  # --- Load pseudobulk cell data (HDF5-backed SummarizedExperiment) ----------
  se <- loadHDF5SummarizedExperiment(Sys.getenv("AGE_CLOCK_SE_DIR", unset = file.path(publication_data_root(), "zenodo_release", "age_clock_se")))

  # Donors profiled in >5 tissues (healthy only) for the per-donor panel.
  donor_select <-
    se %>% colData() %>% as.data.frame() %>% rownames_to_column() %>%
    filter(disease_groups == 'Normal') %>%
    dplyr::select(donor_id, tissue_groups, age_days) %>%
    distinct() %>%
    group_by(donor_id, age_days) %>%
    summarise(n = n_distinct(tissue_groups), .groups = "drop") %>%
    filter(n > 5)

  # --- Pre-trained blood clocks ---------------------------------------------
  feature_set <- readRDS(system.file("data", "all_model_inputfeatures.RDS", package = "scImmuAging"))

  # Apply one blood clock to all tissues, keeping the central 1-99% predictions.
  run_age_clock <- function(test_model) {
    se_cur <- se %>% filter(cell_type_unified_ensemble %in% cell_type_list[[test_model]])
    se_cur <- se_cur %>% keep_abundant() %>% scale_abundance(method = 'TMMwsp')
    se_cur <- se_cur[intersect(feature_set[[test_model]], rownames(se_cur)), ]

    data <- se_cur %>% assay('counts_scaled') %>% t %>% as.data.frame()
    data[, setdiff(feature_set[[test_model]], rownames(se_cur))] <- 0
    meta <- se_cur %>% colData() %>% as.data.frame() %>% rownames_to_column()

    data <- left_join(
      data %>% rownames_to_column(),
      meta %>% dplyr::select('rowname', 'tissue_groups', 'age_days')
    ) %>% column_to_rownames('rowname')

    blood_data <- data %>% filter(tissue_groups == 'blood')

    model <- cv.glmnet(
      x = as.matrix(blood_data %>% dplyr::select(-c('tissue_groups', 'age_days'))),
      y = as.matrix(blood_data %>% dplyr::select(age_days)),
      type.measure = "mae", standardize = FALSE, relax = FALSE, nfolds = 5
    )

    pred <- predict(
      model,
      newx = data %>% dplyr::select(-c('tissue_groups', 'age_days')) %>% as.matrix(),
      s = "lambda.min"
    )
    colnames(pred) <- 'predicted_age'

    pred %>%
      as.data.frame() %>% rownames_to_column() %>% left_join(meta) %>%
      filter(between(
        predicted_age,
        quantile(predicted_age, 0.01, na.rm = TRUE),
        quantile(predicted_age, 0.99, na.rm = TRUE)
      ))
  }

  pred_list  <- lapply(c("CD8T", "CD4T", "MONO", "NK", "B"), run_age_clock)
  pred_whole <- bind_rows(pred_list)
  total_pred <- pred_whole %>% filter(donor_id %in% donor_select$donor_id)

  # Cross-figure export used by Fig 4.
  saveRDS(pred_whole,
          file = file.path(path_intermediate("fig4"), "Fig4_pred_whole.rds"))

  # Panel E: predicted vs actual age (blood highlighted).
  Plot__pred_whole <-
    pred_whole %>%
    mutate(age_years = age_days / 365, predicted_age_scaled = predicted_age / 365) %>%
    ggplot(aes(x = age_years, y = predicted_age_scaled, color = tissue_groups)) +
    geom_point(
      data = ~ filter(.x, tissue_groups == "blood"),
      alpha = 0.15, size = 0.6, shape = 16, color = "grey50"
    ) +
    geom_smooth(
      data = ~ filter(.x, tissue_groups != "blood"),
      method = "lm", se = FALSE, linewidth = 0.4, alpha = 0.2
    ) +
    geom_smooth(
      data = ~ filter(.x, tissue_groups == "blood"),
      method = "lm", se = FALSE, linewidth = 1.2
    ) +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +
    scale_color_manual(values = tissue_colors) +
    labs(x = "Actual Age", y = "Predicted Age", color = "Tissue Group") +
    coord_fixed() +
    theme_multipanel +
    expand_limits(x = 0, y = 0) +
    theme(legend.position = "right")

  # Panel F data: per-donor / per-tissue predicted-age summary.
  total_pred_summary <-
    total_pred %>%
    group_by(donor_id, tissue_groups) %>%
    reframe(
      predicted_age_mean  = mean(predicted_age, na.rm = TRUE),
      predicted_age_upper = max(predicted_age, na.rm = TRUE),
      predicted_age_lower = min(predicted_age, na.rm = TRUE),
      actual_age = unique(age_days),
      n_s = n_distinct(sample_id)
    )

  saveRDS(Plot__pred_whole,   path_pred_whole_plot)
  saveRDS(total_pred_summary, path_total_pred_summary)
}
```

```{r fig.width=8, fig.height=6, label="panel-E-pred-whole"}
Plot__pred_whole
```

```{r fig.width=10, fig.height=6, label="panel-F-donor-lines"}
Plot__total_pred_line_chart <-
  total_pred_summary %>%
  ggplot(aes(x = predicted_age_mean, y = 0)) +
  geom_line(aes(group = donor_id)) +
  geom_point(aes(color = tissue_groups, size = n_s)) +
  geom_point(aes(x = actual_age, y = 0), shape = 8, color = "black", size = 3) +
  scale_color_manual(values = tissue_colors) +
  facet_wrap(~ donor_id, ncol = 1, strip.position = "left") +
  scale_x_continuous(labels = function(x) ceiling(x / 365), name = "Predicted Age (years)") +
  labs(x = "Age", y = "Donors",
       title = "Predicted Age by Tissue Group per Donor (Black Star = Actual Age)") +
  theme_multipanel +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.line.y = element_blank(),
    strip.text.y.left = element_text(angle = 0),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

saveRDS(
  Plot__total_pred_line_chart,
  file.path(path_figure_ready("fig4"), "Fig3_Plot__total_pred_line_chart.rds")
)

Plot__total_pred_line_chart
```

# Final assembly

The six panels are stitched into the final Figure 4 with `patchwork` and
exported as a PDF sized for print (183 mm wide).

```{r fig.width=10, fig.height=10, label="fig4-assembly"}
# --- Panel A (revised for compact multipanel layout) ------------------------
Plot__linear_distance_all_revised <-
  linear_distance_scaled %>%
  ggplot(aes(x = cum_dist_scaled, y = 0, group = version)) +
  geom_line() +
  geom_point(color = "red", size = 2) +
  geom_text(aes(label = age_decade_continuous * 10 - 10), vjust = -1, size = 2) +
  facet_grid(rows = vars(version), switch = "y") +
  theme_multipanel +
  theme(
    plot.title = element_blank(),
    axis.ticks.x.bottom = element_blank(), axis.ticks.y.left = element_blank(),
    axis.line.x.bottom = element_blank(), axis.line.y.left = element_blank(),
    axis.text.x.bottom = element_blank(), axis.text.y.left = element_blank(),
    axis.title.x.bottom = element_blank(), axis.title.y.left = element_blank(),
    strip.text.y.left = element_text(size = 7, margin = margin(0, 0, 0, 0, "pt"),
                                     angle = 0, hjust = 1),
    strip.placement = "inside"
  )

# --- Panel B (revised) ------------------------------------------------------
Plot__sccomp_1d_revised <-
  Plot__sccomp_1d_colored +
  theme_multipanel +
  theme(
    plot.title = element_blank(),
    axis.ticks.x.bottom = element_blank(),
    axis.ticks.y.left = element_blank()
  ) &
  guides(
    color    = guide_legend(nrow = 1, keywidth = unit(4, "mm"), keyheight = unit(4, "mm"), override.aes = list(size = 3)),
    fill     = guide_legend(nrow = 1, keywidth = unit(4, "mm"), keyheight = unit(4, "mm"), override.aes = list(size = 3)),
    shape    = guide_legend(nrow = 1, keywidth = unit(4, "mm"), keyheight = unit(4, "mm"), override.aes = list(size = 3)),
    linetype = guide_legend(nrow = 1, keywidth = unit(4, "mm"), keyheight = unit(4, "mm"), override.aes = list(size = 3)),
    size     = guide_legend(nrow = 1, keywidth = unit(4, "mm"), keyheight = unit(4, "mm"), override.aes = list(size = 3)),
    alpha    = guide_legend(nrow = 1, keywidth = unit(4, "mm"), keyheight = unit(4, "mm"), override.aes = list(size = 3))
  )

row1_left_col <-
  (wrap_elements(Plot__linear_distance_all_revised) / Plot__sccomp_1d_revised) +
  plot_layout(heights = c(1, 2)) &
  theme(
    panel.spacing = unit(0, "pt"),
    plot.margin = margin(0, 0, 0, 0, "pt"),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.text.position = "bottom"
  )

# --- Row 1: (A over B) beside C (UpSet) -------------------------------------
row1 <-
  (row1_left_col | as.ggplot(Plot__upset)) +
  plot_layout(widths = c(2, 3)) &
  theme(
    panel.spacing = unit(0, "pt"),
    plot.margin = margin(0, 0, 0, 0, "pt"),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.text.position = "bottom"
  )

# --- Row 3 right column: E over F -------------------------------------------
Plot__pred_whole_no_legend <-
  Plot__pred_whole +
  labs(x = "Predicted Age (years)") +
  theme(
    axis.line.x = element_line(),
    axis.text.x = element_text(size = 6),
    axis.title.x = element_text(size = 7)
  )

Plot__total_pred_line_chart_no_legend <-
  total_pred_summary %>%
  ggplot(aes(x = predicted_age_mean, y = 0)) +
  geom_line(aes(group = donor_id), linewidth = 0.25) +
  geom_point(aes(color = tissue_groups, size = n_s)) +
  geom_point(aes(x = actual_age, y = 0), shape = 8, color = "black", size = 1) +
  scale_color_manual(values = tissue_colors) +
  scale_size_continuous(range = c(0.25, 2)) +
  facet_wrap(~ donor_id, ncol = 1, strip.position = "left") +
  scale_x_continuous(labels = function(x) ceiling(x / 365), name = "Predicted Age (years)") +
  labs(y = "Donors") +
  theme_multipanel +
  guides(colour = "none", color = "none", fill = "none", shape = "none",
         linetype = "none", size = "none", alpha = "none") +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.line.y = element_blank(),
    strip.text.y.left = element_text(angle = 0)
  )

row3_right_col <-
  (Plot__pred_whole_no_legend / Plot__total_pred_line_chart_no_legend) +
  plot_layout(heights = c(1, 1)) &
  theme(
    panel.spacing = unit(0, "pt"),
    plot.margin = margin(0, 0, 0, 0, "pt"),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.text.position = "bottom"
  )

# --- Row 3: D beside (E over F) ---------------------------------------------
Plot__smoothed_age_dist_all_no_legend <-
  Plot__smoothed_age_dist_all +
  scale_linewidth(range = c(0.2, 1.5), guide = "none") +
  guides(colour = "none", color = "none", fill = "none", shape = "none",
         linetype = "none", size = "none", alpha = "none")

row3 <-
  (Plot__smoothed_age_dist_all_no_legend | row3_right_col) +
  plot_layout(widths = c(3, 1)) &
  theme(
    panel.spacing = unit(0, "pt"),
    plot.margin = margin(0, 0, 0, 0, "pt"),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.text.position = "bottom"
  )

# --- Final figure -----------------------------------------------------------
Fig4 <-
  (row1 / row3) +
  patchwork::plot_layout(heights = c(3, 3), guides = "collect", nrow = 2) &
  theme(
    panel.spacing = unit(0, "pt"),
    plot.margin = margin(0, 0, 0, 0, "pt"),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.text.position = "bottom",
    legend.key.size = unit(5, 'mm'),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 6)
  )

ggsave(
  plot = Fig4,
  filename = file.path(path_figure_ready("fig4"), "fig4.pdf"),
  units = "mm", width = 183, height = 180, dpi = 500, scale = 1
)

Fig4
```

# Supplementary tables

Export plot statistics from ggplot / heatmap `@data` slots for all Fig 4 panels.

```{r export-supplementary-tables}
message(
  "Panel caches updated. Run `Rscript report/build_supplementary_workbooks.R` ",
  "to rebuild the split-tab workbooks with separate plot-data, model-summary and test sheets."
)
```

# Save panels

```{r save-all-plots, eval=FALSE}
# Persist all panel objects for downstream figure assembly / re-use.
save(
  list = ls(pattern = "Plot__"),
  file = file.path(path_figure_ready("fig4"), "Fig4_plot_data_all.RData")
)

sessionInfo()
```

# Appendix: provenance of pre-computed inputs {.appendix}

The following block documents how `new_gene_hyp_df.rds` was assembled (adding the
`monocytic` cell type from the `targets` modelling store). It is disabled by
default because it depends on the full modelling pipeline.

```{r gene-hyp-provenance, eval=FALSE}
library(targets)

root_path <- Sys.getenv("NEW_CELL_TYPE_ROOT", unset = NA_character_)
.x <- 'monocytic'

adjust_tissue <- tar_meta(
  contains('adjust_tissue'),
  store = glue::glue('{root_path}/V1_{make.names(.x)}/_targets/')
) %>% filter(type == 'pattern') %>% pull(name) %>% unique

adjust_tissue <- tar_read_raw(
  adjust_tissue, store = glue::glue('{root_path}/V1_{make.names(.x)}/_targets/')
)

gene_hyp_missing <-
  adjust_tissue %>% dplyr::select(-contains('tissue')) %>%
  unnest(hypothesis_age_monotonic) %>%
  mutate(star = sign(ci_lower) == sign(ci_upper)) %>%
  mutate(cell_type = .x, abs_est = abs(estimate)) %>%
  filter(star)

gene_hyp_df <-
  readRDS(file.path(path_intermediate("fig4"), "new_gene_hyp_df.rds")) %>%
  bind_rows(gene_hyp_missing)

saveRDS(gene_hyp_df,
        file = file.path(path_intermediate("fig4"), "new_gene_hyp_df.rds"))
```
