---
title: "Figure 5 supplementary plots"
date: today
date-format: "DD MMM YYYY"
format:
  html:
    theme: minty
    toc: true
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
---

# Purpose

This report expands panels that the main figure truncates for display:

1. donor coverage across all mapped tissue groups;
2. body-level sex-difference ageing contrasts (later−earlier, split at age
   decade 5) for every tested cell type;
3. the full tissue-by-cell-type sex-effect heatmap;
4. sex×age-split boxplots for **all** age-trend × sex-gap quadrants (Fig. 5b
   shows only the four lowest-FDR effects overall); and
5. plasma-cell sex trajectories for **all** tissues with plasma predictions
   (no coverage filter; Fig. 5d still uses a stricter subset).

On render, each ggplot object is written to
`vignettes/supplementary_plot_objects/`.

```{r setup}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(here)
library(forcats)
library(stringr)
library(knitr)
library(ggnewscale)
library(patchwork)
library(ggh4x)
library(purrr)
library(readr)

source(file.path("..", "R", "paths.R"))
workbook <- path_source_tables("Supplementary_File_Figure_5.xlsx")
if (!file.exists(workbook)) stop("Figure 5 supplementary file not found: ", workbook)

plot_object_dir <- file.path(publication_vignette_root(), "supplementary_plot_objects")
dir.create(plot_object_dir, recursive = TRUE, showWarnings = FALSE)

read_panel_sheet <- function(sheet, path = workbook) {
  if (!sheet %in% excel_sheets(path)) stop("Missing worksheet: ", sheet, " in ", path)
  # Submission workbooks store a caption in row 1 and headers in row 2.
  read_xlsx(path, sheet = sheet, skip = 1, guess_max = 1000000L)
}

save_plot_object <- function(plot, filename) {
  path <- file.path(plot_object_dir, filename)
  saveRDS(plot, path, compress = "xz")
  message("Saved ggplot object: ", path)
  invisible(path)
}

clean_label <- function(x) {
  x <- gsub("_", " ", x, fixed = TRUE)
  tools::toTitleCase(x)
}

# Same asin-sqrt axis transform as Fig2 supplementary proportion plots:
# keep raw proportions in the data; transform only the scale.
asin_sqrt_trans <- scales::trans_new(
  name = "asin_sqrt",
  transform = function(x) asin(sqrt(pmax(pmin(x, 1), 0))),
  inverse = function(y) sin(y)^2
)

# Match Fig5.qmd theme_multipanel exactly for boxplots and line charts.
theme_multipanel <-
  theme_bw() +
  theme(
    panel.border = element_blank(),
    axis.line = element_line(size = 0.1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    strip.background = element_blank(),
    axis.title.y = element_text(margin = margin(0, 0, 0, 0), size = 7),
    axis.title.x = element_text(margin = margin(0, 0, 0, 0), size = 7),
    panel.spacing.x = unit(0.1, "lines"),
    axis.text.x = element_text(size = 6),
    axis.text.y = element_text(size = 6),
    strip.text.x = element_text(size = 7),
    strip.text.y = element_text(size = 7),
    legend.key.size = unit(5, "mm"),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 6),
    strip.clip = "off",
    plot.title = element_text(size = 7),
    axis.line.x = element_line(size = 0.2),
    axis.line.y = element_line(size = 0.2),
    axis.ticks.x = element_line(size = 0.2),
    axis.ticks.y = element_line(size = 0.2)
  )

# Lighter theme retained only for coverage / heatmap summary panels.
theme_supplementary <- theme_bw(base_size = 8.5) +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_blank(),
    legend.position = "bottom",
    plot.title.position = "plot"
  )

# Same single-line abbreviations as Fig5.qmd boxplot facets.
cell_type_abbrev_dict <- c(
  "b memory" = "B Memory", "b naive" = "B Naive",
  "cd14 mono" = "CD14 Mono", "cd16 mono" = "CD16 Mono",
  "cd4 fh em" = "CD4 FH EM", "cd4 naive" = "CD4 Naive", "cd4 tcm" = "CD4 TCM",
  "cd4 th1 em" = "CD4 Th1 EM", "cd4 th1/th17 em" = "CD4 Th1/Th17 EM",
  "cd4 th17 em" = "CD4 Th17 EM", "cd4 th2 em" = "CD4 Th2 EM", "treg" = "Treg",
  "cd8 naive" = "CD8 Naive", "cd8 tcm" = "CD8 TCM", "cd8 tem" = "CD8 TEM",
  "cdc" = "CDC", "granulocyte" = "Gran", "ilc" = "ILC", "macrophage" = "Mac",
  "mait" = "MAIT", "mast" = "Mast", "nk" = "NK", "nkt" = "NKT", "pdc" = "PDC",
  "tgd" = "TGD", "plasma" = "Plasma"
)

tissue_name_dict <- c(
  "blood" = "Blood",
  "sensory-related structures" = "Sensory",
  "cerebral lobes and cortical areas" = "Cerebral cortex",
  "limbic and basal systems" = "Limbic/basal",
  "liver" = "Liver",
  "bone marrow" = "Bone marrow",
  "large intestine" = "Large int",
  "lymphatic system" = "Lymphatic",
  "small intestine" = "Small int",
  "respiratory system" = "Respiratory",
  "renal system" = "Renal",
  "brainstem and cerebellar structures" = "Hindbrain",
  "cardiovascular system" = "Cardiovascular",
  "general brain and major structures" = "Brain",
  "vasculature" = "Vasculature",
  "ovary" = "Ovary",
  "integumentary system (skin)" = "Skin",
  "nasal, oral, and pharyngeal regions" = "Pharyngeal",
  "muscular system (skeletal muscles)" = "Skeletal muscle",
  "prostate" = "Prostate",
  "breast" = "Breast",
  "digestive system (general)" = "Digestive",
  "oesophagus" = "Oesophagus",
  "female reproductive system" = "Female repro",
  "spleen" = "Spleen",
  "peritoneal and abdominal cavity structures" = "Peritoneal",
  "thymus" = "Thymus",
  "adipose tissue" = "Adipose",
  "trachea" = "Trachea",
  "endocrine system" = "Endocrine",
  "pancreas" = "Pancreas",
  "miscellaneous glands" = "Glands",
  "stomach" = "Stomach",
  "gallbladder" = "Gallbladder"
)

colmap_sex <- c("female" = "#D7907B", "male" = "#0A014F")

resolve_fig5_cache <- function(filename) {
  candidates <- c(
    path_intermediate("fig5", filename),
    {
      arch <- path_age_clock_archive("data", filename)
      if (is.null(arch)) character() else arch
    },
    {
      arch <- path_age_clock_archive("data", "age_fig", filename)
      if (is.null(arch)) character() else arch
    }
  )
  hit <- candidates[nzchar(candidates) & file.exists(candidates)][1]
  if (is.na(hit)) {
    stop(
      "Missing Fig5 cache '", filename, "'. Expected under path_intermediate('fig5') ",
      "or AGE_CLOCK_ARCHIVE/data/.",
      call. = FALSE
    )
  }
  hit
}

format_fdr <- function(x) {
  case_when(
    is.na(x) ~ "NA",
    x < 0.001 ~ gsub("^0\\.", ".", formatC(x, format = "e", digits = 2)),
    TRUE ~ gsub("^0\\.", ".", formatC(x, format = "f", digits = 3))
  )
}
```

```{r read-source-data}
donor_counts <- read_panel_sheet("Fig5_A_donor_counts")
sex_tests <- read_panel_sheet("Fig5_B_sex_tests")
tissue_effects <- read_panel_sheet("Fig5_C_tissue_effects")

# Prefer the compact later−earlier split-at-5 table used for Fig. 5b / S23.
resolve_split_at_5 <- function() {
  candidates <- c(
    path_intermediate("fig5", "fig5_test_sex_diff_split_at_5.csv"),
    {
      arch <- path_age_clock_archive("data", "age_fig", "fig5_test_sex_diff_split_at_5.csv")
      if (is.null(arch)) character() else arch
    },
    path_intermediate("fig5", "Fig4_split_at_5_compact.rds"),
    {
      arch <- path_age_clock_archive("data", "age_fig", "Fig4_split_at_5_compact.rds")
      if (is.null(arch)) character() else arch
    }
  )
  hit <- candidates[nzchar(candidates) & file.exists(candidates)][1]
  if (is.na(hit)) {
    stop(
      "Missing fig5_test_sex_diff_split_at_5.csv under path_intermediate('fig5') ",
      "or AGE_CLOCK_ARCHIVE/data/age_fig/.",
      call. = FALSE
    )
  }
  if (grepl("\\.csv$", hit, ignore.case = TRUE)) {
    readr::read_csv(hit, show_col_types = FALSE) %>%
      mutate(
        sig = if ("sig" %in% names(.)) as.logical(sig) else c_FDR < 0.05,
        sign_age_trend = as.integer(sign_age_trend)
      )
  } else {
    readRDS(hit)
  }
}

test_sex_diff_split_at_5 <- resolve_split_at_5() %>%
  mutate(
    selection_direction = case_when(
      sign_age_trend > 0 & c_effect > 0 ~ "age_up_male_more",
      sign_age_trend > 0 & c_effect < 0 ~ "age_up_female_more",
      sign_age_trend < 0 & c_effect > 0 ~ "age_down_female_more",
      sign_age_trend < 0 & c_effect < 0 ~ "age_down_male_more",
      TRUE ~ "other"
    ),
    same_signed = sign(sign_age_trend) == sign(c_effect) & sign_age_trend != 0
  )
```

# Complete age-by-sex cohort coverage

```{r fig5-full-cohort-coverage, fig.width=8, fig.height=12}
coverage_data <- donor_counts |>
  transmute(
    age_groups = .data[["Age stratum"]],
    sex = .data[["Sex"]],
    tissue_label = coalesce(
      na_if(.data[["Figure tissue label"]], ""),
      .data[["Tissue group"]]
    ),
    donor_count = .data[["Number of unique donors"]],
    displayed_main = .data[["Displayed in Fig. 5a"]] == "Yes"
  ) |>
  filter(!is.na(tissue_label), tissue_label != "Unmapped tissue") |>
  mutate(
    sex = factor(
      str_to_lower(sex),
      levels = c("female", "male"),
      labels = c("Female", "Male")
    ),
    age_groups = factor(age_groups, levels = c("Young", "Mid", "Old"))
  )

tissue_order <- coverage_data |>
  group_by(tissue_label) |>
  summarise(total = sum(donor_count), .groups = "drop") |>
  arrange(total) |>
  pull(tissue_label)

coverage_data <- coverage_data |>
  mutate(tissue_label = factor(tissue_label, levels = tissue_order))

unmapped_n <- donor_counts |>
  filter(.data[["Tissue group"]] == "Unmapped tissue") |>
  summarise(n = sum(.data[["Number of unique donors"]], na.rm = TRUE)) |>
  pull(n)
if (length(unmapped_n) == 0 || is.na(unmapped_n)) unmapped_n <- 0

Plot_SFig5_full_cohort_coverage <-
  ggplot(coverage_data, aes(x = age_groups, y = tissue_label, fill = donor_count)) +
  geom_tile(colour = "white", linewidth = 0.25) +
  geom_text(aes(label = ifelse(donor_count == 0, "", donor_count)), size = 2.2) +
  facet_grid(. ~ sex) +
  scale_fill_viridis_c(trans = "sqrt", option = "C", name = "Unique donors") +
  labs(
    x = "Age group",
    y = NULL,
    title = "Donor coverage across all mapped tissues, age groups and sexes",
    subtitle = sprintf(
      "All mapped tissues are shown; %d donor records are unmapped",
      unmapped_n
    )
  ) +
  theme_supplementary +
  theme(panel.grid.major = element_blank())

Plot_SFig5_full_cohort_coverage
save_plot_object(
  Plot_SFig5_full_cohort_coverage,
  "Fig5_supp_full_cohort_coverage.rds"
)
```

# Body-level sex-difference ageing contrasts (split at age decade 5)

```{r fig5-all-global-sex-effects, fig.width=9, fig.height=8}
global_effect_data <- test_sex_diff_split_at_5 |>
  filter(!is.na(cell_type_unified_ensemble)) |>
  mutate(
    cell_label = recode(cell_type_unified_ensemble, !!!cell_type_abbrev_dict),
    cell_label = if_else(
      cell_label == cell_type_unified_ensemble,
      clean_label(cell_type_unified_ensemble),
      cell_label
    ),
    supported = !is.na(c_FDR) & c_FDR < 0.05,
    sign_class = if_else(same_signed, "Same-signed with S4", "Opposite-signed vs S4")
  ) |>
  arrange(c_effect) |>
  mutate(cell_label = factor(cell_label, levels = unique(cell_label)))

global_effect_limit <- max(abs(global_effect_data$c_effect), na.rm = TRUE)

Plot_SFig5_all_global_sex_effects <-
  ggplot(
    global_effect_data,
    aes(x = c_effect, y = cell_label, colour = sign_class)
  ) +
  geom_vline(xintercept = 0, linewidth = 0.3, linetype = "dashed") +
  geom_errorbarh(
    aes(xmin = c_lower, xmax = c_upper),
    height = 0.25,
    linewidth = 0.35,
    alpha = 0.7
  ) +
  geom_point(aes(shape = supported), size = 2) +
  scale_colour_manual(
    values = c(
      "Same-signed with S4" = "#0A014F",
      "Opposite-signed vs S4" = "#D7907B"
    ),
    name = NULL
  ) +
  scale_shape_manual(
    values = c(`TRUE` = 16, `FALSE` = 1),
    labels = c(`TRUE` = "FDR < 0.05", `FALSE` = "FDR ≥ 0.05"),
    name = NULL
  ) +
  coord_cartesian(xlim = c(-global_effect_limit, global_effect_limit) * 1.05) +
  labs(
    x = "Sex-difference ageing contrast\n(later−earlier males) − (later−earlier females)",
    y = NULL,
    title = "Body-level sex-difference ageing effects (split at age decade 5)",
    subtitle = "Filled points denote FDR < 0.05; Fig. 5b highlights overall top-4 by FDR"
  ) +
  theme_supplementary +
  theme(panel.grid.major.y = element_blank())

Plot_SFig5_all_global_sex_effects
save_plot_object(
  Plot_SFig5_all_global_sex_effects,
  "Fig5_supp_all_global_sex_effects.rds"
)
```

# Full-resolution tissue-by-cell-type sex effects

```{r fig5-full-tissue-sex-heatmap, fig.width=13, fig.height=10}
tissue_effect_data <- tissue_effects |>
  transmute(
    tissue_label = .data[["Tissue group"]],
    cell_label = .data[["Immune cell type"]],
    c_effect = .data[[
      "Male-minus-female posterior mean effect (unconstrained sccomp composition scale)"
    ]],
    c_FDR = .data[["Bayesian FDR"]],
    supported = .data[["FDR < 0.05"]] == "Yes"
  ) |>
  filter(!is.na(tissue_label), !is.na(cell_label), is.finite(c_effect))

tissue_order <- tissue_effect_data |>
  group_by(tissue_label) |>
  summarise(mean_effect = mean(c_effect, na.rm = TRUE), .groups = "drop") |>
  arrange(mean_effect) |>
  pull(tissue_label)
cell_order <- tissue_effect_data |>
  group_by(cell_label) |>
  summarise(mean_effect = mean(c_effect, na.rm = TRUE), .groups = "drop") |>
  arrange(mean_effect) |>
  pull(cell_label)

tissue_effect_data <- tissue_effect_data |>
  mutate(
    tissue_label = factor(tissue_label, levels = tissue_order),
    cell_label = factor(cell_label, levels = cell_order)
  )

tissue_effect_limit <- max(abs(tissue_effect_data$c_effect), na.rm = TRUE)

Plot_SFig5_full_tissue_sex_heatmap <-
  ggplot(tissue_effect_data, aes(x = cell_label, y = tissue_label, fill = c_effect)) +
  geom_tile(colour = "white", linewidth = 0.15) +
  geom_point(
    data = filter(tissue_effect_data, supported),
    size = 0.8, colour = "black"
  ) +
  scale_fill_gradient2(
    low = "#D7907B", mid = "white", high = "#0A014F", midpoint = 0,
    limits = c(-tissue_effect_limit, tissue_effect_limit),
    name = "Male − female\neffect"
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "Full tissue-resolved landscape of sex effects",
    subtitle = "All estimates are shown; black points denote FDR < 0.05"
  ) +
  theme_supplementary +
  theme(
    panel.grid.major = element_blank(),
    axis.text.x = element_text(angle = 55, hjust = 1, size = 6)
  )

Plot_SFig5_full_tissue_sex_heatmap
save_plot_object(
  Plot_SFig5_full_tissue_sex_heatmap,
  "Fig5_supp_full_tissue_sex_heatmap.rds"
)
```

# Sex×age-split boxplots by age-trend × sex-gap quadrant

Main Fig. 5b shows the four lowest-FDR effects overall (no same-signed filter).
Here every cell type is plotted in its age-trend × sex-gap quadrant, using
adjusted sample points and four sex × earlier/later groups (split at age
decade 5).

```{r fig5-quadrant-boxplots-data}
earlier_age_decades <- as.character(1:4)
later_age_decades <- as.character(5:10)
age_sex_group_levels <- c(
  "female earlier", "female later", "male earlier", "male later"
)

assign_age_sex_group <- function(df) {
  df %>%
    mutate(
      age_decade_chr = as.character(age_decade),
      age_split = case_when(
        age_decade_chr %in% earlier_age_decades ~ "earlier",
        age_decade_chr %in% later_age_decades ~ "later",
        TRUE ~ NA_character_
      ),
      age_sex_group = paste(sex, age_split),
      age_sex_group = factor(age_sex_group, levels = age_sex_group_levels)
    ) %>%
    filter(!is.na(age_split), sex %in% c("female", "male"))
}

standardise_cell_col <- function(df) {
  if ("cell_type_unified_ensemble" %in% names(df) && !"L3" %in% names(df)) {
    df <- df %>% dplyr::rename(L3 = cell_type_unified_ensemble)
  } else if ("L3" %in% names(df) && !"cell_type_unified_ensemble" %in% names(df)) {
    df <- df %>% mutate(cell_type_unified_ensemble = L3)
  }
  df
}

format_y_prop_label <- function(x) {
  gsub("^0\\.", ".", formatC(x, format = "fg", digits = 3, drop0trailing = TRUE))
}

pred_draws <- readRDS(
  resolve_fig5_cache("sccomp_sex_age_split_pred_draws_fig4.rds")
) %>%
  standardise_cell_col() %>%
  assign_age_sex_group()

sex_adjusted <- readRDS(
  resolve_fig5_cache("sccomp_sex_age_split_adjusted_fig4_processed.rds")
) %>%
  standardise_cell_col()
if (!"age_sex_group" %in% names(sex_adjusted)) {
  sex_adjusted <- assign_age_sex_group(sex_adjusted)
}

build_age_sex_boxplot_df <- function(cell_types, test_df) {
  bind_rows(
    pred_draws %>%
      filter(L3 %in% cell_types) %>%
      dplyr::select(sample_id, L3, sex, age_decade, age_split, age_sex_group, proportion_mean) %>%
      mutate(version = "prediction"),
    sex_adjusted %>%
      filter(L3 %in% cell_types) %>%
      dplyr::select(sample_id, L3, sex, age_decade, age_split, age_sex_group, proportion_mean) %>%
      mutate(version = "adjusted")
  ) %>%
    mutate(L3_abbrev = recode(L3, !!!cell_type_abbrev_dict)) %>%
    left_join(
      test_df %>% dplyr::select(cell_type_unified_ensemble, c_FDR, c_effect),
      by = c("L3" = "cell_type_unified_ensemble")
    ) %>%
    mutate(fdr_formatted = format_fdr(c_FDR)) %>%
    {
      df_temp <- .
      unique_levels <- df_temp %>%
        distinct(L3, L3_abbrev) %>%
        arrange(match(L3, cell_types)) %>%
        pull(L3_abbrev)
      df_temp %>%
        mutate(
          L3 = factor(L3, levels = cell_types),
          L3_abbrev = factor(L3_abbrev, levels = unique_levels),
          age_sex_group = factor(age_sex_group, levels = age_sex_group_levels)
        )
    }
}

plot_age_sex_boxplot <- function(
    df,
    box_version = c("adjusted", "prediction"),
    y_pad = 0.08,
    ncol = 4,
    nrow = NULL
) {
  box_version <- match.arg(box_version)
  box_df <- df %>% filter(version == box_version)
  jitter_df <- df %>% filter(version == "adjusted")
  if (base::nrow(box_df) == 0) stop("No rows for box_version = '", box_version, "'.")

  facet_levels <- if (is.factor(box_df$L3_abbrev)) {
    levels(box_df$L3_abbrev)[levels(box_df$L3_abbrev) %in% unique(as.character(box_df$L3_abbrev))]
  } else {
    unique(as.character(box_df$L3_abbrev))
  }

  box_lims <- box_df %>%
    group_by(L3_abbrev, age_sex_group) %>%
    summarise(
      stats = list(boxplot.stats(proportion_mean, do.conf = FALSE, do.out = FALSE)$stats),
      .groups = "drop"
    ) %>%
    mutate(
      ymin = purrr::map_dbl(stats, 1),
      ymax = purrr::map_dbl(stats, 5)
    ) %>%
    group_by(L3_abbrev) %>%
    summarise(ymin = min(ymin), ymax = max(ymax), .groups = "drop") %>%
    mutate(
      ymin_t = asin(sqrt(pmax(ymin, 0))),
      ymax_t = asin(sqrt(pmax(ymax, 0))),
      pad = y_pad * pmax(ymax_t - ymin_t, 0.02),
      ymin = sin(pmax(ymin_t - pad, 0))^2,
      ymax = sin(pmin(ymax_t + pad, pi / 2))^2,
      fdr_y = ymin + 0.92 * (ymax - ymin)
    ) %>%
    mutate(L3_abbrev = factor(as.character(L3_abbrev), levels = facet_levels)) %>%
    arrange(L3_abbrev)

  fdr_labels <- box_df %>%
    distinct(L3_abbrev, fdr_formatted) %>%
    left_join(box_lims %>% dplyr::select(L3_abbrev, fdr_y), by = "L3_abbrev") %>%
    mutate(
      age_sex_group = factor(
        age_sex_group_levels[ceiling(length(age_sex_group_levels) / 2)],
        levels = age_sex_group_levels
      ),
      proportion_mean = fdr_y,
      label = paste0("FDR: ", fdr_formatted)
    )

  y_scales <- purrr::map(seq_len(base::nrow(box_lims)), function(i) {
    facet_val <- as.character(box_lims$L3_abbrev[i])
    rlang::new_formula(
      lhs = rlang::expr(L3_abbrev == !!facet_val),
      rhs = scale_y_continuous(
        trans = "asn",
        limits = c(box_lims$ymin[i], box_lims$ymax[i]),
        labels = format_y_prop_label,
        oob = scales::oob_keep,
        expand = expansion(mult = 0)
      )
    )
  })

  ggplot(box_df, aes(x = age_sex_group, y = proportion_mean)) +
    geom_jitter(
      data = jitter_df, width = 0.15, height = 0,
      colour = "grey50", shape = 16, size = 0.15, alpha = 0.2
    ) +
    geom_boxplot(aes(fill = sex), outlier.shape = NA, alpha = 0.3, linewidth = 0.25) +
    geom_text(
      data = fdr_labels, aes(label = label),
      size = 2, colour = "grey20", vjust = 0.5
    ) +
    scale_fill_manual(values = colmap_sex) +
    scale_x_discrete(labels = c(
      "female earlier" = "F(e)", "female later" = "F(l)",
      "male earlier" = "M(e)", "male later" = "M(l)"
    )) +
    facet_wrap(~ L3_abbrev, ncol = ncol, nrow = nrow, scales = "free_y") +
    ggh4x::facetted_pos_scales(y = y_scales) +
    labs(title = "", x = "Sex × Age groups (earlier vs later)", y = "Proportion Mean") +
    theme_multipanel +
    theme(
      strip.text.x.top = element_text(margin = margin(0, 0, 0, 0, "pt")),
      strip.placement = "inside"
    )
}

# Titles follow the S23 combined interpretation of S4 aging sign × sex-gap change.
supp_quadrants <- tibble::tribble(
  ~sign_age_trend, ~c_effect_dir, ~panel_id, ~panel_title,
  1,  "positive", "A", "Increases with age; males increase more than females",
  1,  "negative", "B", "Increases with age; females increase more than males",
  -1, "positive", "C", "Decreases with age; females decrease more than males",
  -1, "negative", "D", "Decreases with age; males decrease more than females"
)

build_quadrant_boxplot <- function(sign_age, effect_dir, panel_title) {
  cell_types <- test_sex_diff_split_at_5 %>%
    filter(
      sign_age_trend == sign_age,
      if (effect_dir == "positive") c_effect > 0 else c_effect < 0
    ) %>%
    arrange(desc(abs(c_effect)), c_FDR) %>%
    pull(cell_type_unified_ensemble)

  if (length(cell_types) == 0) {
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5, label = "No cell types", size = 3) +
        labs(title = panel_title) +
        theme_void() +
        theme(plot.title = element_text(size = 8, face = "bold"))
    )
  }

  df_q <- build_age_sex_boxplot_df(cell_types, test_sex_diff_split_at_5)
  plot_age_sex_boxplot(df_q, box_version = "adjusted", ncol = 4) +
    labs(title = panel_title) +
    theme(plot.title = element_text(size = 8, face = "bold", hjust = 0))
}

n_per_panel <- purrr::map_int(seq_len(nrow(supp_quadrants)), function(i) {
  test_sex_diff_split_at_5 %>%
    filter(
      sign_age_trend == supp_quadrants$sign_age_trend[i],
      if (supp_quadrants$c_effect_dir[i] == "positive") c_effect > 0 else c_effect < 0
    ) %>%
    nrow()
})
rows_per_panel <- pmax(1, ceiling(n_per_panel / 4))
panel_heights <- rows_per_panel / sum(rows_per_panel)
```

```{r fig5-quadrant-boxplots, fig.width=7.2, fig.height=12}
Plot_SFig5_boxplot_quadrants <- purrr::pmap(
  list(
    supp_quadrants$sign_age_trend,
    supp_quadrants$c_effect_dir,
    supp_quadrants$panel_title,
    seq_along(supp_quadrants$panel_id)
  ),
  function(sign_age, effect_dir, panel_title, i) {
    p <- build_quadrant_boxplot(sign_age, effect_dir, panel_title)
    if (i < length(supp_quadrants$panel_id)) {
      p +
        labs(x = NULL) +
        theme(legend.position = "none", axis.title.x = element_blank())
    } else {
      p + theme(legend.position = "bottom")
    }
  }
)

Plot_SFig5_all_boxplot <-
  wrap_plots(Plot_SFig5_boxplot_quadrants, ncol = 1, heights = panel_heights) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 9, face = "bold"))

Plot_SFig5_all_boxplot
save_plot_object(Plot_SFig5_all_boxplot, "Fig5_supp_boxplot_age_trend_x_sex_gap.rds")

fig_width_mm <- 183
fig_height_mm <- min(220, 48 + sum(rows_per_panel) * 42)
ggsave(
  file.path(plot_object_dir, "Fig5_supp_boxplot_age_trend_x_sex_gap.png"),
  Plot_SFig5_all_boxplot,
  width = fig_width_mm,
  height = fig_height_mm,
  units = "mm",
  dpi = 300,
  bg = "white"
)

# Also keep per-quadrant objects for reuse.
purrr::walk2(
  Plot_SFig5_boxplot_quadrants,
  supp_quadrants$panel_id,
  function(p, id) {
    save_plot_object(
      p + theme(legend.position = "bottom"),
      sprintf("Fig5_supp_boxplot_quadrant_%s.rds", id)
    )
  }
)
```

# Plasma sex trajectories for every tissue

No coverage filter: every tissue with plasma-cell predictions is shown (female/male
only; unknown sex dropped). Main Fig. 5d still uses a stricter subset. Styling
matches Fig2 supplementary plasma trajectories (raw proportions, asin-sqrt y scale).

```{r fig5-all-plasma-line, fig.width=15, fig.height=14}
sex_tissue_pred <- readRDS(
  resolve_fig5_cache(
    "predictions_age_decade_sex_tissue___L3___disease_FALSE___immune_only_TRUE.rds"
  )
)

plasma_line_data <- sex_tissue_pred |>
  filter(
    cell_type_unified_ensemble == "plasma",
    sex %in% c("female", "male"),
    !is.na(age_decade),
    !is.na(tissue_groups)
  ) |>
  group_by(tissue_groups, cell_type_unified_ensemble, age_decade, sex) |>
  summarise(
    n_samples = n_distinct(sample_id),
    proportion_mean = first(proportion_mean),
    proportion_lower = first(proportion_lower),
    proportion_upper = first(proportion_upper),
    .groups = "drop"
  ) |>
  mutate(
    Age = factor(as.character(age_decade), levels = as.character(1:10)),
    Sex = factor(sex, levels = c("female", "male")),
    tissue_groups_short = {
      out <- unname(tissue_name_dict[tissue_groups])
      out <- ifelse(
        is.na(out),
        tools::toTitleCase(gsub("_", " ", tissue_groups, fixed = TRUE)),
        out
      )
      factor(out, levels = c(setdiff(sort(unique(out)), "Large int"), "Large int"))
    }
  )

Plot_SFig5_all_plasma_line <-
  plasma_line_data |>
  ggplot() +
  geom_hline(yintercept = 0, lwd = 0.1) +
  geom_point(aes(x = Age, y = proportion_mean), shape = NA) +
  geom_line(
    aes(x = as.numeric(Age), y = proportion_mean, colour = Sex, fill = Sex),
    show.legend = FALSE
  ) +
  geom_ribbon(
    aes(
      x = as.numeric(Age),
      colour = Sex,
      fill = Sex,
      ymin = proportion_lower,
      ymax = proportion_upper
    ),
    colour = NA,
    alpha = 0.1
  ) +
  scale_colour_manual(
    values = c("#D7907B", "#0A014F"),
    aesthetics = c("colour", "fill"),
    guide = guide_legend(override.aes = list(alpha = 1))
  ) +
  ggnewscale::new_scale_colour() +
  scale_colour_brewer(palette = "Set2", name = "Comparison") +
  scale_y_continuous(
    trans = asin_sqrt_trans,
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.05)),
    labels = function(x) gsub("^0\\.", ".", format(x, nsmall = 1))
  ) +
  labs(x = "Age decade", y = "Proportion Mean (arcsin-sqrt scale)") +
  facet_wrap(~ tissue_groups_short, ncol = 5, scales = "free_y") +
  theme_multipanel +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 0, hjust = 0.5)
  )

Plot_SFig5_all_plasma_line
save_plot_object(
  Plot_SFig5_all_plasma_line,
  "Fig5_supp_all_plasma_line.rds"
)
```

# Output objects

```{r output-manifest}
data.frame(
  object = c(
    "Plot_SFig5_full_cohort_coverage",
    "Plot_SFig5_all_global_sex_effects",
    "Plot_SFig5_full_tissue_sex_heatmap",
    "Plot_SFig5_all_boxplot",
    "Plot_SFig5_boxplot_quadrants",
    "Plot_SFig5_all_plasma_line"
  ),
  rds_file = c(
    "Fig5_supp_full_cohort_coverage.rds",
    "Fig5_supp_all_global_sex_effects.rds",
    "Fig5_supp_full_tissue_sex_heatmap.rds",
    "Fig5_supp_boxplot_age_trend_x_sex_gap.rds",
    "Fig5_supp_boxplot_quadrant_{A–D}.rds",
    "Fig5_supp_all_plasma_line.rds"
  )
)
```
