---
title: "Figure 5 supplementary plots"
date: today
date-format: "DD MMM YYYY"
format:
  html:
    theme: minty
    toc: true
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
---

# Purpose

This report uses the finalized Figure 5 source-data workbook to show:

1. age-by-sex donor coverage across all 36 mapped tissue groups;
2. every body-level sex contrast across age decades;
3. a full-resolution tissue-by-cell-type sex-effect heatmap; and
4. a plasma-cell tissue audit showing which tested tissues have trajectories in
   the main panel.

The main plasma trajectories are not duplicated. The exported workbook contains
only the trajectories that passed the main panel's display rules, so omitted
trajectories cannot be reconstructed faithfully from that sheet alone.

```{r setup}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(here)
library(knitr)

source(file.path("..", "R", "paths.R"))
workbook <- path_source_tables("SupplData_AgeClock_Fig5.xlsx")
if (!file.exists(workbook)) stop("Figure 5 source-data workbook not found: ", workbook)

plot_object_dir <- file.path(here::here(), "vignettes", "supplementary_plot_objects")
dir.create(plot_object_dir, recursive = TRUE, showWarnings = FALSE)

read_panel_sheet <- function(sheet) {
  if (!sheet %in% excel_sheets(workbook)) stop("Missing worksheet: ", sheet)
  read_xlsx(workbook, sheet = sheet, guess_max = 1000000L)
}

save_plot_object <- function(plot, filename) {
  path <- file.path(plot_object_dir, filename)
  saveRDS(plot, path, compress = "xz")
  message("Saved ggplot object: ", path)
  invisible(path)
}

clean_label <- function(x) {
  x <- gsub("_", " ", x, fixed = TRUE)
  tools::toTitleCase(x)
}

theme_supplementary <- theme_bw(base_size = 8.5) +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_blank(),
    legend.position = "bottom",
    plot.title.position = "plot"
  )
```

```{r read-source-data}
cohort_counts <- read_panel_sheet("Fig5_A_all_tissue_donor_counts")
global_sex_tests <- read_panel_sheet("Fig5_B_all_global_sex_tests_by_")
tissue_sex_tests <- read_panel_sheet("Fig5_C_tissue_sex_effect_tests")
plasma_trajectories <- read_panel_sheet("Fig5_D_plotted_estimates")
```

# Complete age-by-sex cohort coverage

The three records with missing tissue-group mapping are excluded from the
36-tissue display and reported separately below.

```{r fig5-full-cohort-coverage, fig.width=8, fig.height=12}
unmapped_donor_records <- cohort_counts |>
  filter(is.na(tissue_groups)) |>
  summarise(donor_count = sum(donor_count, na.rm = TRUE)) |>
  pull(donor_count)

tissue_map <- cohort_counts |>
  filter(!is.na(tissue_groups), !is.na(tissue_label)) |>
  distinct(tissue_groups, tissue_label)

coverage_data <- cohort_counts |>
  filter(!is.na(tissue_groups)) |>
  select(tissue_groups, age_groups, sex, donor_count) |>
  complete(
    tissue_groups,
    age_groups = c("Young", "Mid", "Old"),
    sex = c("female", "male"),
    fill = list(donor_count = 0)
  ) |>
  left_join(tissue_map, by = "tissue_groups")

tissue_order <- coverage_data |>
  group_by(tissue_label) |>
  summarise(total = sum(donor_count), .groups = "drop") |>
  arrange(total) |>
  pull(tissue_label)

coverage_data <- coverage_data |>
  mutate(
    tissue_label = factor(tissue_label, levels = tissue_order),
    age_groups = factor(age_groups, levels = c("Young", "Mid", "Old")),
    sex = factor(sex, levels = c("female", "male"), labels = c("Female", "Male"))
  )

Plot_SFig5_full_cohort_coverage <-
  ggplot(coverage_data, aes(x = age_groups, y = tissue_label, fill = donor_count)) +
  geom_tile(colour = "white", linewidth = 0.25) +
  geom_text(aes(label = ifelse(donor_count == 0, "", donor_count)), size = 2.2) +
  facet_grid(. ~ sex) +
  scale_fill_viridis_c(trans = "sqrt", option = "C", name = "Unique donors") +
  labs(
    x = "Age group",
    y = NULL,
    title = "Donor coverage across all mapped tissues, age groups and sexes",
    subtitle = sprintf(
      "Thirty-six mapped tissue groups are shown; %d donor records have missing tissue-group mapping",
      unmapped_donor_records
    )
  ) +
  theme_supplementary +
  theme(panel.grid.major = element_blank())

Plot_SFig5_full_cohort_coverage
save_plot_object(
  Plot_SFig5_full_cohort_coverage,
  "Fig5_supp_full_cohort_coverage.rds"
)
```

# Complete body-level sex effects across age decades

Positive effects indicate male enrichment relative to the female baseline.
Stars denote FDR-supported contrasts.

```{r fig5-all-global-sex-effects, fig.width=9, fig.height=8}
global_effect_data <- global_sex_tests |>
  filter(!is.na(cell_type_unified_ensemble), !is.na(Age)) |>
  mutate(
    cell_label = clean_label(cell_type_unified_ensemble),
    age_decade = factor(as.character(Age), levels = as.character(1:10)),
    supported = !is.na(c_FDR) & c_FDR < 0.05
  )

cell_order <- global_effect_data |>
  group_by(cell_label) |>
  summarise(mean_effect = mean(c_effect, na.rm = TRUE), .groups = "drop") |>
  arrange(mean_effect) |>
  pull(cell_label)

global_effect_data <- global_effect_data |>
  mutate(cell_label = factor(cell_label, levels = cell_order))

global_effect_limit <- max(abs(global_effect_data$c_effect), na.rm = TRUE)

Plot_SFig5_all_global_sex_effects <-
  ggplot(global_effect_data, aes(x = age_decade, y = cell_label, fill = c_effect)) +
  geom_tile(colour = "white", linewidth = 0.2) +
  geom_point(
    data = filter(global_effect_data, supported),
    shape = 8,
    size = 1.4,
    colour = "black"
  ) +
  scale_fill_gradient2(
    low = "#D7907B", mid = "white", high = "#0A014F", midpoint = 0,
    limits = c(-global_effect_limit, global_effect_limit),
    name = "Male − female\neffect"
  ) +
  labs(
    x = "Age decade",
    y = NULL,
    title = "Complete body-level sex effects across age decades",
    subtitle = "The fifth decade is the model reference; stars denote FDR < 0.05"
  ) +
  theme_supplementary +
  theme(panel.grid.major = element_blank())

Plot_SFig5_all_global_sex_effects
save_plot_object(
  Plot_SFig5_all_global_sex_effects,
  "Fig5_supp_all_global_sex_effects.rds"
)
```

# Full-resolution tissue-by-cell-type sex effects

This is intended as the legible supplementary counterpart to the compressed
main Figure 5C. Non-significant estimates remain visible; black points denote
FDR-supported effects.

```{r fig5-full-tissue-sex-heatmap, fig.width=13, fig.height=10}
tissue_effect_data <- tissue_sex_tests |>
  filter(!is.na(Tissue), !is.na(cell_type_unified_ensemble)) |>
  mutate(
    tissue_label = ifelse(is.na(tissue_short), clean_label(Tissue), tissue_short),
    cell_label = ifelse(
      is.na(cell_type_abbrev),
      clean_label(cell_type_unified_ensemble),
      cell_type_abbrev
    ),
    supported = !is.na(c_FDR) & c_FDR < 0.05
  )

tissue_order <- tissue_effect_data |>
  group_by(tissue_label) |>
  summarise(mean_effect = mean(c_effect, na.rm = TRUE), .groups = "drop") |>
  arrange(mean_effect) |>
  pull(tissue_label)
cell_order <- tissue_effect_data |>
  group_by(cell_label) |>
  summarise(mean_effect = mean(c_effect, na.rm = TRUE), .groups = "drop") |>
  arrange(mean_effect) |>
  pull(cell_label)

tissue_effect_data <- tissue_effect_data |>
  mutate(
    tissue_label = factor(tissue_label, levels = tissue_order),
    cell_label = factor(cell_label, levels = cell_order)
  )

tissue_effect_limit <- max(abs(tissue_effect_data$c_effect), na.rm = TRUE)

Plot_SFig5_full_tissue_sex_heatmap <-
  ggplot(tissue_effect_data, aes(x = cell_label, y = tissue_label, fill = c_effect)) +
  geom_tile(colour = "white", linewidth = 0.15) +
  geom_point(
    data = filter(tissue_effect_data, supported),
    size = 0.8,
    colour = "black"
  ) +
  scale_fill_gradient2(
    low = "#D7907B", mid = "white", high = "#0A014F", midpoint = 0,
    limits = c(-tissue_effect_limit, tissue_effect_limit),
    name = "Male − female\neffect"
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "Full tissue-resolved landscape of sex effects",
    subtitle = "All estimates are shown; black points denote FDR < 0.05"
  ) +
  theme_supplementary +
  theme(
    panel.grid.major = element_blank(),
    axis.text.x = element_text(angle = 55, hjust = 1)
  )

Plot_SFig5_full_tissue_sex_heatmap
save_plot_object(
  Plot_SFig5_full_tissue_sex_heatmap,
  "Fig5_supp_full_tissue_sex_heatmap.rds"
)
```

# Plasma-cell trajectory inclusion audit

This audit avoids drawing unsupported trajectories. It shows all tissues with a
plasma-cell sex-effect estimate and identifies which tissues are represented in
the main trajectory panel.

```{r fig5-plasma-inclusion-audit, fig.width=9, fig.height=7}
trajectory_tissues <- unique(plasma_trajectories$Tissue)

plasma_audit_data <- tissue_effect_data |>
  filter(cell_type_unified_ensemble == "plasma") |>
  mutate(
    trajectory_displayed = Tissue %in% trajectory_tissues,
    tissue_label = as.character(tissue_label)
  ) |>
  arrange(c_effect) |>
  mutate(tissue_label = factor(tissue_label, levels = unique(tissue_label)))

Plot_SFig5_plasma_tissue_audit <-
  ggplot(plasma_audit_data, aes(x = c_effect, y = tissue_label)) +
  geom_vline(xintercept = 0, colour = "grey55", linewidth = 0.35) +
  geom_segment(aes(x = 0, xend = c_effect, yend = tissue_label), colour = "grey75") +
  geom_point(
    aes(
      fill = trajectory_displayed,
      shape = supported,
      size = NSamplesTissue
    ),
    colour = "black"
  ) +
  scale_fill_manual(
    values = c(`TRUE` = "#4C9F70", `FALSE` = "white"),
    name = "Trajectory displayed"
  ) +
  scale_shape_manual(values = c(`TRUE` = 21, `FALSE` = 24), name = "FDR < 0.05") +
  scale_size_continuous(name = "Tissue samples", range = c(1.5, 5)) +
  labs(
    x = "Male − female plasma-cell effect",
    y = NULL,
    title = "Plasma-cell tissue effects and trajectory availability"
  ) +
  theme_supplementary

Plot_SFig5_plasma_tissue_audit
save_plot_object(
  Plot_SFig5_plasma_tissue_audit,
  "Fig5_supp_plasma_tissue_audit.rds"
)
```

```{r excluded-plasma-tissues}
plasma_audit_data |>
  filter(!trajectory_displayed) |>
  transmute(
    tissue = as.character(tissue_label),
    effect = c_effect,
    FDR = c_FDR,
    tissue_samples = NSamplesTissue,
    reason = "No trajectory exported under the main-panel display rules"
  ) |>
  kable(digits = 3, caption = "Tissues tested for plasma-cell sex effects without a displayed trajectory")
```

# Output objects

```{r output-manifest}
data.frame(
  object = c(
    "Plot_SFig5_full_cohort_coverage",
    "Plot_SFig5_all_global_sex_effects",
    "Plot_SFig5_full_tissue_sex_heatmap",
    "Plot_SFig5_plasma_tissue_audit"
  ),
  rds_file = c(
    "Fig5_supp_full_cohort_coverage.rds",
    "Fig5_supp_all_global_sex_effects.rds",
    "Fig5_supp_full_tissue_sex_heatmap.rds",
    "Fig5_supp_plasma_tissue_audit.rds"
  )
)
```
