---
title: "Fig 5"
date: today
date-format: "DD MMM YYYY"
title-block-banner: true
author: 
  - name: Chen Zhan
    orcid: 0000-0002-4794-8339
    email: chen.zhan@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
format: 
  html:
    theme: minty
    df-print: paged
    code-line-numbers: true
    embed-resources: true
    code-fold: true
    code-fold-default: true
    toc: true
    toc-location: left
    toc-title: "Menu"
knitr:
  opts_chunk: 
    message: false
    warning: false
comments:
  hypothesis: 
    theme: clean
editor: source
abstract: |
  Sex-differences in immune composition ageing across tissues
---

# Overview

This report reproduces **Figure 5** (sex differences in immune-composition
ageing across tissues). The final figure is a patchwork of:

- **A** – Donor counts by age group and sex across tissues (signed log bar chart).
- **B** – Sex differences in composition for the three most male- and three most
  female-biased cell types (arcsin-sqrt proportions).
- **C** – Heatmap of male-vs-female age effects (cell type × tissue).
- **D** – Plasma-cell proportion trajectories by sex across tissues.

Every panel **reads its saved object from `vignettes/data/figure_ready/` by default** and
only recomputes (loading the sccomp estimates, running `sccomp_test` /
`sccomp_predict`) when the cache is absent. This keeps the report fast to render
while remaining fully reproducible.

The cache filenames retain their historical `Fig4_` prefix for compatibility;
all manuscript-facing figures and supplementary exports use Figure 5.

# Library


```{r publication-paths}
# Portable publication data roots (no HPC absolute paths).
.path_candidates <- c(
  file.path("..", "R", "paths.R"),
  if (requireNamespace("here", quietly = TRUE)) {
    file.path(here::here(), "vignettes", "R", "paths.R")
  },
  file.path("R", "paths.R")
)
.path_candidates <- .path_candidates[!is.na(.path_candidates) & nzchar(.path_candidates)]
.path_file <- .path_candidates[file.exists(.path_candidates)][1]
if (is.na(.path_file)) {
  stop(
    "Cannot find vignettes/R/paths.R. Render from the immuneHealthyBodyMap ",
    "repository root or from vignettes/reports/, or set IMMUNE_HEALTHY_BODY_MAP_ROOT."
  )
}
source(.path_file)
fig_cache_dir <- path_figure_ready()
dir.create(path_intermediate("fig2"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_intermediate("fig4"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_intermediate("fig5"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_figure_ready("fig1"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_figure_ready("fig2"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_figure_ready("fig4"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_figure_ready("fig5"), recursive = TRUE, showWarnings = FALSE)
```

```{r setup-libraries}
# Set your R library path here if needed
# .libPaths removed for portability; use renv/conda or site library
library(here)  # project-root aware paths

# Ensure required output directories exist
required_dirs <- c(
  path_figure_ready("fig5"),
  path_intermediate("fig5")
)
invisible(lapply(required_dirs, function(d) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE, showWarnings = FALSE)
}))

library(magrittr)
library(tidyverse)
library(glue)
library(patchwork)          # figure assembly
library(ggplotify)          # as.ggplot() for the heatmap panel
library(ComplexHeatmap)
library(tidyHeatmap)        # heatmap(), layer_point(), annotation_bar()
library(ggnewscale)         # second colour scale in the line chart
library(scico)              # 'vik' diverging palette for the heatmap
library(circlize)           # colorRamp2()
library(ggplot2)
library(grid)               # unit()
# NOTE: `sccomp` and `cellNexus` are loaded on-demand inside the computation
# chunks (panels A–D) so the report still renders from cache without them.
```

# Setup: theme, dictionaries, palettes and paths

```{r define-theme}
theme_multipanel <-
  theme_bw() +
  theme(
    panel.border = element_blank(),
    axis.line = element_line(size = 0.1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    strip.background = element_blank(),
    axis.title.y = element_text(margin = margin(0, 0, 0, 0), size = 7),
    axis.title.x = element_text(margin = margin(0, 0, 0, 0), size = 7),
    panel.spacing.x = unit(0.1, "lines"),
    axis.text.x = element_text(size = 6),
    axis.text.y = element_text(size = 6),
    strip.text.x = element_text(size = 7),
    strip.text.y = element_text(size = 7),
    legend.key.size = unit(5, 'mm'),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 6),
    strip.clip = "off",
    plot.title = element_text(size = 7),
    axis.line.x = element_line(size = 0.2),
    axis.line.y = element_line(size = 0.2),
    axis.ticks.x = element_line(size = 0.2),
    axis.ticks.y = element_line(size = 0.2)
  )
```

```{r define-dictionaries}
# Shortened tissue names (shared across all figures)
tissue_name_dict <- c(
  "blood" = "Blood",
  "sensory-related structures" = "Sensory",
  "cerebral lobes and cortical areas" = "Cerebral cortex",
  "limbic and basal systems" = "Limbic/basal",
  "liver" = "Liver",
  "bone marrow" = "Bone marrow",
  "large intestine" = "Large int",
  "lymphatic system" = "Lymphatic",
  "small intestine" = "Small int",
  "respiratory system" = "Respiratory",
  "renal system" = "Renal",
  "brainstem and cerebellar structures" = "Hindbrain",
  "cardiovascular system" = "Cardiovascular",
  "general brain and major structures" = "Brain",
  "vasculature" = "Vasculature",
  "ovary" = "Ovary",
  "integumentary system (skin)" = "Skin",
  "nasal, oral, and pharyngeal regions" = "Pharyngeal",
  "muscular system (skeletal muscles)" = "Skeletal muscle",
  "prostate" = "Prostate",
  "breast" = "Breast",
  "digestive system (general)" = "Digestive",
  "oesophagus" = "Oesophagus",
  "female reproductive system" = "Female repro",
  "spleen" = "Spleen",
  "peritoneal and abdominal cavity structures" = "Peritoneal",
  "thymus" = "Thymus",
  "adipose tissue" = "Adipose",
  "trachea" = "Trachea",
  "endocrine system" = "Endocrine",
  "pancreas" = "Pancreas",
  "miscellaneous glands" = "Glands",
  "stomach" = "Stomach",
  "gallbladder" = "Gallbladder"
)

# Cell-type abbreviations (single-line, used for boxplot facets)
cell_type_abbrev_dict <- c(
  "b memory" = "B Memory", "b naive" = "B Naive",
  "cd14 mono" = "CD14 Mono", "cd16 mono" = "CD16 Mono",
  "cd4 fh em" = "CD4 FH EM", "cd4 naive" = "CD4 Naive", "cd4 tcm" = "CD4 TCM",
  "cd4 th1 em" = "CD4 Th1 EM", "cd4 th1/th17 em" = "CD4 Th1/Th17 EM",
  "cd4 th17 em" = "CD4 Th17 EM", "cd4 th2 em" = "CD4 Th2 EM", "treg" = "Treg",
  "cd8 naive" = "CD8 Naive", "cd8 tcm" = "CD8 TCM", "cd8 tem" = "CD8 TEM",
  "cdc" = "CDC", "granulocyte" = "Gran", "ilc" = "ILC", "macrophage" = "Mac",
  "mait" = "MAIT", "mast" = "Mast", "nk" = "NK", "nkt" = "NKT", "pdc" = "PDC",
  "tgd" = "TGD", "plasma" = "Plasma"
)

# Cell-type abbreviations for the heatmap columns
cell_type_abbrev_dict_heatmap <- c(
  "macrophage" = "Mac", "cd4 th1 em" = "CD4 Th1 EM", "cd4 th1/th17 em" = "CD4 Th1/17 EM",
  "cd4 th17 em" = "CD4 Th17 EM", "cd4 th2 em" = "CD4 Th2 EM", "cd8 naive" = "CD8 Naive",
  "plasma" = "Plasma", "granulocyte" = "Gran", "cd4 naive" = "CD4 Naive",
  "b naive" = "B Naive", "nk" = "NK", "nkt" = "NKT", "cdc" = "CDC", "pdc" = "PDC",
  "b memory" = "B Memory", "cd14 mono" = "CD14 Mono", "cd16 mono" = "CD16 Mono",
  "cd4 fh em" = "CD4 FH EM", "cd4 tcm" = "CD4 TCM", "cd8 tcm" = "CD8 TCM",
  "cd8 tem" = "CD8 TEM", "ilc" = "ILC", "mait" = "MAIT", "mast" = "Mast",
  "tgd" = "TGD", "treg" = "Treg"
)

colmap_sex <- c("female" = "#D7907B", "male" = "#0A014F")
```

```{r define-paths-and-loaders}
# --- sccomp estimates (large; loaded lazily only when a cache is missing) -----
# Update this path to your sccomp estimates object.
estimates_file <-
  path_sccomp_estimates("estimates_age_decade___L3___disease_FALSE___immune_only_TRUE.rds")
cell_type_col <- "L3"

.est_raw <- NULL
get_est <- function() {
  if (is.null(.est_raw)) .est_raw <<- readRDS(estimates_file)
  .est_raw
}
get_estimates_renamed <- function() {
  get_est() |> dplyr::rename(cell_type_unified_ensemble = !!sym(cell_type_col))
}

# --- Cached panel objects -----------------------------------------------------
fig_files <- path_figure_ready("fig5")
path_bar        <- file.path(fig_files, "Fig4_Plot__age_sex_ethnicity_tissue_bar_consensus.rds")
path_bar_all_counts <- file.path(fig_files, "Fig5_A_donor_counts_all_tissues.rds")
path_heatmap    <- file.path(fig_files, "Fig4_Plot__heatmap_male_vs_female.rds")
path_box_red    <- file.path(fig_files, "Fig4_Plot__boxplot_uniformly_red.rds")
path_box_blue   <- file.path(fig_files, "Fig4_Plot__boxplot_uniformly_blue.rds")
path_line       <- file.path(fig_files, "Fig4_Plot__line_chart_plasma.rds")
```

# Panel A – Donor counts by age group and sex

```{r fig.width=25, fig.height=5, label="panel-A-bar"}
rebuild_panel_a <- tolower(Sys.getenv("REBUILD_FIG5_PANEL_A", "false")) %in%
  c("1", "true", "yes")

if (file.exists(path_bar) && file.exists(path_bar_all_counts) && !rebuild_panel_a) {

  Plot__age_sex_ethnicity_tissue_bar <- readRDS(path_bar)
  donor_count_all <- readRDS(path_bar_all_counts)

} else {

  library(cellNexus)  # get_metadata_url() / get_metadata()

  edit_covariates_path <- path_edit_covariates()
  source(edit_covariates_path)

  cell_metadata <-
    get_publication_metadata() %>%
    filter(disease == "normal") |> filter(age_days > 365)

  donor_metadata <-
    cell_metadata |>
    distinct(donor_id, self_reported_ethnicity, assay, age_days, sex, tissue, disease) |>
    as_tibble() %>%
    edit_covariates() %>%
    filter(sex %in% c("female", "male"))

  donor_count_all <-
    donor_metadata |>
    mutate(age_groups = case_when(
      age_decade %in% c('1', '2', '3') ~ 'Young',
      age_decade %in% c('4', '5', '6') ~ 'Mid',
      age_decade %in% c('7', '8', '9') ~ 'Old',
      .default = 'Old'
    )) |>
    # Count each donor once per tissue/age/sex cohort. A donor can have several
    # samples or assay records within a tissue, which must not inflate panel A.
    distinct(donor_id, age_groups, sex, tissue_groups) |>
    dplyr::count(age_groups, sex, tissue_groups, name = "donor_count") |>
    as_tibble() |>
    mutate(
      tissue_label = recode(tissue_groups, !!!tissue_name_dict),
      tissue_label = stringr::str_replace_all(tissue_label, "/", "/\n"),
      tissue_label = stringr::str_wrap(tissue_label, width = 10)
    )

  saveRDS(donor_count_all, path_bar_all_counts)

  tissue_order <- donor_count_all |>
    group_by(tissue_label) |>
    summarise(total_n = sum(donor_count), .groups = "drop") |>
    arrange(desc(total_n)) |>
    pull(tissue_label)

  # Keep the 25 best-represented tissues for legibility
  top_tissues <- head(tissue_order, 25)

  df_bar <- donor_count_all |>
    filter(tissue_label %in% top_tissues) |>
    mutate(
      tissue_groups = factor(tissue_label, levels = tissue_order),
      signed_donor_count = ifelse(sex == "female", -donor_count, donor_count)
    ) |>
    mutate(age_groups = factor(age_groups, levels = c("Old", "Mid", "Young")),
           Sign = sign(signed_donor_count))

  Plot__age_sex_ethnicity_tissue_bar <-
    df_bar |>
    ggplot(aes(x = age_groups, y = log(abs(signed_donor_count)) * Sign, fill = sex)) +
    geom_bar(stat = "identity", position = "stack") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
    facet_grid(. ~ tissue_groups) +
    scale_fill_manual(values = colmap_sex, guide = guide_legend(nrow = 1)) +
    labs(x = "", y = expression(log(abs(donor[count])) %.% Sign), fill = "Sex") +
    theme_multipanel +
    theme(
      strip.placement = "inside",
      strip.text.x.top = element_text(angle = 45, hjust = 1),
      strip.background = element_blank(),
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
      plot.title = element_blank()
    )

  saveRDS(Plot__age_sex_ethnicity_tissue_bar, path_bar)
}

print(Plot__age_sex_ethnicity_tissue_bar)
```

# Sex-difference composition tests (shared inputs)

These `sccomp` contrasts feed panels **B** and **C**. The whole block is skipped
when all three panel caches (`path_heatmap`, `path_box_red`, `path_box_blue`) exist.

```{r sex-difference-sccomp}
path_global_sex_tests <- file.path(path_intermediate("fig5"), "Fig4_global_sex_tests_compact.rds")
need_sex <- !file.exists(path_heatmap) || !file.exists(path_box_red) ||
  !file.exists(path_box_blue) || !file.exists(path_global_sex_tests)

if (need_sex) {

  library(sccomp)  # sccomp_test(), sccomp_predict()

  estimates_age_bins <- get_estimates_renamed()
  load(file.path(path_intermediate("fig2"), "df_cell_statistics.RData"))  # df_cells_ct, df_cells_tissue

  age_bins      <- c("1", "2", "3", "4", "6", "7", "8", '9', '10')
  age_binary    <- setNames(c("Early", "Early", "Early", "Early",
                              "Late", "Late", "Late", "Late", "Late"), age_bins)

  # Parameter table for age / sex / age:sex (fixed + tissue-random)
  param_df <-
    estimates_age_bins |>
    filter(grepl("age_decade|sex", parameter)) |>
    mutate(
      params = str_split(parameter, "___", simplify = TRUE),
      params = (\(x) { colnames(x) <- c("parameter_tissue_removed", "Tissue"); as.data.frame(x) })(params)
    ) |>
    dplyr::select(params, parameter) |>
    tidyr::unpack(params) |>
    mutate(
      Age = str_extract(parameter_tissue_removed, "(?<=age_decade)(10|[1-9])"),
      Sex = str_extract(parameter_tissue_removed, "(?<=sex)(male|unknown)"),
      Tissue = if_else(Tissue == "", "global", Tissue),
      AgeBinary = age_binary[Age]
    ) |>
    distinct()

  # -- Tissue-specific male-vs-female contrasts (Panel C) ----------------------
  tissue_age_combo <- param_df %>%
    filter(Sex == 'male', Tissue != 'global', !is.na(Age)) %>% distinct(Tissue, Age)

  contrasts_df <-
    tissue_age_combo %>%
    mutate(contr = map2_chr(Tissue, Age, ~ {
      param_df %>%
        filter(Tissue %in% c(.x, 'global'), Age %in% c(.y, NA), Sex == 'male') %>%
        mutate(parameter = glue::glue('`{parameter}`')) %>% pull(parameter) %>%
        paste0(collapse = ' + ')
    })) %>%
    mutate(contrast_name = glue::glue('male-female___age_decade{Age} | {Tissue}'))

  contrasts_sex_diff_overall <-
    tissue_age_combo %>%
    mutate(contr = map2_chr(Tissue, Age, ~ {
      param_df %>%
        filter(Tissue %in% c(.x, 'global'), Age %in% c(.y, NA), Sex == 'male') %>%
        mutate(parameter = glue::glue('`{parameter}`')) %>% pull(parameter) %>%
        paste0(collapse = ' + ')
    })) %>%
    group_by(Tissue) %>%
    summarise(contr = { n <- n(); glue::glue("({paste(contr, collapse = ' + ')}) / {n}") },
              .groups = "drop") %>%
    mutate(contrast_name = Tissue)

  contrasts <- contrasts_sex_diff_overall %>%
    dplyr::select(contrast_name, contr) |> deframe()

  test_mixed_all <- estimates_age_bins |>
    sccomp_test(contrasts = contrasts) |>
    dplyr::rename(cell_type_unified_ensemble = !!sym(cell_type_col))

  # The contrast names are already the tissue names.  The previous join used
  # per-decade contrast names and therefore could not match these overall
  # contrasts when rebuilding from scratch.
  test_mixed_max <- test_mixed_all |>
    rename(Tissue = parameter) |>
    mutate(significant = c_FDR < 0.05)

  # -- Global male-vs-female contrasts (Panel B cell-type selection) -----------
  tissue_age_combo2 <- param_df %>%
    filter(Sex == 'male', Tissue == 'global', !is.na(Age)) %>% distinct(Tissue, Age)

  contrasts_df2 <-
    tissue_age_combo2 %>%
    mutate(contr = map2_chr(Tissue, Age, ~ {
      param_df %>%
        filter(Tissue %in% c(.x, 'global'), Age %in% c(.y, NA), Sex == 'male') %>%
        mutate(parameter = glue::glue('`{parameter}`')) %>% pull(parameter) %>%
        paste0(collapse = ' + ')
    })) %>%
    mutate(contrast_name = glue::glue('male-female___age_decade{Age} | {Tissue}'))

  contrasts2 <- contrasts_df2 %>%
    dplyr::select(contrast_name, contr) |> deframe()

  if (file.exists(path_global_sex_tests)) {
    test_mixed_all2 <- readRDS(path_global_sex_tests)
  } else {
    test_mixed_all2 <- estimates_age_bins |>
      sccomp_test(contrasts = contrasts2) |>
      dplyr::rename(cell_type_unified_ensemble = !!sym(cell_type_col)) |>
      left_join(contrasts_df2, by = join_by(parameter == contrast_name)) |>
      mutate(significant = c_FDR < 0.05)

    # Strip the very large sccomp attributes before caching the tabular tests.
    test_mixed_all2 <- tibble::as_tibble(setNames(
      lapply(names(test_mixed_all2), function(nm) test_mixed_all2[[nm]]),
      names(test_mixed_all2)
    ))
    saveRDS(test_mixed_all2, path_global_sex_tests)
  }

  test_mixed_max2 <- test_mixed_all2 |>
    group_by(cell_type_unified_ensemble) |>
    arrange(c_FDR, desc(abs(c_effect))) |>
    slice_head(n = 1) |>
    ungroup() %>%
    mutate(sig = c_FDR < 0.05)

  # -- Sex predictions for the boxplots (own caches) --------------------------
  path_pred_draws  <- file.path(path_intermediate("fig5"), "sccomp_sex_pred_draws_fig4.rds")
  path_sex_adjust  <- file.path(path_intermediate("fig5"), "sccomp_sex_adjusted_fig4.rds")

  if (file.exists(path_pred_draws)) {
    pred_draws <- readRDS(path_pred_draws)
  } else {
    est <- get_est()
    pred_draws <-
      est |>
      sccomp_predict(
        formula_composition = ~ age_decade * sex + ethnicity_groups_imputed,
        new_data =
          est |>
          attr("count_data") |>
          sccomp:::.subset(sample_id) |>
          distinct(sex) |>
          arrange(sex) |>
          mutate(age_decade = NA, ethnicity_groups_imputed = NA) |>
          mutate(sample_id = glue::glue("my_sample_{1:n()}")),
        number_of_draws = 100,
        summary_instead_of_draws = FALSE
      ) |>
      rename(proportion_mean = proportion) |>
      mutate(sample_id_original = sample_id) |>
      tidyr::unite("sample_id", c(sample_id, .draw), remove = FALSE)
    saveRDS(pred_draws, file = path_pred_draws)
  }

  if (file.exists(path_sex_adjust)) {
    sex_adjusted <- readRDS(path_sex_adjust)
  } else {
    est <- get_est()
    sex_adjusted <-
      est |>
      sccomp_predict(
        formula_composition = ~ age_decade * sex + ethnicity_groups_imputed,
        new_data = est |> attr("count_data") |> sccomp:::.subset(sample_id),
        number_of_draws = 100,
        summary_instead_of_draws = TRUE
      )
    saveRDS(sex_adjusted, file = path_sex_adjust)
  }
}
```

# Panel B – Sex differences for the top biased cell types

```{r define-boxplot-builder}
# Build one sex boxplot for a given set of cell types (facet order preserved).
build_sex_boxplot <- function(cell_types) {
  bind_rows(
    pred_draws %>%
      filter(L3 %in% cell_types) %>%
      dplyr::select(sample_id, L3, sex, proportion_mean) %>%
      mutate(version = 'prediction'),
    sex_adjusted %>%
      filter(L3 %in% cell_types) %>%
      dplyr::select(sample_id, L3, sex, proportion_mean) %>%
      mutate(version = 'adjusted')
  ) %>%
    filter(sex %in% c('male', 'female')) %>%
    mutate(L3_abbrev_base = recode(L3, !!!cell_type_abbrev_dict)) %>%
    left_join(
      test_mixed_max2 %>% dplyr::select(cell_type_unified_ensemble, c_FDR),
      by = c("L3" = "cell_type_unified_ensemble")
    ) %>%
    mutate(
      fdr_formatted = if_else(
        c_FDR < 0.001,
        gsub("^0\\.", ".", formatC(c_FDR, format = "e", digits = 2)),
        gsub("^0\\.", ".", formatC(c_FDR, format = "f", digits = 3))
      ),
      L3_abbrev = paste0(L3_abbrev_base, "\n(FDR: ", fdr_formatted, ')')
    ) %>%
    {
      lv <- distinct(., L3, L3_abbrev) %>% arrange(match(L3, cell_types)) %>% pull(L3_abbrev)
      mutate(., L3_abbrev = factor(L3_abbrev, levels = lv)) %>%
        dplyr::select(-L3_abbrev_base, -fdr_formatted)
    } %>%
    ggplot(aes(x = sex, y = proportion_mean %>% sqrt %>% asin)) +
    geom_jitter(
      data = . %>% filter(version == "adjusted"),
      width = 0.15, colour = "grey50", shape = '.', size = 0.6, alpha = 0.9
    ) +
    geom_boxplot(
      aes(fill = sex),
      data = . %>% filter(version == "prediction"),
      outlier.shape = NA, alpha = 0.3, size = 0.25
    ) +
    scale_fill_manual(values = colmap_sex) +
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.02))) +
    facet_wrap(~ L3_abbrev, nrow = 1, scales = "fixed") +
    labs(title = '', x = "Sex", y = "Proportion Mean (arcsin sqrt)") +
    theme_multipanel +
    theme(
      strip.text.x.top = element_text(margin = margin(0, 0, 0, 0, "pt")),
      strip.placement = "inside"
    )
}
```

```{r fig.width=14, fig.height=4, label="panel-B-boxplot-red"}
if (file.exists(path_box_red)) {
  Plot__boxplot_uniformly_red <- readRDS(path_box_red)
} else {
  # Three cell types with the strongest positive (male-biased) global effect
  uniformly_red_cell_types <-
    test_mixed_max2 %>% filter(sig, c_effect > 0) %>%
    arrange(-c_effect) %>% head(3) %>% pull(cell_type_unified_ensemble)
  Plot__boxplot_uniformly_red <- build_sex_boxplot(uniformly_red_cell_types)
  saveRDS(Plot__boxplot_uniformly_red, path_box_red)
}
print(Plot__boxplot_uniformly_red)
```

```{r fig.width=14, fig.height=4, label="panel-B-boxplot-blue"}
if (file.exists(path_box_blue)) {
  Plot__boxplot_uniformly_blue <- readRDS(path_box_blue)
} else {
  # Three cell types with the strongest negative (female-biased) global effect
  uniformly_blue_cell_types <-
    test_mixed_max2 %>% filter(sig, c_effect < 0) %>%
    arrange(c_effect) %>% head(3) %>% pull(cell_type_unified_ensemble)
  Plot__boxplot_uniformly_blue <- build_sex_boxplot(uniformly_blue_cell_types)
  saveRDS(Plot__boxplot_uniformly_blue, path_box_blue)
}
print(Plot__boxplot_uniformly_blue)
```

# Panel C – Heatmap of male-vs-female age effects

```{r fig.width=10, fig.height=9, label="panel-C-heatmap"}
if (file.exists(path_heatmap)) {
  Plot__heatmap_male_vs_female <- readRDS(path_heatmap)
} else {
  heat_data <-
    test_mixed_max |>
    mutate(size = -log10(c_FDR)) |>
    drop_na() |>
    complete(cell_type_unified_ensemble, Tissue,
             fill = list(c_FDR = 1, c_effect = 0, size = 0)) |>
    left_join(df_cells_ct) |>
    left_join(df_cells_tissue, by = join_by(Tissue == tissue)) |>
    mutate(tissue_short = recode(Tissue, !!!tissue_name_dict)) |>
    mutate(cell_type_abbrev = recode(cell_type_unified_ensemble, !!!cell_type_abbrev_dict_heatmap))

  Plot__heatmap_male_vs_female <-
    heat_data |>
    heatmap(
      .row = tissue_short, .column = cell_type_abbrev, .value = c_effect,
      name = "Effect\nsize",
      palette_value = circlize::colorRamp2(seq(-2, 2, length.out = 7), scico::scico(7, palette = "vik")),
      row_title = "Tissue", column_title = "Cell type"
    ) |>
    layer_point(c_FDR < 0.05, .size = size) |>
    annotation_bar(NCellsCellType, annotation_label = "log(Cells)") |>
    annotation_bar(NSamplesCellType, annotation_label = "log(Samples)") |>
    annotation_bar(NCellsTissue, annotation_label = "log(Cells)", annotation_name_rot = 45) |>
    annotation_bar(NSamplesTissue, annotation_label = "log(Samples)", annotation_name_rot = 45)

  saveRDS(Plot__heatmap_male_vs_female, path_heatmap)
}

Plot__heatmap_male_vs_female
```

# Panel D – Plasma-cell proportion trajectories by sex

```{r fig.width=15, fig.height=10, label="panel-D-line-chart"}
if (file.exists(path_line)) {

  Plot__line_chart_plasma <- readRDS(path_line)

} else {

  library(sccomp)  # sccomp_predict()

  x <- 'plasma'

  # Tissue-level predictions (own cache; recomputed from estimates if absent)
  outfile <- file.path(path_intermediate("fig5"),
    sprintf("predictions_age_decade_sex_tissue___%s___disease_FALSE___immune_only_TRUE.rds", cell_type_col)
  )
  if (file.exists(outfile)) {
    df_tissue <- readRDS(outfile)
  } else {
    estimates_age_bins <- get_estimates_renamed()
    new_df <- estimates_age_bins %>% attr("count_data")
    new_df$ethnicity_groups_imputed <- NA
    new_df$assay_groups___altered <- NA
    new_df$dataset_id___altered <- NA
    new_df <- new_df %>% dplyr::select(-!!cell_type_col) %>% dplyr::select(-n) %>% distinct()
    df_tissue <- sccomp_predict(
      estimates_age_bins,
      new_data = new_df,
      formula_composition = ~ 1 + age_decade + sex + age_decade:sex +
        (1 + age_decade + sex + age_decade:sex | tissue_groups),
      number_of_draws = 500,
      summary_instead_of_draws = TRUE
    ) |>
      dplyr::rename(cell_type_unified_ensemble = !!sym(cell_type_col))
    saveRDS(df_tissue, outfile)
  }

  tissue_groups_of_interest <-
    df_tissue %>%
    filter(cell_type_unified_ensemble == x) %>%
    distinct(tissue_groups, age_decade) %>% group_by(tissue_groups) %>%
    count(tissue_groups) %>% filter(n > 3) %>% pull(tissue_groups) %>%
    intersect(
      df_tissue %>%
        filter(cell_type_unified_ensemble == x) %>%
        distinct(tissue_groups, sex) %>% group_by(tissue_groups) %>%
        count(tissue_groups) %>% filter(n > 1) %>% pull(tissue_groups)
    ) %>%
    setdiff(c("adipose tissue", "endocrine system", "sensory-related structures", "spleen"))

  plasma_plot_data <-
    df_tissue |>
    dplyr::rename(Age = age_decade, Sex = sex, Tissue = tissue_groups) |>
    filter(!is.na(Age), cell_type_unified_ensemble == x, Sex != "unknown",
           Tissue %in% tissue_groups_of_interest) |>
    mutate(
      Age = Age |> fct_relevel(as.character(1:10)),
      tissue_groups_short = recode(Tissue, !!!tissue_name_dict),
      tissue_groups_short = factor(
        tissue_groups_short,
        levels = c(setdiff(unique(tissue_groups_short), "Large int"), "Large int")
      )
    ) |>
    # Predictions are identical for all source samples in a tissue/age/sex
    # cohort. Keep one plotted estimate and retain the cohort size explicitly.
    group_by(Tissue, tissue_groups_short, cell_type_unified_ensemble, Age, Sex) |>
    summarise(
      n_source_rows = n(),
      n_samples = n_distinct(sample_id),
      proportion_mean = first(proportion_mean),
      proportion_lower = first(proportion_lower),
      proportion_upper = first(proportion_upper),
      .groups = "drop"
    )

  Plot__line_chart_plasma <-
    plasma_plot_data |>
    ggplot() +
    geom_hline(yintercept = 0, lwd = 0.1) +
    geom_point(aes(x = Age, y = proportion_mean), shape = NA) +
    geom_line(aes(x = as.numeric(Age), y = proportion_mean, colour = Sex, fill = Sex),
              show.legend = FALSE) +
    geom_ribbon(aes(x = as.numeric(Age), colour = Sex, fill = Sex,
                    ymin = proportion_lower, ymax = proportion_upper),
                colour = NA, alpha = 0.1) +
    scale_colour_manual(values = c("#D7907B", "#0A014F"),
                        aesthetics = c("colour", "fill"),
                        guide = guide_legend(override.aes = list(alpha = 1))) +
    ggnewscale::new_scale_colour() +
    scale_colour_brewer(palette = "Set2", name = "Comparison") +
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.02))) +
    facet_wrap(~ tissue_groups_short, ncol = 4, scales = 'free_y') +
    theme_multipanel +
    theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1))

  saveRDS(Plot__line_chart_plasma, path_line)
}

print(Plot__line_chart_plasma)
```

# Final assembly

Layout (ported from `report/Fig_files/Fig4.R`): panel **A** spans the top;
below, the left column stacks **B** over **D** and the right column stacks the
**C** heatmap over a spacer.

```{r fig.width=10, fig.height=12, label="fig4-assembly"}
# --- Panel A (compact) ------------------------------------------------------
Plot__A <-
  Plot__age_sex_ethnicity_tissue_bar +
  theme(
    plot.title = element_blank(),
    legend.position = "bottom", legend.direction = "horizontal", legend.text.position = "bottom",
    strip.text.x.top = element_text(margin = margin(0, 0, 0, 0, "pt")),
    strip.placement = "inside",
    panel.spacing = unit(0, "pt"),
    plot.margin = margin(0, 0, 0, 0, "pt")
  )

# --- Panel B (red over blue) ------------------------------------------------
Plot__box_red <-
  Plot__boxplot_uniformly_red +
  guides(fill = "none") +
  theme(
    axis.line.x = element_blank(), axis.text.x = element_blank(), axis.title.x = element_blank(),
    panel.spacing = unit(0, "pt"), plot.margin = margin(0, 0, 0, 0, "pt")
  )

Plot__box_blue <-
  Plot__boxplot_uniformly_blue +
  guides(fill = "none") +
  theme(
    legend.position = "none", axis.title.y = element_blank(),
    panel.spacing = unit(0, "pt"), plot.margin = margin(0, 0, 0, 0, "pt")
  )

Plot__boxplot_2p <-
  (Plot__box_red / Plot__box_blue) +
  plot_layout() &
  theme(
    legend.position = "none", legend.direction = "horizontal", legend.text.position = "bottom",
    panel.spacing = unit(0, "pt"), plot.margin = margin(0, 0, 0, 0, "pt")
  )

# --- Panel D (thin lines, y starting at 0) ----------------------------------
Plot__line <- Plot__line_chart_plasma
Plot__line$layers <- lapply(Plot__line$layers, function(layer) {
  if (inherits(layer$geom, "GeomLine")) layer$aes_params$linewidth <- 0.25
  layer
})
Plot__line <-
  Plot__line +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  theme(
    plot.title = element_blank(),
    axis.text.x.bottom = element_text(angle = 0),
    legend.position = "none", legend.direction = "horizontal", legend.text.position = "bottom",
    strip.text.x.top = element_text(margin = margin(0, 0, 0, 0, "pt")),
    strip.placement = "inside",
    panel.spacing = unit(0, "pt"), plot.margin = margin(0, 0, 0, 0, "pt")
  )

# --- Panel C (heatmap -> ggplot); apply an FDR floor so point sizes are finite
heat_final <- {
  d <- Plot__heatmap_male_vs_female@data
  min_fdr <- min(d$c_FDR[d$c_FDR > 0], na.rm = TRUE)
  d$c_FDR[d$c_FDR == 0] <- min_fdr
  d %>%
    mutate(size = -log10(c_FDR)) %>%
    heatmap(
      .row = tissue_short, .column = cell_type_abbrev, .value = c_effect,
      name = "Effect\nsize",
      palette_value = circlize::colorRamp2(seq(-2, 2, length.out = 7), scico::scico(7, palette = "vik")),
      row_title = "Tissue", column_title = "Cell type"
    ) |>
    layer_point(c_FDR < 0.05, .size = size) |>
    annotation_bar(NCellsCellType, annotation_label = "log(Cells)") |>
    annotation_bar(NSamplesCellType, annotation_label = "log(Samples)") |>
    annotation_bar(NCellsTissue, annotation_label = "log(Cells)", annotation_name_rot = 45) |>
    annotation_bar(NSamplesTissue, annotation_label = "log(Samples)", annotation_name_rot = 45)
}

ht_temp <- heat_final %>% as_ComplexHeatmap()
ht_temp@heatmap_param$show_heatmap_legend <- FALSE
ht_temp@column_names_param$gp$fontsize <- 13
ht_temp@row_names_param$gp$fontsize <- 13

heatmap_ggplot <-
  as.ggplot(ht_temp) +
  theme(
    axis.line = element_blank(), axis.text = element_blank(),
    axis.title = element_blank(), axis.ticks = element_blank(),
    plot.margin = margin(0, 0, 0, 0, "pt")
  )

# --- Rows --------------------------------------------------------------------
row2_left <-
  (Plot__boxplot_2p / Plot__line) +
  plot_layout(heights = c(1, 1, 3)) &
  theme(panel.spacing = unit(0, "pt"), plot.margin = margin(0, 0, 0, 0, "pt"),
        legend.position = "bottom", legend.direction = "horizontal", legend.text.position = "bottom")

row2_right <-
  (heatmap_ggplot / plot_spacer()) +
  plot_layout(heights = c(6, 1)) &
  theme(panel.spacing = unit(0, "pt"), plot.margin = margin(0, 0, 0, 0, "pt"),
        legend.position = "bottom", legend.direction = "horizontal", legend.text.position = "bottom")

row2 <-
  (row2_left | row2_right) +
  plot_layout(widths = c(2, 2)) &
  theme(panel.spacing = unit(0, "pt"), plot.margin = margin(0, 0, 0, 0, "pt"),
        legend.position = "bottom", legend.direction = "horizontal", legend.text.position = "bottom")

# --- Final figure ------------------------------------------------------------
fig5 <-
  (Plot__A / row2) +
  plot_layout(guides = 'collect', heights = c(1, 6)) &
  theme(
    panel.spacing = unit(0, "pt"), plot.margin = margin(0, 0, 0, 0, "pt"),
    legend.position = "bottom", legend.direction = "horizontal", legend.text.position = "bottom",
    legend.key.size = unit(5, 'mm'),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 6)
  )

ggsave(
  plot = fig5,
  filename = file.path(path_figure_ready("fig5"), "fig5.pdf"),
  units = "mm", width = 183, height = 180, dpi = 500, scale = 1
)

fig5
```

# Supplementary tables

Export plot statistics from ggplot / heatmap `@data` slots for all Fig 5 panels.

```{r export-supplementary-tables}
message(
  "Panel caches updated. Run `Rscript report/build_supplementary_workbooks.R` ",
  "to rebuild the split-tab workbooks with separate plot-data, summary and test sheets."
)
```

# Save panels

```{r save-all-plots, eval=FALSE}
# Persist all panel objects for downstream figure assembly / re-use.
save(
  list = ls(pattern = "Plot__"),
  file = file.path(path_figure_ready("fig5"), "Fig5_plot_data_all.RData")
)

sessionInfo()
```
