---
title: "Figure 5 — Sex differences in immune ageing across tissues"
date: today
date-format: "DD MMM YYYY"
title-block-banner: true
author: 
  - name: Chen Zhan
    orcid: 0000-0002-4794-8339
    email: chen.zhan@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
format: 
  html:
    theme: minty
    df-print: paged
    code-line-numbers: true
    embed-resources: true
    code-fold: true
    code-fold-default: true
    toc: true
    toc-location: left
    toc-title: "Menu"
knitr:
  opts_chunk: 
    message: false
    warning: false
comments:
  hypothesis: 
    theme: clean
editor: source
abstract: |
  Sex differences in immune-cell composition across age and tissues.
---

# Overview

This report reproduces **Figure 5** (sex differences in immune-cell composition
across age and tissues). The final figure is a patchwork of:

- **A** – Donor counts by age group and sex across tissues (signed log bar chart).
- **B** – Age-split sex×age boxplots for the four lowest-FDR sex-difference
  ageing effects (later−earlier male−female gap change, split at age decade 5;
  arcsin-sqrt proportions).
- **C** – Heatmap of male-vs-female age effects (cell type × tissue).
- **D** – Plasma-cell proportion trajectories by sex across tissues.

Every panel **reads its saved object from `vignettes/data/figure_ready/` by default** and
only recomputes (loading the sccomp estimates, running `sccomp_test` /
`sccomp_predict`) when the cache is absent. This keeps the report fast to render
while remaining fully reproducible.

The cache filenames retain their historical `Fig4_` prefix for compatibility;
all manuscript-facing figures and supplementary exports use Figure 5.

# Library


```{r publication-paths}
# Portable publication data roots (no HPC absolute paths).
.path_candidates <- c(
  file.path("..", "R", "paths.R"),
  if (requireNamespace("here", quietly = TRUE)) {
    file.path(here::here(), "vignettes", "R", "paths.R")
  },
  file.path("R", "paths.R")
)
.path_candidates <- .path_candidates[!is.na(.path_candidates) & nzchar(.path_candidates)]
.path_file <- .path_candidates[file.exists(.path_candidates)][1]
if (is.na(.path_file)) {
  stop(
    "Cannot find vignettes/R/paths.R. Render from the immuneHealthyBodyMap ",
    "repository root or from vignettes/reports/, or set IMMUNE_HEALTHY_BODY_MAP_ROOT."
  )
}
source(.path_file)
fig_cache_dir <- path_figure_ready()
dir.create(path_intermediate("fig2"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_intermediate("fig4"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_intermediate("fig5"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_figure_ready("fig1"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_figure_ready("fig2"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_figure_ready("fig4"), recursive = TRUE, showWarnings = FALSE)
dir.create(path_figure_ready("fig5"), recursive = TRUE, showWarnings = FALSE)
```

```{r setup-libraries}
# Set your R library path here if needed
# .libPaths removed for portability; use renv/conda or site library
library(here)  # project-root aware paths

# Ensure required output directories exist
required_dirs <- c(
  path_figure_ready("fig5"),
  path_intermediate("fig5")
)
invisible(lapply(required_dirs, function(d) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE, showWarnings = FALSE)
}))

library(magrittr)
library(tidyverse)
library(glue)
library(patchwork)          # figure assembly
library(ggplotify)          # as.ggplot() for the heatmap panel
library(ComplexHeatmap)
library(tidyHeatmap)        # heatmap(), layer_point(), annotation_bar()
library(ggnewscale)         # second colour scale in the line chart
library(scico)              # 'vik' diverging palette for the heatmap
library(circlize)           # colorRamp2()
library(ggplot2)
library(grid)               # unit()
library(ggh4x)              # facetted_pos_scales() for Panel B free-y limits
# NOTE: `sccomp` and `cellNexus` are loaded on-demand inside the computation
# chunks (panels A–D) so the report still renders from cache without them.
```

# Setup: theme, dictionaries, palettes and paths

```{r define-theme}
theme_multipanel <-
  theme_bw() +
  theme(
    panel.border = element_blank(),
    axis.line = element_line(size = 0.1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    strip.background = element_blank(),
    axis.title.y = element_text(margin = margin(0, 0, 0, 0), size = 7),
    axis.title.x = element_text(margin = margin(0, 0, 0, 0), size = 7),
    panel.spacing.x = unit(0.1, "lines"),
    axis.text.x = element_text(size = 6),
    axis.text.y = element_text(size = 6),
    strip.text.x = element_text(size = 7),
    strip.text.y = element_text(size = 7),
    legend.key.size = unit(5, 'mm'),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 6),
    strip.clip = "off",
    plot.title = element_text(size = 7),
    axis.line.x = element_line(size = 0.2),
    axis.line.y = element_line(size = 0.2),
    axis.ticks.x = element_line(size = 0.2),
    axis.ticks.y = element_line(size = 0.2)
  )
```

```{r define-dictionaries}
# Shortened tissue names (shared across all figures)
tissue_name_dict <- c(
  "blood" = "Blood",
  "sensory-related structures" = "Sensory",
  "cerebral lobes and cortical areas" = "Cerebral cortex",
  "limbic and basal systems" = "Limbic/basal",
  "liver" = "Liver",
  "bone marrow" = "Bone marrow",
  "large intestine" = "Large int",
  "lymphatic system" = "Lymphatic",
  "small intestine" = "Small int",
  "respiratory system" = "Respiratory",
  "renal system" = "Renal",
  "brainstem and cerebellar structures" = "Hindbrain",
  "cardiovascular system" = "Cardiovascular",
  "general brain and major structures" = "Brain",
  "vasculature" = "Vasculature",
  "ovary" = "Ovary",
  "integumentary system (skin)" = "Skin",
  "nasal, oral, and pharyngeal regions" = "Pharyngeal",
  "muscular system (skeletal muscles)" = "Skeletal muscle",
  "prostate" = "Prostate",
  "breast" = "Breast",
  "digestive system (general)" = "Digestive",
  "oesophagus" = "Oesophagus",
  "female reproductive system" = "Female repro",
  "spleen" = "Spleen",
  "peritoneal and abdominal cavity structures" = "Peritoneal",
  "thymus" = "Thymus",
  "adipose tissue" = "Adipose",
  "trachea" = "Trachea",
  "endocrine system" = "Endocrine",
  "pancreas" = "Pancreas",
  "miscellaneous glands" = "Glands",
  "stomach" = "Stomach",
  "gallbladder" = "Gallbladder"
)

# Cell-type abbreviations (single-line, used for boxplot facets)
cell_type_abbrev_dict <- c(
  "b memory" = "B Memory", "b naive" = "B Naive",
  "cd14 mono" = "CD14 Mono", "cd16 mono" = "CD16 Mono",
  "cd4 fh em" = "CD4 FH EM", "cd4 naive" = "CD4 Naive", "cd4 tcm" = "CD4 TCM",
  "cd4 th1 em" = "CD4 Th1 EM", "cd4 th1/th17 em" = "CD4 Th1/Th17 EM",
  "cd4 th17 em" = "CD4 Th17 EM", "cd4 th2 em" = "CD4 Th2 EM", "treg" = "Treg",
  "cd8 naive" = "CD8 Naive", "cd8 tcm" = "CD8 TCM", "cd8 tem" = "CD8 TEM",
  "cdc" = "CDC", "granulocyte" = "Gran", "ilc" = "ILC", "macrophage" = "Mac",
  "mait" = "MAIT", "mast" = "Mast", "nk" = "NK", "nkt" = "NKT", "pdc" = "PDC",
  "tgd" = "TGD", "plasma" = "Plasma"
)

# Cell-type abbreviations for the heatmap columns
cell_type_abbrev_dict_heatmap <- c(
  "macrophage" = "Mac", "cd4 th1 em" = "CD4 Th1 EM", "cd4 th1/th17 em" = "CD4 Th1/17 EM",
  "cd4 th17 em" = "CD4 Th17 EM", "cd4 th2 em" = "CD4 Th2 EM", "cd8 naive" = "CD8 Naive",
  "plasma" = "Plasma", "granulocyte" = "Gran", "cd4 naive" = "CD4 Naive",
  "b naive" = "B Naive", "nk" = "NK", "nkt" = "NKT", "cdc" = "CDC", "pdc" = "PDC",
  "b memory" = "B Memory", "cd14 mono" = "CD14 Mono", "cd16 mono" = "CD16 Mono",
  "cd4 fh em" = "CD4 FH EM", "cd4 tcm" = "CD4 TCM", "cd8 tcm" = "CD8 TCM",
  "cd8 tem" = "CD8 TEM", "ilc" = "ILC", "mait" = "MAIT", "mast" = "Mast",
  "tgd" = "TGD", "treg" = "Treg"
)

colmap_sex <- c("female" = "#D7907B", "male" = "#0A014F")

# Keep raw proportions in the data; transform only the axis (same as Fig2).
asin_sqrt_trans <- scales::trans_new(
  name = "asin_sqrt",
  transform = function(x) asin(sqrt(x)),
  inverse = function(y) sin(y)^2
)
```

```{r define-paths-and-loaders}
# --- sccomp estimates (large; loaded lazily only when a cache is missing) -----
# Resolve the configured L3 sccomp estimates object.
estimates_file <-
  path_sccomp_estimates("estimates_age_decade___L3___disease_FALSE___immune_only_TRUE.rds")
cell_type_col <- "L3"

.est_raw <- NULL
get_est <- function() {
  if (is.null(.est_raw)) .est_raw <<- readRDS(estimates_file)
  .est_raw
}
get_estimates_renamed <- function() {
  get_est() |> dplyr::rename(cell_type_unified_ensemble = !!sym(cell_type_col))
}

# --- Cached panel objects -----------------------------------------------------
fig_files <- path_figure_ready("fig5")
# Prefer Fig5_-prefixed caches; historical Fig4_* names remain for older panels.
path_bar        <- file.path(fig_files, "Fig4_Plot__age_sex_ethnicity_tissue_bar_consensus.rds")
path_bar_all_counts <- file.path(fig_files, "Fig5_A_donor_counts_all_tissues.rds")
path_heatmap    <- file.path(fig_files, "Fig4_Plot__heatmap_male_vs_female.rds")
path_box_red    <- file.path(fig_files, "Fig5_Plot__boxplot_uniformly_red.rds")
path_box_blue   <- file.path(fig_files, "Fig5_Plot__boxplot_uniformly_blue.rds")
path_line       <- file.path(fig_files, "Fig5_Plot__line_chart_plasma.rds")

save_fig5_proportion_plot <- function(plot, path) {
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  saveRDS(plot, path)
  message("Saved ", path)
  invisible(plot)
}
```

# Panel A – Donor counts by age group and sex

```{r fig.width=25, fig.height=5, label="panel-A-bar"}
rebuild_panel_a <- tolower(Sys.getenv("REBUILD_FIG5_PANEL_A", "false")) %in%
  c("1", "true", "yes")

if (file.exists(path_bar) && file.exists(path_bar_all_counts) && !rebuild_panel_a) {

  Plot__age_sex_ethnicity_tissue_bar <- readRDS(path_bar)
  donor_count_all <- readRDS(path_bar_all_counts)

} else {

  library(cellNexus)  # get_metadata_url() / get_metadata()

  edit_covariates_path <- path_edit_covariates()
  source(edit_covariates_path)

  cell_metadata <-
    get_publication_metadata() %>%
    filter(disease == "normal") |> filter(age_days > 365)

  donor_metadata <-
    cell_metadata |>
    distinct(donor_id, self_reported_ethnicity, assay, age_days, sex, tissue, disease) |>
    as_tibble() %>%
    edit_covariates() %>%
    filter(sex %in% c("female", "male"))

  donor_count_all <-
    donor_metadata |>
    mutate(age_groups = case_when(
      age_decade %in% c('1', '2', '3') ~ 'Young',
      age_decade %in% c('4', '5', '6') ~ 'Mid',
      age_decade %in% c('7', '8', '9') ~ 'Old',
      .default = 'Old'
    )) |>
    # Count each donor once per tissue/age/sex cohort. A donor can have several
    # samples or assay records within a tissue, which must not inflate panel A.
    distinct(donor_id, age_groups, sex, tissue_groups) |>
    dplyr::count(age_groups, sex, tissue_groups, name = "donor_count") |>
    as_tibble() |>
    mutate(
      tissue_label = recode(tissue_groups, !!!tissue_name_dict),
      tissue_label = stringr::str_replace_all(tissue_label, "/", "/\n"),
      tissue_label = stringr::str_wrap(tissue_label, width = 10)
    )

  saveRDS(donor_count_all, path_bar_all_counts)

  tissue_order <- donor_count_all |>
    group_by(tissue_label) |>
    summarise(total_n = sum(donor_count), .groups = "drop") |>
    arrange(desc(total_n)) |>
    pull(tissue_label)

  # Keep the 25 best-represented tissues for legibility
  top_tissues <- head(tissue_order, 25)

  df_bar <- donor_count_all |>
    filter(tissue_label %in% top_tissues) |>
    mutate(
      tissue_groups = factor(tissue_label, levels = tissue_order),
      signed_donor_count = ifelse(sex == "female", -donor_count, donor_count)
    ) |>
    mutate(age_groups = factor(age_groups, levels = c("Old", "Mid", "Young")),
           Sign = sign(signed_donor_count))

  Plot__age_sex_ethnicity_tissue_bar <-
    df_bar |>
    ggplot(aes(x = age_groups, y = log(abs(signed_donor_count)) * Sign, fill = sex)) +
    geom_bar(stat = "identity", position = "stack") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
    facet_grid(. ~ tissue_groups) +
    scale_fill_manual(values = colmap_sex, guide = guide_legend(nrow = 1)) +
    labs(x = "", y = expression(log(abs(donor[count])) %.% Sign), fill = "Sex") +
    theme_multipanel +
    theme(
      strip.placement = "inside",
      strip.text.x.top = element_text(angle = 45, hjust = 1),
      strip.background = element_blank(),
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
      plot.title = element_blank()
    )

  saveRDS(Plot__age_sex_ethnicity_tissue_bar, path_bar)
}

print(Plot__age_sex_ethnicity_tissue_bar)
```

# Sex-difference composition tests (Panel C heatmap inputs)

Tissue-level male-vs-female contrasts for panel **C**. Skipped when the heatmap
cache already exists.

```{r sex-difference-sccomp}
need_heatmap <- !file.exists(path_heatmap)

if (need_heatmap) {

  library(sccomp)  # sccomp_test()

  estimates_age_bins <- get_estimates_renamed()
  load(file.path(path_intermediate("fig2"), "df_cell_statistics.RData"))  # df_cells_ct, df_cells_tissue

  age_bins      <- c("1", "2", "3", "4", "6", "7", "8", '9', '10')
  age_binary    <- setNames(c("Early", "Early", "Early", "Early",
                              "Late", "Late", "Late", "Late", "Late"), age_bins)

  # Parameter table for age / sex / age:sex (fixed + tissue-random)
  param_df <-
    estimates_age_bins |>
    filter(grepl("age_decade|sex", parameter)) |>
    mutate(
      params = str_split(parameter, "___", simplify = TRUE),
      params = (\(x) { colnames(x) <- c("parameter_tissue_removed", "Tissue"); as.data.frame(x) })(params)
    ) |>
    dplyr::select(params, parameter) |>
    tidyr::unpack(params) |>
    mutate(
      Age = str_extract(parameter_tissue_removed, "(?<=age_decade)(10|[1-9])"),
      Sex = str_extract(parameter_tissue_removed, "(?<=sex)(male|unknown)"),
      Tissue = if_else(Tissue == "", "global", Tissue),
      AgeBinary = age_binary[Age]
    ) |>
    distinct()

  # -- Tissue-specific male-vs-female contrasts (Panel C) ----------------------
  tissue_age_combo <- param_df %>%
    filter(Sex == 'male', Tissue != 'global', !is.na(Age)) %>% distinct(Tissue, Age)

  contrasts_sex_diff_overall <-
    tissue_age_combo %>%
    mutate(contr = map2_chr(Tissue, Age, ~ {
      param_df %>%
        filter(Tissue %in% c(.x, 'global'), Age %in% c(.y, NA), Sex == 'male') %>%
        mutate(parameter = glue::glue('`{parameter}`')) %>% pull(parameter) %>%
        paste0(collapse = ' + ')
    })) %>%
    group_by(Tissue) %>%
    summarise(contr = { n <- n(); glue::glue("({paste(contr, collapse = ' + ')}) / {n}") },
              .groups = "drop") %>%
    mutate(contrast_name = Tissue)

  contrasts <- contrasts_sex_diff_overall %>%
    dplyr::select(contrast_name, contr) |> deframe()

  test_mixed_all <- estimates_age_bins |>
    sccomp_test(contrasts = contrasts) |>
    dplyr::rename(cell_type_unified_ensemble = !!sym(cell_type_col))

  # Contrast names are tissue names for these overall tissue contrasts.
  test_mixed_max <- test_mixed_all |>
    rename(Tissue = parameter) |>
    mutate(significant = c_FDR < 0.05)
}
```

# Panel B – Age-split sex×age boxplots (top-4 by FDR)

Body-level sex-difference ageing contrast split at age decade 5:
`(later−earlier males) − (later−earlier females)`. Fig. 5b highlights the four
lowest-FDR effects (no same-signed filter). Boxes use sex × earlier/later
groups; grey points are sample-level adjusted proportions.

```{r define-boxplot-builder}
earlier_age_decades <- as.character(1:4)
later_age_decades <- as.character(5:10)
age_sex_group_levels <- c(
  "female earlier", "female later", "male earlier", "male later"
)

assign_age_sex_group <- function(df) {
  df %>%
    mutate(
      age_decade_chr = as.character(age_decade),
      age_split = case_when(
        age_decade_chr %in% earlier_age_decades ~ "earlier",
        age_decade_chr %in% later_age_decades ~ "later",
        TRUE ~ NA_character_
      ),
      age_sex_group = paste(sex, age_split),
      age_sex_group = factor(age_sex_group, levels = age_sex_group_levels)
    ) %>%
    filter(!is.na(age_split), sex %in% c("female", "male"))
}

standardise_cell_col <- function(df) {
  if ("cell_type_unified_ensemble" %in% names(df) && !"L3" %in% names(df)) {
    df <- df %>% dplyr::rename(L3 = cell_type_unified_ensemble)
  } else if ("L3" %in% names(df) && !"cell_type_unified_ensemble" %in% names(df)) {
    df <- df %>% mutate(cell_type_unified_ensemble = L3)
  }
  df
}

format_fdr_label <- function(x) {
  if_else(
    x < 0.001,
    gsub("^0\\.", ".", formatC(x, format = "e", digits = 2)),
    gsub("^0\\.", ".", formatC(x, format = "f", digits = 3))
  )
}

format_y_prop_label <- function(x) {
  gsub("^0\\.", ".", formatC(x, format = "fg", digits = 3, drop0trailing = TRUE))
}

resolve_fig5_analysis_cache <- function(filename) {
  candidates <- c(
    path_intermediate("fig5", filename),
    {
      arch <- path_age_clock_archive("data", "age_fig", filename)
      if (is.null(arch)) character() else arch
    },
    {
      arch <- path_age_clock_archive("data", filename)
      if (is.null(arch)) character() else arch
    }
  )
  hit <- candidates[nzchar(candidates) & file.exists(candidates)][1]
  if (is.na(hit)) NA_character_ else hit
}

build_age_sex_boxplot_df <- function(cell_types, test_df) {
  bind_rows(
    pred_draws %>%
      filter(L3 %in% cell_types) %>%
      dplyr::select(sample_id, L3, sex, age_decade, age_split, age_sex_group, proportion_mean) %>%
      mutate(version = "prediction"),
    sex_adjusted %>%
      filter(L3 %in% cell_types) %>%
      dplyr::select(sample_id, L3, sex, age_decade, age_split, age_sex_group, proportion_mean) %>%
      mutate(version = "adjusted")
  ) %>%
    mutate(L3_abbrev = recode(L3, !!!cell_type_abbrev_dict)) %>%
    left_join(
      test_df %>% dplyr::select(cell_type_unified_ensemble, c_FDR, c_effect),
      by = c("L3" = "cell_type_unified_ensemble")
    ) %>%
    mutate(fdr_formatted = format_fdr_label(c_FDR)) %>%
    {
      df_temp <- .
      unique_levels <- df_temp %>%
        distinct(L3, L3_abbrev) %>%
        arrange(match(L3, cell_types)) %>%
        pull(L3_abbrev)
      df_temp %>%
        mutate(
          L3 = factor(L3, levels = cell_types),
          L3_abbrev = factor(L3_abbrev, levels = unique_levels),
          age_sex_group = factor(age_sex_group, levels = age_sex_group_levels)
        )
    }
}

plot_age_sex_boxplot <- function(
    df,
    box_version = c("adjusted", "prediction"),
    y_pad = 0.08,
    ncol = 4,
    nrow = 1
) {
  box_version <- match.arg(box_version)
  box_df <- df %>% filter(version == box_version)
  jitter_df <- df %>% filter(version == "adjusted")
  if (base::nrow(box_df) == 0) stop("No rows for box_version = '", box_version, "'.")

  facet_levels <- if (is.factor(box_df$L3_abbrev)) {
    levels(box_df$L3_abbrev)[levels(box_df$L3_abbrev) %in% unique(as.character(box_df$L3_abbrev))]
  } else {
    unique(as.character(box_df$L3_abbrev))
  }

  box_lims <- box_df %>%
    group_by(L3_abbrev, age_sex_group) %>%
    summarise(
      stats = list(boxplot.stats(proportion_mean, do.conf = FALSE, do.out = FALSE)$stats),
      .groups = "drop"
    ) %>%
    mutate(
      ymin = purrr::map_dbl(stats, 1),
      ymax = purrr::map_dbl(stats, 5)
    ) %>%
    group_by(L3_abbrev) %>%
    summarise(ymin = min(ymin), ymax = max(ymax), .groups = "drop") %>%
    mutate(
      ymin_t = asin(sqrt(pmax(ymin, 0))),
      ymax_t = asin(sqrt(pmax(ymax, 0))),
      pad = y_pad * pmax(ymax_t - ymin_t, 0.02),
      ymin = sin(pmax(ymin_t - pad, 0))^2,
      ymax = sin(pmin(ymax_t + pad, pi / 2))^2,
      fdr_y = ymin + 0.92 * (ymax - ymin)
    ) %>%
    mutate(L3_abbrev = factor(as.character(L3_abbrev), levels = facet_levels)) %>%
    arrange(L3_abbrev)

  fdr_labels <- box_df %>%
    distinct(L3_abbrev, fdr_formatted) %>%
    left_join(box_lims %>% dplyr::select(L3_abbrev, fdr_y), by = "L3_abbrev") %>%
    mutate(
      age_sex_group = factor(
        age_sex_group_levels[ceiling(length(age_sex_group_levels) / 2)],
        levels = age_sex_group_levels
      ),
      proportion_mean = fdr_y,
      label = paste0("FDR: ", fdr_formatted)
    )

  y_scales <- purrr::map(seq_len(base::nrow(box_lims)), function(i) {
    facet_val <- as.character(box_lims$L3_abbrev[i])
    rlang::new_formula(
      lhs = rlang::expr(L3_abbrev == !!facet_val),
      rhs = scale_y_continuous(
        trans = "asn",
        limits = c(box_lims$ymin[i], box_lims$ymax[i]),
        labels = format_y_prop_label,
        oob = scales::oob_keep,
        expand = expansion(mult = 0)
      )
    )
  })

  ggplot(box_df, aes(x = age_sex_group, y = proportion_mean)) +
    geom_jitter(
      data = jitter_df, width = 0.15, height = 0,
      colour = "grey50", shape = 16, size = 0.15, alpha = 0.2
    ) +
    geom_boxplot(aes(fill = sex), outlier.shape = NA, alpha = 0.3, linewidth = 0.25) +
    geom_text(
      data = fdr_labels, aes(label = label),
      size = 2, colour = "grey20", vjust = 0.5
    ) +
    scale_fill_manual(values = colmap_sex) +
    scale_x_discrete(labels = c(
      "female earlier" = "F(e)", "female later" = "F(l)",
      "male earlier" = "M(e)", "male later" = "M(l)"
    )) +
    facet_wrap(~ L3_abbrev, ncol = ncol, nrow = nrow, scales = "free_y") +
    ggh4x::facetted_pos_scales(y = y_scales) +
    labs(
      title = "",
      x = "Sex × Age groups (earlier vs later)",
      y = "Proportion Mean"
    ) +
    theme_multipanel +
    theme(
      strip.text.x.top = element_text(margin = margin(0, 0, 0, 0, "pt")),
      strip.placement = "inside"
    )
}
```

```{r load-panel-B-inputs}
rebuild_fig5_boxplots <- tolower(Sys.getenv("REBUILD_FIG5_BOXPLOTS", "false")) %in%
  c("1", "true", "yes")

# Rebuild when missing, forced, or when an older sex-only (no age split) cache remains.
needs_boxplot_rebuild <- function(path) {
  if (!file.exists(path) || rebuild_fig5_boxplots) return(TRUE)
  p <- readRDS(path)
  x_map <- tryCatch(rlang::as_label(p$mapping$x), error = function(e) "")
  has_age_sex <- !is.null(p$data) && "age_sex_group" %in% names(p$data)
  y_map <- tryCatch(rlang::as_label(p$mapping$y), error = function(e) "")
  !has_age_sex || !grepl("age_sex_group", x_map) || grepl("asin|sqrt", y_map)
}

need_box_inputs <- needs_boxplot_rebuild(path_box_red) ||
  needs_boxplot_rebuild(path_box_blue)

if (need_box_inputs) {
  path_split_csv <- resolve_fig5_analysis_cache("fig5_test_sex_diff_split_at_5.csv")
  path_split_rds <- resolve_fig5_analysis_cache("Fig4_split_at_5_compact.rds")
  path_pred <- resolve_fig5_analysis_cache("sccomp_sex_age_split_pred_draws_fig4.rds")
  path_adj_proc <- resolve_fig5_analysis_cache(
    "sccomp_sex_age_split_adjusted_fig4_processed.rds"
  )
  path_adj <- resolve_fig5_analysis_cache("sccomp_sex_age_split_adjusted_fig4.rds")

  if (!is.na(path_split_csv)) {
    test_sex_diff_split_at_5 <- readr::read_csv(path_split_csv, show_col_types = FALSE) %>%
      mutate(
        sig = if ("sig" %in% names(.)) as.logical(sig) else c_FDR < 0.05,
        sign_age_trend = as.integer(sign_age_trend)
      )
  } else if (!is.na(path_split_rds) && file.info(path_split_rds)$size < 5e6) {
    test_sex_diff_split_at_5 <- readRDS(path_split_rds)
  } else {
    stop(
      "Missing fig5_test_sex_diff_split_at_5.csv under path_intermediate('fig5') ",
      "or AGE_CLOCK_ARCHIVE/data/age_fig/.",
      call. = FALSE
    )
  }

  if (is.na(path_pred)) {
    stop(
      "Missing sccomp_sex_age_split_pred_draws_fig4.rds under path_intermediate('fig5') ",
      "or AGE_CLOCK_ARCHIVE/data/age_fig/.",
      call. = FALSE
    )
  }
  pred_draws <- readRDS(path_pred) %>% standardise_cell_col() %>% assign_age_sex_group()

  if (!is.na(path_adj_proc)) {
    sex_adjusted <- readRDS(path_adj_proc) %>% standardise_cell_col()
    if (!"age_sex_group" %in% names(sex_adjusted)) {
      sex_adjusted <- assign_age_sex_group(sex_adjusted)
    }
  } else if (!is.na(path_adj)) {
    library(sccomp)
    est <- get_est()
    sex_adjusted <- readRDS(path_adj) %>%
      left_join(
        est |> attr("count_data") |> sccomp:::.subset(sample_id),
        by = "sample_id"
      ) %>%
      rename(proportion_mean = adjusted_proportion) %>%
      standardise_cell_col() %>%
      assign_age_sex_group()
    saveRDS(sex_adjusted, path_intermediate("fig5", "sccomp_sex_age_split_adjusted_fig4_processed.rds"))
  } else {
    stop(
      "Missing sccomp_sex_age_split_adjusted_fig4_processed.rds (or raw adjusted RDS) ",
      "under path_intermediate('fig5') or AGE_CLOCK_ARCHIVE/data/age_fig/.",
      call. = FALSE
    )
  }
}
```

```{r fig.width=14, fig.height=4, label="panel-B-boxplot-top4"}
# Force rebuild when an older dual-row / same-signed cache is still present.
if (!needs_boxplot_rebuild(path_box_red)) {
  p_chk <- readRDS(path_box_red)
  n_facets <- if (!is.null(p_chk$data$L3)) length(unique(as.character(p_chk$data$L3))) else 0L
  if (n_facets != 4L) {
    # stale selection size
    unlink(path_box_red)
  }
}

if (!needs_boxplot_rebuild(path_box_red) && file.exists(path_box_red)) {
  Plot__boxplot_fig5b <- readRDS(path_box_red)
} else {
  # Overall top-4 by FDR (no same-signed / direction filter)
  fig5b_cell_types <-
    test_sex_diff_split_at_5 %>%
    arrange(c_FDR, desc(abs(c_effect))) %>%
    head(4) %>%
    pull(cell_type_unified_ensemble)
  Plot__boxplot_fig5b <- plot_age_sex_boxplot(
    build_age_sex_boxplot_df(fig5b_cell_types, test_sex_diff_split_at_5),
    box_version = "adjusted",
    ncol = 4,
    nrow = 1
  )
  save_fig5_proportion_plot(Plot__boxplot_fig5b, path_box_red)
  # Keep legacy blue cache path in sync so older assembly scripts do not reuse stale plots.
  save_fig5_proportion_plot(Plot__boxplot_fig5b, path_box_blue)
}
Plot__boxplot_uniformly_red <- Plot__boxplot_fig5b
Plot__boxplot_uniformly_blue <- Plot__boxplot_fig5b
print(Plot__boxplot_fig5b)
```

# Panel C – Heatmap of male-vs-female age effects

```{r fig.width=10, fig.height=9, label="panel-C-heatmap"}
if (file.exists(path_heatmap)) {
  Plot__heatmap_male_vs_female <- readRDS(path_heatmap)
} else {
  heat_data <-
    test_mixed_max |>
    mutate(size = -log10(c_FDR)) |>
    drop_na() |>
    complete(cell_type_unified_ensemble, Tissue,
             fill = list(c_FDR = 1, c_effect = 0, size = 0)) |>
    left_join(df_cells_ct) |>
    left_join(df_cells_tissue, by = join_by(Tissue == tissue)) |>
    mutate(tissue_short = recode(Tissue, !!!tissue_name_dict)) |>
    mutate(cell_type_abbrev = recode(cell_type_unified_ensemble, !!!cell_type_abbrev_dict_heatmap))

  Plot__heatmap_male_vs_female <-
    heat_data |>
    heatmap(
      .row = tissue_short, .column = cell_type_abbrev, .value = c_effect,
      name = "Effect\nsize",
      palette_value = circlize::colorRamp2(seq(-2, 2, length.out = 7), scico::scico(7, palette = "vik")),
      row_title = "Tissue", column_title = "Cell type"
    ) |>
    layer_point(c_FDR < 0.05, .size = size) |>
    annotation_bar(NCellsCellType, annotation_label = "log(Cells)") |>
    annotation_bar(NSamplesCellType, annotation_label = "log(Samples)") |>
    annotation_bar(NCellsTissue, annotation_label = "log(Cells)", annotation_name_rot = 45) |>
    annotation_bar(NSamplesTissue, annotation_label = "log(Samples)", annotation_name_rot = 45)

  saveRDS(Plot__heatmap_male_vs_female, path_heatmap)
}

Plot__heatmap_male_vs_female
```

# Panel D – Plasma-cell proportion trajectories by sex

```{r fig.width=15, fig.height=10, label="panel-D-line-chart"}
rebuild_fig5_line <- tolower(Sys.getenv("REBUILD_FIG5_LINE", "false")) %in%
  c("1", "true", "yes")
needs_line_rebuild <- function(path) {
  if (!file.exists(path) || rebuild_fig5_line) return(TRUE)
  p <- readRDS(path)
  ysc <- tryCatch(p$scales$get_scales("y"), error = function(e) NULL)
  tr <- tryCatch(ysc$get_transformation()$name, error = function(e) NA_character_)
  !identical(tr, "asin_sqrt")
}

if (!needs_line_rebuild(path_line)) {

  Plot__line_chart_plasma <- readRDS(path_line)

} else {

  library(sccomp)  # sccomp_predict()

  x <- 'plasma'

  # Tissue-level predictions (own cache; recomputed from estimates if absent)
  outfile <- file.path(path_intermediate("fig5"),
    sprintf("predictions_age_decade_sex_tissue___%s___disease_FALSE___immune_only_TRUE.rds", cell_type_col)
  )
  if (file.exists(outfile)) {
    df_tissue <- readRDS(outfile)
  } else {
    estimates_age_bins <- get_estimates_renamed()
    new_df <- estimates_age_bins %>% attr("count_data")
    new_df$ethnicity_groups_imputed <- NA
    new_df$assay_groups___altered <- NA
    new_df$dataset_id___altered <- NA
    new_df <- new_df %>% dplyr::select(-!!cell_type_col) %>% dplyr::select(-n) %>% distinct()
    df_tissue <- sccomp_predict(
      estimates_age_bins,
      new_data = new_df,
      formula_composition = ~ 1 + age_decade + sex + age_decade:sex +
        (1 + age_decade + sex + age_decade:sex | tissue_groups),
      number_of_draws = 500,
      summary_instead_of_draws = TRUE
    ) |>
      dplyr::rename(cell_type_unified_ensemble = !!sym(cell_type_col))
    saveRDS(df_tissue, outfile)
  }

  tissue_groups_of_interest <-
    df_tissue %>%
    filter(cell_type_unified_ensemble == x) %>%
    distinct(tissue_groups, age_decade) %>% group_by(tissue_groups) %>%
    count(tissue_groups) %>% filter(n > 3) %>% pull(tissue_groups) %>%
    intersect(
      df_tissue %>%
        filter(cell_type_unified_ensemble == x) %>%
        distinct(tissue_groups, sex) %>% group_by(tissue_groups) %>%
        count(tissue_groups) %>% filter(n > 1) %>% pull(tissue_groups)
    ) %>%
    setdiff(c("adipose tissue", "endocrine system", "sensory-related structures", "spleen"))

  plasma_plot_data <-
    df_tissue |>
    dplyr::rename(Age = age_decade, Sex = sex, Tissue = tissue_groups) |>
    filter(!is.na(Age), cell_type_unified_ensemble == x, Sex != "unknown",
           Tissue %in% tissue_groups_of_interest) |>
    mutate(
      Age = Age |> fct_relevel(as.character(1:10)),
      tissue_groups_short = recode(Tissue, !!!tissue_name_dict),
      tissue_groups_short = factor(
        tissue_groups_short,
        levels = c(setdiff(unique(tissue_groups_short), "Large int"), "Large int")
      )
    ) |>
    # Predictions are identical for all source samples in a tissue/age/sex
    # cohort. Keep one plotted estimate and retain the cohort size explicitly.
    group_by(Tissue, tissue_groups_short, cell_type_unified_ensemble, Age, Sex) |>
    summarise(
      n_source_rows = n(),
      n_samples = n_distinct(sample_id),
      proportion_mean = first(proportion_mean),
      proportion_lower = first(proportion_lower),
      proportion_upper = first(proportion_upper),
      .groups = "drop"
    )

  Plot__line_chart_plasma <-
    plasma_plot_data |>
    ggplot() +
    geom_hline(yintercept = 0, linewidth = 0.1) +
    geom_point(aes(x = Age, y = proportion_mean), shape = NA) +
    geom_line(aes(x = as.numeric(Age), y = proportion_mean, colour = Sex, fill = Sex),
              show.legend = FALSE) +
    geom_ribbon(aes(x = as.numeric(Age), colour = Sex, fill = Sex,
                    ymin = proportion_lower, ymax = proportion_upper),
                colour = NA, alpha = 0.1) +
    scale_colour_manual(values = c("#D7907B", "#0A014F"),
                        aesthetics = c("colour", "fill"),
                        guide = guide_legend(override.aes = list(alpha = 1))) +
    ggnewscale::new_scale_colour() +
    scale_colour_brewer(palette = "Set2", name = "Comparison") +
    scale_y_continuous(
      trans = asin_sqrt_trans,
      limits = c(0, NA),
      expand = expansion(mult = c(0, 0.05)),
      labels = function(x) gsub("^0\\.", ".", format(x, nsmall = 1))
    ) +
    labs(x = "Age decade", y = "Proportion Mean (arcsin-sqrt scale)") +
    facet_wrap(~ tissue_groups_short, ncol = 4, scales = 'free_y') +
    theme_multipanel +
    theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1))

  save_fig5_proportion_plot(Plot__line_chart_plasma, path_line)
}

print(Plot__line_chart_plasma)
```

# Final assembly

Layout (ported from `report/Fig_files/Fig4.R`): panel **A** spans the top;
below, the left column stacks **B** over **D** and the right column stacks the
**C** heatmap over a spacer.

```{r fig.width=10, fig.height=12, label="fig4-assembly"}
# --- Panel A (compact) ------------------------------------------------------
Plot__A <-
  Plot__age_sex_ethnicity_tissue_bar +
  theme(
    plot.title = element_blank(),
    legend.position = "bottom", legend.direction = "horizontal", legend.text.position = "bottom",
    strip.text.x.top = element_text(margin = margin(0, 0, 0, 0, "pt")),
    strip.placement = "inside",
    panel.spacing = unit(0, "pt"),
    plot.margin = margin(0, 0, 0, 0, "pt")
  )

# --- Panel B (top-4 by FDR) --------------------------------------------------
Plot__boxplot_2p <-
  Plot__boxplot_fig5b +
  guides(fill = "none") +
  theme(
    legend.position = "none", legend.direction = "horizontal", legend.text.position = "bottom",
    panel.spacing = unit(0, "pt"), plot.margin = margin(0, 0, 0, 0, "pt")
  )

# --- Panel D (thin lines, y starting at 0) ----------------------------------
Plot__line <- Plot__line_chart_plasma
Plot__line$layers <- lapply(Plot__line$layers, function(layer) {
  if (inherits(layer$geom, "GeomLine")) layer$aes_params$linewidth <- 0.25
  layer
})
Plot__line <-
  Plot__line +
  scale_y_continuous(
    trans = asin_sqrt_trans,
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.05)),
    labels = function(x) gsub("^0\\.", ".", format(x, nsmall = 1))
  ) +
  theme(
    plot.title = element_blank(),
    axis.text.x.bottom = element_text(angle = 0),
    legend.position = "none", legend.direction = "horizontal", legend.text.position = "bottom",
    strip.text.x.top = element_text(margin = margin(0, 0, 0, 0, "pt")),
    strip.placement = "inside",
    panel.spacing = unit(0, "pt"), plot.margin = margin(0, 0, 0, 0, "pt")
  )

# --- Panel C (heatmap -> ggplot); apply an FDR floor so point sizes are finite
heat_final <- {
  d <- Plot__heatmap_male_vs_female@data
  min_fdr <- min(d$c_FDR[d$c_FDR > 0], na.rm = TRUE)
  d$c_FDR[d$c_FDR == 0] <- min_fdr
  d %>%
    mutate(size = -log10(c_FDR)) %>%
    heatmap(
      .row = tissue_short, .column = cell_type_abbrev, .value = c_effect,
      name = "Effect\nsize",
      palette_value = circlize::colorRamp2(seq(-2, 2, length.out = 7), scico::scico(7, palette = "vik")),
      row_title = "Tissue", column_title = "Cell type"
    ) |>
    layer_point(c_FDR < 0.05, .size = size) |>
    annotation_bar(NCellsCellType, annotation_label = "log(Cells)") |>
    annotation_bar(NSamplesCellType, annotation_label = "log(Samples)") |>
    annotation_bar(NCellsTissue, annotation_label = "log(Cells)", annotation_name_rot = 45) |>
    annotation_bar(NSamplesTissue, annotation_label = "log(Samples)", annotation_name_rot = 45)
}

ht_temp <- heat_final %>% as_ComplexHeatmap()
ht_temp@heatmap_param$show_heatmap_legend <- FALSE
ht_temp@column_names_param$gp$fontsize <- 13
ht_temp@row_names_param$gp$fontsize <- 13

heatmap_ggplot <-
  as.ggplot(ht_temp) +
  theme(
    axis.line = element_blank(), axis.text = element_blank(),
    axis.title = element_blank(), axis.ticks = element_blank(),
    plot.margin = margin(0, 0, 0, 0, "pt")
  )

# --- Rows --------------------------------------------------------------------
row2_left <-
  (Plot__boxplot_2p / Plot__line) +
  plot_layout(heights = c(1, 3)) &
  theme(panel.spacing = unit(0, "pt"), plot.margin = margin(0, 0, 0, 0, "pt"),
        legend.position = "bottom", legend.direction = "horizontal", legend.text.position = "bottom")

row2_right <-
  (heatmap_ggplot / plot_spacer()) +
  plot_layout(heights = c(6, 1)) &
  theme(panel.spacing = unit(0, "pt"), plot.margin = margin(0, 0, 0, 0, "pt"),
        legend.position = "bottom", legend.direction = "horizontal", legend.text.position = "bottom")

row2 <-
  (row2_left | row2_right) +
  plot_layout(widths = c(2, 2)) &
  theme(panel.spacing = unit(0, "pt"), plot.margin = margin(0, 0, 0, 0, "pt"),
        legend.position = "bottom", legend.direction = "horizontal", legend.text.position = "bottom")

# --- Final figure ------------------------------------------------------------
fig5 <-
  (Plot__A / row2) +
  plot_layout(guides = 'collect', heights = c(1, 6)) &
  theme(
    panel.spacing = unit(0, "pt"), plot.margin = margin(0, 0, 0, 0, "pt"),
    legend.position = "bottom", legend.direction = "horizontal", legend.text.position = "bottom",
    legend.key.size = unit(5, 'mm'),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 6)
  )

ggsave(
  plot = fig5,
  filename = file.path(path_figure_ready("fig5"), "fig5.pdf"),
  units = "mm", width = 183, height = 180, dpi = 500, scale = 1
)

fig5
```

# Supplementary tables

Export plot statistics from ggplot / heatmap `@data` slots for all Figure 5 panels.

```{r export-supplementary-tables}
message(
  "Panel caches updated. Run `Rscript report/build_supplementary_workbooks.R` ",
  "to rebuild the split-tab workbooks with separate plot-data, summary and test sheets."
)
```

# Save panels

```{r save-all-plots, eval=FALSE}
# Persist all panel objects for downstream figure assembly / re-use.
save(
  list = ls(pattern = "Plot__"),
  file = file.path(path_figure_ready("fig5"), "Fig5_plot_data_all.RData")
)

sessionInfo()
```
