---
title: "Figure 5 supplementary plots"
date: today
date-format: "DD MMM YYYY"
format:
  html:
    theme: minty
    toc: true
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
---

# Purpose

This report expands panels that the main figure truncates for display:

1. donor coverage across all mapped tissue groups;
2. every body-level sex contrast across age decades;
3. the full tissue-by-cell-type sex-effect heatmap;
4. sex boxplots for **all** tested cell types (not only the six highlighted in
   Fig. 5b); and
5. plasma-cell sex trajectories for **all** tissues with plasma predictions
   (no coverage filter; Fig. 5d still uses a stricter subset).

On render, each ggplot object is written to
`vignettes/supplementary_plot_objects/`.

```{r setup}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(here)
library(forcats)
library(stringr)
library(knitr)
library(ggnewscale)
library(patchwork)

source(file.path("..", "R", "paths.R"))
workbook <- path_source_tables("Supplementary_File_Figure_5.xlsx")
if (!file.exists(workbook)) stop("Figure 5 supplementary file not found: ", workbook)

plot_object_dir <- file.path(publication_vignette_root(), "supplementary_plot_objects")
dir.create(plot_object_dir, recursive = TRUE, showWarnings = FALSE)

read_panel_sheet <- function(sheet, path = workbook) {
  if (!sheet %in% excel_sheets(path)) stop("Missing worksheet: ", sheet, " in ", path)
  # Submission workbooks store a caption in row 1 and headers in row 2.
  read_xlsx(path, sheet = sheet, skip = 1, guess_max = 1000000L)
}

save_plot_object <- function(plot, filename) {
  path <- file.path(plot_object_dir, filename)
  saveRDS(plot, path, compress = "xz")
  message("Saved ggplot object: ", path)
  invisible(path)
}

clean_label <- function(x) {
  x <- gsub("_", " ", x, fixed = TRUE)
  tools::toTitleCase(x)
}

# Same asin-sqrt axis transform as Fig2 supplementary proportion plots:
# keep raw proportions in the data; transform only the scale.
asin_sqrt_trans <- scales::trans_new(
  name = "asin_sqrt",
  transform = function(x) asin(sqrt(pmax(pmin(x, 1), 0))),
  inverse = function(y) sin(y)^2
)

# Match Fig5.qmd theme_multipanel exactly for boxplots and line charts.
theme_multipanel <-
  theme_bw() +
  theme(
    panel.border = element_blank(),
    axis.line = element_line(size = 0.1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    strip.background = element_blank(),
    axis.title.y = element_text(margin = margin(0, 0, 0, 0), size = 7),
    axis.title.x = element_text(margin = margin(0, 0, 0, 0), size = 7),
    panel.spacing.x = unit(0.1, "lines"),
    axis.text.x = element_text(size = 6),
    axis.text.y = element_text(size = 6),
    strip.text.x = element_text(size = 7),
    strip.text.y = element_text(size = 7),
    legend.key.size = unit(5, "mm"),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 6),
    strip.clip = "off",
    plot.title = element_text(size = 7),
    axis.line.x = element_line(size = 0.2),
    axis.line.y = element_line(size = 0.2),
    axis.ticks.x = element_line(size = 0.2),
    axis.ticks.y = element_line(size = 0.2)
  )

# Lighter theme retained only for coverage / heatmap summary panels.
theme_supplementary <- theme_bw(base_size = 8.5) +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_blank(),
    legend.position = "bottom",
    plot.title.position = "plot"
  )

# Same single-line abbreviations as Fig5.qmd boxplot facets.
cell_type_abbrev_dict <- c(
  "b memory" = "B Memory", "b naive" = "B Naive",
  "cd14 mono" = "CD14 Mono", "cd16 mono" = "CD16 Mono",
  "cd4 fh em" = "CD4 FH EM", "cd4 naive" = "CD4 Naive", "cd4 tcm" = "CD4 TCM",
  "cd4 th1 em" = "CD4 Th1 EM", "cd4 th1/th17 em" = "CD4 Th1/Th17 EM",
  "cd4 th17 em" = "CD4 Th17 EM", "cd4 th2 em" = "CD4 Th2 EM", "treg" = "Treg",
  "cd8 naive" = "CD8 Naive", "cd8 tcm" = "CD8 TCM", "cd8 tem" = "CD8 TEM",
  "cdc" = "CDC", "granulocyte" = "Gran", "ilc" = "ILC", "macrophage" = "Mac",
  "mait" = "MAIT", "mast" = "Mast", "nk" = "NK", "nkt" = "NKT", "pdc" = "PDC",
  "tgd" = "TGD", "plasma" = "Plasma"
)

tissue_name_dict <- c(
  "blood" = "Blood",
  "sensory-related structures" = "Sensory",
  "cerebral lobes and cortical areas" = "Cerebral cortex",
  "limbic and basal systems" = "Limbic/basal",
  "liver" = "Liver",
  "bone marrow" = "Bone marrow",
  "large intestine" = "Large int",
  "lymphatic system" = "Lymphatic",
  "small intestine" = "Small int",
  "respiratory system" = "Respiratory",
  "renal system" = "Renal",
  "brainstem and cerebellar structures" = "Hindbrain",
  "cardiovascular system" = "Cardiovascular",
  "general brain and major structures" = "Brain",
  "vasculature" = "Vasculature",
  "ovary" = "Ovary",
  "integumentary system (skin)" = "Skin",
  "nasal, oral, and pharyngeal regions" = "Pharyngeal",
  "muscular system (skeletal muscles)" = "Skeletal muscle",
  "prostate" = "Prostate",
  "breast" = "Breast",
  "digestive system (general)" = "Digestive",
  "oesophagus" = "Oesophagus",
  "female reproductive system" = "Female repro",
  "spleen" = "Spleen",
  "peritoneal and abdominal cavity structures" = "Peritoneal",
  "thymus" = "Thymus",
  "adipose tissue" = "Adipose",
  "trachea" = "Trachea",
  "endocrine system" = "Endocrine",
  "pancreas" = "Pancreas",
  "miscellaneous glands" = "Glands",
  "stomach" = "Stomach",
  "gallbladder" = "Gallbladder"
)

colmap_sex <- c("female" = "#D7907B", "male" = "#0A014F")

resolve_fig5_cache <- function(filename) {
  candidates <- c(
    path_intermediate("fig5", filename),
    {
      arch <- path_age_clock_archive("data", filename)
      if (is.null(arch)) character() else arch
    },
    {
      arch <- path_age_clock_archive("data", "age_fig", filename)
      if (is.null(arch)) character() else arch
    }
  )
  hit <- candidates[nzchar(candidates) & file.exists(candidates)][1]
  if (is.na(hit)) {
    stop(
      "Missing Fig5 cache '", filename, "'. Expected under path_intermediate('fig5') ",
      "or AGE_CLOCK_ARCHIVE/data/.",
      call. = FALSE
    )
  }
  hit
}

format_fdr <- function(x) {
  case_when(
    is.na(x) ~ "NA",
    x < 0.001 ~ gsub("^0\\.", ".", formatC(x, format = "e", digits = 2)),
    TRUE ~ gsub("^0\\.", ".", formatC(x, format = "f", digits = 3))
  )
}
```

```{r read-source-data}
donor_counts <- read_panel_sheet("Fig5_A_donor_counts")
sex_tests <- read_panel_sheet("Fig5_B_sex_tests")
tissue_effects <- read_panel_sheet("Fig5_C_tissue_effects")

global_sex_tests_l3 <- readRDS(
  resolve_fig5_cache("Fig4_global_sex_tests_compact.rds")
) |>
  mutate(significant = !is.na(c_FDR) & c_FDR < 0.05)

# Best age-band contrast per cell type (same selection logic as the main panel).
best_global_sex <- global_sex_tests_l3 |>
  group_by(cell_type_unified_ensemble) |>
  arrange(c_FDR, desc(abs(c_effect)), .by_group = TRUE) |>
  slice_head(n = 1) |>
  ungroup() |>
  mutate(
    selection_direction = case_when(
      c_effect > 0 ~ "positive",
      c_effect < 0 ~ "negative",
      TRUE ~ "zero"
    )
  )
```

# Complete age-by-sex cohort coverage

```{r fig5-full-cohort-coverage, fig.width=8, fig.height=12}
coverage_data <- donor_counts |>
  transmute(
    age_groups = .data[["Age stratum"]],
    sex = .data[["Sex"]],
    tissue_label = coalesce(
      na_if(.data[["Figure tissue label"]], ""),
      .data[["Tissue group"]]
    ),
    donor_count = .data[["Number of unique donors"]],
    displayed_main = .data[["Displayed in Fig. 5a"]] == "Yes"
  ) |>
  filter(!is.na(tissue_label), tissue_label != "Unmapped tissue") |>
  mutate(
    sex = factor(
      str_to_lower(sex),
      levels = c("female", "male"),
      labels = c("Female", "Male")
    ),
    age_groups = factor(age_groups, levels = c("Young", "Mid", "Old"))
  )

tissue_order <- coverage_data |>
  group_by(tissue_label) |>
  summarise(total = sum(donor_count), .groups = "drop") |>
  arrange(total) |>
  pull(tissue_label)

coverage_data <- coverage_data |>
  mutate(tissue_label = factor(tissue_label, levels = tissue_order))

unmapped_n <- donor_counts |>
  filter(.data[["Tissue group"]] == "Unmapped tissue") |>
  summarise(n = sum(.data[["Number of unique donors"]], na.rm = TRUE)) |>
  pull(n)
if (length(unmapped_n) == 0 || is.na(unmapped_n)) unmapped_n <- 0

Plot_SFig5_full_cohort_coverage <-
  ggplot(coverage_data, aes(x = age_groups, y = tissue_label, fill = donor_count)) +
  geom_tile(colour = "white", linewidth = 0.25) +
  geom_text(aes(label = ifelse(donor_count == 0, "", donor_count)), size = 2.2) +
  facet_grid(. ~ sex) +
  scale_fill_viridis_c(trans = "sqrt", option = "C", name = "Unique donors") +
  labs(
    x = "Age group",
    y = NULL,
    title = "Donor coverage across all mapped tissues, age groups and sexes",
    subtitle = sprintf(
      "All mapped tissues are shown; %d donor records are unmapped",
      unmapped_n
    )
  ) +
  theme_supplementary +
  theme(panel.grid.major = element_blank())

Plot_SFig5_full_cohort_coverage
save_plot_object(
  Plot_SFig5_full_cohort_coverage,
  "Fig5_supp_full_cohort_coverage.rds"
)
```

# Complete body-level sex effects across age decades

```{r fig5-all-global-sex-effects, fig.width=9, fig.height=8}
global_effect_data <- global_sex_tests_l3 |>
  filter(!is.na(cell_type_unified_ensemble), !is.na(Age)) |>
  mutate(
    cell_label = clean_label(cell_type_unified_ensemble),
    age_decade = factor(as.character(Age), levels = as.character(1:10)),
    supported = significant
  )

cell_order <- global_effect_data |>
  group_by(cell_label) |>
  summarise(mean_effect = mean(c_effect, na.rm = TRUE), .groups = "drop") |>
  arrange(mean_effect) |>
  pull(cell_label)

global_effect_data <- global_effect_data |>
  mutate(cell_label = factor(cell_label, levels = cell_order))

global_effect_limit <- max(abs(global_effect_data$c_effect), na.rm = TRUE)

Plot_SFig5_all_global_sex_effects <-
  ggplot(global_effect_data, aes(x = age_decade, y = cell_label, fill = c_effect)) +
  geom_tile(colour = "white", linewidth = 0.2) +
  geom_point(
    data = filter(global_effect_data, supported),
    shape = 8, size = 1.4, colour = "black"
  ) +
  scale_fill_gradient2(
    low = "#D7907B", mid = "white", high = "#0A014F", midpoint = 0,
    limits = c(-global_effect_limit, global_effect_limit),
    name = "Male − female\neffect"
  ) +
  labs(
    x = "Age decade",
    y = NULL,
    title = "Complete body-level sex effects across age decades",
    subtitle = "The fifth decade is the model reference; stars denote FDR < 0.05"
  ) +
  theme_supplementary +
  theme(panel.grid.major = element_blank())

Plot_SFig5_all_global_sex_effects
save_plot_object(
  Plot_SFig5_all_global_sex_effects,
  "Fig5_supp_all_global_sex_effects.rds"
)
```

# Full-resolution tissue-by-cell-type sex effects

```{r fig5-full-tissue-sex-heatmap, fig.width=13, fig.height=10}
tissue_effect_data <- tissue_effects |>
  transmute(
    tissue_label = .data[["Tissue group"]],
    cell_label = .data[["Immune cell type"]],
    c_effect = .data[[
      "Male-minus-female posterior mean effect (unconstrained sccomp composition scale)"
    ]],
    c_FDR = .data[["Bayesian FDR"]],
    supported = .data[["FDR < 0.05"]] == "Yes"
  ) |>
  filter(!is.na(tissue_label), !is.na(cell_label), is.finite(c_effect))

tissue_order <- tissue_effect_data |>
  group_by(tissue_label) |>
  summarise(mean_effect = mean(c_effect, na.rm = TRUE), .groups = "drop") |>
  arrange(mean_effect) |>
  pull(tissue_label)
cell_order <- tissue_effect_data |>
  group_by(cell_label) |>
  summarise(mean_effect = mean(c_effect, na.rm = TRUE), .groups = "drop") |>
  arrange(mean_effect) |>
  pull(cell_label)

tissue_effect_data <- tissue_effect_data |>
  mutate(
    tissue_label = factor(tissue_label, levels = tissue_order),
    cell_label = factor(cell_label, levels = cell_order)
  )

tissue_effect_limit <- max(abs(tissue_effect_data$c_effect), na.rm = TRUE)

Plot_SFig5_full_tissue_sex_heatmap <-
  ggplot(tissue_effect_data, aes(x = cell_label, y = tissue_label, fill = c_effect)) +
  geom_tile(colour = "white", linewidth = 0.15) +
  geom_point(
    data = filter(tissue_effect_data, supported),
    size = 0.8, colour = "black"
  ) +
  scale_fill_gradient2(
    low = "#D7907B", mid = "white", high = "#0A014F", midpoint = 0,
    limits = c(-tissue_effect_limit, tissue_effect_limit),
    name = "Male − female\neffect"
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "Full tissue-resolved landscape of sex effects",
    subtitle = "All estimates are shown; black points denote FDR < 0.05"
  ) +
  theme_supplementary +
  theme(
    panel.grid.major = element_blank(),
    axis.text.x = element_text(angle = 55, hjust = 1, size = 6)
  )

Plot_SFig5_full_tissue_sex_heatmap
save_plot_object(
  Plot_SFig5_full_tissue_sex_heatmap,
  "Fig5_supp_full_tissue_sex_heatmap.rds"
)
```

# Sex boxplots for every tested cell type

Main Fig. 5b shows three male-biased and three female-biased cell types. Here
every cell type with a body-level sex contrast is plotted, using sample-level
sex predictions from the analysis cache (age structure of the cohort retained).

```{r fig5-all-sex-boxplots-data}
sex_pred <- readRDS(
  resolve_fig5_cache(
    "predictions_age_decade_sex___L3___disease_FALSE___immune_only_TRUE.rds"
  )
) |>
  filter(sex %in% c("female", "male")) |>
  transmute(
    sample_id,
    L3 = cell_type_unified_ensemble,
    sex,
    age_decade,
    proportion_mean
  )

build_sex_boxplot_data <- function(direction) {
  selected <- best_global_sex |>
    filter(selection_direction == direction)
  selected <- if (identical(direction, "positive")) {
    arrange(selected, desc(c_effect), c_FDR)
  } else {
    arrange(selected, c_effect, c_FDR)
  }

  cell_types <- selected$cell_type_unified_ensemble
  sex_pred |>
    filter(L3 %in% cell_types) |>
    left_join(
      selected |>
        transmute(L3 = cell_type_unified_ensemble, c_FDR, c_effect),
      by = "L3"
    ) |>
    mutate(
      sex = factor(sex, levels = c("female", "male")),
      version = "prediction",
      L3_abbrev_base = recode(L3, !!!cell_type_abbrev_dict),
      L3_abbrev = paste0(L3_abbrev_base, "\n(FDR: ", format_fdr(c_FDR), ")"),
      L3 = factor(L3, levels = cell_types)
    ) |>
    arrange(L3) |>
    mutate(L3_abbrev = factor(L3_abbrev, levels = unique(as.character(L3_abbrev))))
}

build_sex_boxplot <- function(df) {
  # Match Fig2 supplementary proportion styling: raw fractions in data,
  # asin-sqrt applied via the y scale (not in the aesthetic mapping).
  # Without sex_adjusted draws we only have prediction rows, so the jitter
  # layer is omitted when absent.
  has_adjusted <- "version" %in% names(df) && any(df$version == "adjusted")
  if (!"version" %in% names(df)) {
    df <- mutate(df, version = "prediction")
  }

  p <- ggplot(df, aes(x = sex, y = proportion_mean))
  if (has_adjusted) {
    p <- p +
      geom_jitter(
        data = df %>% dplyr::filter(version == "adjusted"),
        width = 0.15, colour = "grey50", shape = ".", size = 0.6, alpha = 0.9
      )
  }
  p +
    geom_boxplot(
      aes(fill = sex),
      data = df %>% dplyr::filter(version == "prediction"),
      outlier.shape = NA, alpha = 0.3, size = 0.25
    ) +
    scale_fill_manual(values = colmap_sex) +
    scale_y_continuous(
      trans = asin_sqrt_trans,
      expand = expansion(mult = c(0.02, 0.02)),
      labels = function(x) gsub("^0\\.", ".", format(x, nsmall = 1))
    ) +
    facet_wrap(~ L3_abbrev, ncol = 5, scales = "fixed") +
    labs(title = "", x = "Sex", y = "Proportion Mean (arcsin-sqrt scale)") +
    theme_multipanel +
    theme(
      strip.text.x.top = element_text(margin = margin(0, 0, 0, 0, "pt")),
      strip.placement = "inside"
    )
}

df_sex_pos <- build_sex_boxplot_data("positive")
df_sex_neg <- build_sex_boxplot_data("negative")
```

```{r fig5-all-sex-boxplots-pos, fig.width=14, fig.height=8}
Plot_SFig5_all_boxplot_male_biased <- build_sex_boxplot(df_sex_pos)
Plot_SFig5_all_boxplot_male_biased
save_plot_object(
  Plot_SFig5_all_boxplot_male_biased,
  "Fig5_supp_all_boxplot_male_biased.rds"
)
```

```{r fig5-all-sex-boxplots-neg, fig.width=14, fig.height=8}
Plot_SFig5_all_boxplot_female_biased <- build_sex_boxplot(df_sex_neg)
Plot_SFig5_all_boxplot_female_biased
save_plot_object(
  Plot_SFig5_all_boxplot_female_biased,
  "Fig5_supp_all_boxplot_female_biased.rds"
)
```

```{r fig5-all-sex-boxplots-combined, fig.width=14, fig.height=16}
# Single SI figure: A = male-biased, B = female-biased.
Plot_SFig5_all_boxplot <-
  (Plot_SFig5_all_boxplot_male_biased / Plot_SFig5_all_boxplot_female_biased) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(
    plot.tag = element_text(size = 9, face = "bold"),
    plot.margin = margin(2, 2, 2, 2, "mm"),
    legend.position = "bottom"
  )
Plot_SFig5_all_boxplot
save_plot_object(Plot_SFig5_all_boxplot, "Fig5_supp_all_boxplot.rds")
ggsave(
  file.path(plot_object_dir, "Fig5_supp_all_boxplot.png"),
  Plot_SFig5_all_boxplot,
  width = 183,
  height = 240,
  units = "mm",
  dpi = 500,
  bg = "white"
)
```

# Plasma sex trajectories for every tissue

No coverage filter: every tissue with plasma-cell predictions is shown (female/male
only; unknown sex dropped). Main Fig. 5d still uses a stricter subset. Styling
matches Fig2 supplementary plasma trajectories (raw proportions, asin-sqrt y scale).

```{r fig5-all-plasma-line, fig.width=15, fig.height=14}
sex_tissue_pred <- readRDS(
  resolve_fig5_cache(
    "predictions_age_decade_sex_tissue___L3___disease_FALSE___immune_only_TRUE.rds"
  )
)

plasma_line_data <- sex_tissue_pred |>
  filter(
    cell_type_unified_ensemble == "plasma",
    sex %in% c("female", "male"),
    !is.na(age_decade),
    !is.na(tissue_groups)
  ) |>
  group_by(tissue_groups, cell_type_unified_ensemble, age_decade, sex) |>
  summarise(
    n_samples = n_distinct(sample_id),
    proportion_mean = first(proportion_mean),
    proportion_lower = first(proportion_lower),
    proportion_upper = first(proportion_upper),
    .groups = "drop"
  ) |>
  mutate(
    Age = factor(as.character(age_decade), levels = as.character(1:10)),
    Sex = factor(sex, levels = c("female", "male")),
    tissue_groups_short = {
      out <- unname(tissue_name_dict[tissue_groups])
      out <- ifelse(
        is.na(out),
        tools::toTitleCase(gsub("_", " ", tissue_groups, fixed = TRUE)),
        out
      )
      factor(out, levels = c(setdiff(sort(unique(out)), "Large int"), "Large int"))
    }
  )

Plot_SFig5_all_plasma_line <-
  plasma_line_data |>
  ggplot() +
  geom_hline(yintercept = 0, lwd = 0.1) +
  geom_point(aes(x = Age, y = proportion_mean), shape = NA) +
  geom_line(
    aes(x = as.numeric(Age), y = proportion_mean, colour = Sex, fill = Sex),
    show.legend = FALSE
  ) +
  geom_ribbon(
    aes(
      x = as.numeric(Age),
      colour = Sex,
      fill = Sex,
      ymin = proportion_lower,
      ymax = proportion_upper
    ),
    colour = NA,
    alpha = 0.1
  ) +
  scale_colour_manual(
    values = c("#D7907B", "#0A014F"),
    aesthetics = c("colour", "fill"),
    guide = guide_legend(override.aes = list(alpha = 1))
  ) +
  ggnewscale::new_scale_colour() +
  scale_colour_brewer(palette = "Set2", name = "Comparison") +
  scale_y_continuous(
    trans = asin_sqrt_trans,
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.05)),
    labels = function(x) gsub("^0\\.", ".", format(x, nsmall = 1))
  ) +
  labs(x = "Age decade", y = "Proportion Mean (arcsin-sqrt scale)") +
  facet_wrap(~ tissue_groups_short, ncol = 5, scales = "free_y") +
  theme_multipanel +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 0, hjust = 0.5)
  )

Plot_SFig5_all_plasma_line
save_plot_object(
  Plot_SFig5_all_plasma_line,
  "Fig5_supp_all_plasma_line.rds"
)
```

# Output objects

```{r output-manifest}
data.frame(
  object = c(
    "Plot_SFig5_full_cohort_coverage",
    "Plot_SFig5_all_global_sex_effects",
    "Plot_SFig5_full_tissue_sex_heatmap",
    "Plot_SFig5_all_boxplot_male_biased",
    "Plot_SFig5_all_boxplot_female_biased",
    "Plot_SFig5_all_boxplot",
    "Plot_SFig5_all_plasma_line"
  ),
  rds_file = c(
    "Fig5_supp_full_cohort_coverage.rds",
    "Fig5_supp_all_global_sex_effects.rds",
    "Fig5_supp_full_tissue_sex_heatmap.rds",
    "Fig5_supp_all_boxplot_male_biased.rds",
    "Fig5_supp_all_boxplot_female_biased.rds",
    "Fig5_supp_all_boxplot.rds",
    "Fig5_supp_all_plasma_line.rds"
  )
)
```
