---
title: "Single-cell integration and QC pseudobulk generation for plasma niche analyses"
date: today
date-format: "DD MMM YYYY"
title-block-banner: true
author:
  - name: Chen Zhan
    orcid: 0000-0002-4794-8339
    email: chen.zhan@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
format:
  html:
    theme: minty
    toc: true
    toc-location: left
    toc-title: "Menu"
    code-fold: true
    code-line-numbers: true
    embed-resources: true
execute:
  warning: false
  message: false
params:
  tissue_systems: "respiratory_system,large_intestine"
  cell_compartments: "plasma,stromal,epithelial"
  min_cells_per_pseudobulk: 10
  cores: 4
  output_dir: ""
abstract: |
  Integrates healthy single-cell data with Seurat and Harmony for UMAP
  visualisation, then generates separate, quality-controlled pseudobulk
  expression objects for plasma, stromal and epithelial cells from the
  respiratory system and large intestine. These objects are inputs for plasma
  niche transcriptomics and pathway analyses.
---

# Purpose

This report merges the shared data-preparation logic from the original
multi-tissue plasma-niche and large-intestine reports. It performs one consistent
workflow and writes one pseudobulk `SummarizedExperiment` for each selected
tissue-system and cell-compartment combination. It also creates an integrated
Seurat object and single-cell UMAPs for checking tissue, compartment and sample
structure before aggregation.

With the default parameters, the report generates one integrated Seurat object
and six pseudobulk files:

| Tissue system | Separate cell-compartment outputs |
|---|---|
| Respiratory system | plasma, stromal, epithelial |
| Large intestine | plasma, stromal, epithelial |

The HTML report includes single-cell Seurat UMAPs and interactive
sample-retention and library-quality plots. The generated pseudobulk RDS files
are the analysis inputs for **plasma niche
transcriptomics and pathway analyses**; this report does not perform the
downstream differential-expression or pathway modelling.

# Configuration

Render the default six-output workflow from the repository root:

```bash
quarto render vignettes/reports/plasma_niche_pseudobulk_qc.qmd
```

Select a subset with Quarto parameters, for example:

```bash
quarto render vignettes/reports/plasma_niche_pseudobulk_qc.qmd \
  -P tissue_systems:respiratory_system \
  -P cell_compartments:plasma,stromal
```

Set `CELLNEXUS_CACHE_DIR` to a writable cellNexus cache. Set
`PLASMA_NICHE_OUTPUT_DIR` to override the default output directory
`vignettes/data/processed/plasma_niche/`.

```{r setup}
source(file.path("..", "R", "paths.R"))

Sys.setenv(IMMUNE_HEALTHY_BODY_MAP_ROOT = publication_repo_root())
if (!nzchar(Sys.getenv("DISEASE_GROUPING_CSV", unset = ""))) {
  Sys.setenv(
    DISEASE_GROUPING_CSV = path_source_tables(
      "disease_data_grouped_further.csv"
    )
  )
}

suppressPackageStartupMessages({
  library(BiocParallel)
  library(cellNexus)
  library(edgeR)
  library(ggplot2)
  library(plotly)
  library(S4Vectors)
  library(Seurat)
  library(SummarizedExperiment)
  library(harmony)
  library(scuttle)
  library(tidyverse)
})

source(path_edit_covariates())

split_parameter <- function(value) {
  value |>
    str_split(",", simplify = TRUE) |>
    as.character() |>
    str_trim() |>
    discard(~ !nzchar(.x)) |>
    unique()
}

tissue_system_lookup <- c(
  respiratory_system = "respiratory system",
  large_intestine = "large intestine"
)

feature_count_threshold <- c(
  respiratory_system = 8000L,
  large_intestine = 5000L
)

valid_compartments <- c("plasma", "stromal", "epithelial")
selected_systems <- split_parameter(params$tissue_systems)
selected_compartments <- split_parameter(params$cell_compartments)

unknown_systems <- setdiff(selected_systems, names(tissue_system_lookup))
unknown_compartments <- setdiff(selected_compartments, valid_compartments)
if (length(unknown_systems)) {
  stop("Unknown tissue_systems value(s): ", paste(unknown_systems, collapse = ", "))
}
if (length(unknown_compartments)) {
  stop(
    "Unknown cell_compartments value(s): ",
    paste(unknown_compartments, collapse = ", ")
  )
}

cores <- max(1L, as.integer(params$cores))
min_cells <- max(1L, as.integer(params$min_cells_per_pseudobulk))

output_dir <- params$output_dir
if (!nzchar(output_dir)) {
  output_dir <- Sys.getenv(
    "PLASMA_NICHE_OUTPUT_DIR",
    unset = path_processed("plasma_niche")
  )
}
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
output_dir <- normalizePath(output_dir, winslash = "/", mustWork = TRUE)

cellnexus_cache <- Sys.getenv(
  "CELLNEXUS_CACHE_DIR",
  unset = path_processed("cellNexus_cache")
)
dir.create(cellnexus_cache, recursive = TRUE, showWarnings = FALSE)

bpparam <- if (cores > 1L) {
  BiocParallel::MulticoreParam(cores)
} else {
  BiocParallel::SerialParam()
}

tibble(
  setting = c(
    "Tissue systems",
    "Cell compartments",
    "Minimum contributing cells",
    "Workers",
    "Output directory",
    "cellNexus cache"
  ),
  value = c(
    paste(selected_systems, collapse = ", "),
    paste(selected_compartments, collapse = ", "),
    min_cells,
    cores,
    output_dir,
    cellnexus_cache
  )
) |>
  knitr::kable()
```

# Select healthy single-cell data

The two legacy reports used the same metadata filters and different
feature-count thresholds. This merged report retains the original threshold for
each tissue system: 8,000 detected features for respiratory tissues and 5,000
for large-intestine tissues.

```{r load-and-filter-metadata}
metadata_selected <- get_publication_metadata() |>
  filter(
    disease == "normal",
    empty_droplet %in% FALSE,
    alive %in% TRUE,
    scDblFinder.class != "doublet",
    !is.na(feature_count)
  ) |>
  edit_covariates() |>
  filter(tissue_groups %in% unname(tissue_system_lookup[selected_systems])) |>
  mutate(
    tissue_system = names(tissue_system_lookup)[
      match(tissue_groups, unname(tissue_system_lookup))
    ],
    min_feature_count = unname(feature_count_threshold[tissue_system])
  ) |>
  filter(feature_count >= min_feature_count)

if (!nrow(metadata_selected)) {
  stop("No metadata rows remain after tissue and cell-level QC filters.")
}

count_samples <- function(data, step, step_label) {
  data |>
    filter(cell_type_unified_ensemble %in% selected_compartments) |>
    distinct(tissue_system, cell_type_unified_ensemble, sample_id) |>
    count(
      tissue_system,
      cell_compartment = cell_type_unified_ensemble,
      name = "n_samples"
    ) |>
    complete(
      tissue_system = selected_systems,
      cell_compartment = selected_compartments,
      fill = list(n_samples = 0L)
    ) |>
    mutate(step = step, step_label = step_label, .before = 1)
}

sample_tracking <- list(
  count_samples(
    metadata_selected,
    "1_selected_metadata",
    "1. Selected metadata"
  )
)

metadata_selected |>
  distinct(tissue_system, tissue_groups, tissue) |>
  count(tissue_system, tissue_groups, name = "n_tissues") |>
  knitr::kable(caption = "Selected tissue annotations")
```

# Dataset-diversity QC

Following the intent of the source reports, collections represented by only one
immune-cell annotation within a selected tissue system are removed. Such
collections have zero immune-composition entropy and are likely to have been
experimentally enriched or depleted in a way that makes them unsuitable for
plasma-niche comparisons.

```{r dataset-diversity-qc}
dataset_entropy <- metadata_selected |>
  filter(is_immune) |>
  count(
    collection_id,
    dataset_id,
    tissue_system,
    cell_type_unified_ensemble,
    name = "n_cells"
  ) |>
  group_by(collection_id, dataset_id, tissue_system) |>
  mutate(
    proportion = n_cells / sum(n_cells),
    entropy = -sum(proportion * log(proportion))
  ) |>
  summarise(
    n_immune_annotations = n_distinct(cell_type_unified_ensemble),
    n_immune_cells = sum(n_cells),
    entropy = first(entropy),
    .groups = "drop"
  )

low_diversity_collections <- dataset_entropy |>
  filter(n_immune_annotations < 2L | entropy <= sqrt(.Machine$double.eps)) |>
  distinct(collection_id) |>
  pull(collection_id)

metadata_qc <- metadata_selected |>
  filter(!collection_id %in% low_diversity_collections)

sample_tracking[[2]] <- count_samples(
  metadata_qc,
  "2_dataset_diversity_qc",
  "2. Dataset-diversity QC"
)

dataset_entropy |>
  arrange(tissue_system, entropy) |>
  knitr::kable(
    caption = "Immune-composition entropy by dataset before diversity QC",
    digits = 3
  )
```

# Plasma proportion and target compartments

Plasma abundance is calculated against all QC-passing cells from the same sample
before retaining plasma, stromal and epithelial cells. It is stored as a sample
covariate in every resulting pseudobulk object for downstream niche analyses.

```{r prepare-target-metadata}
sample_composition <- metadata_qc |>
  count(sample_id, tissue_system, name = "total_n") |>
  left_join(
    metadata_qc |>
      filter(cell_type_unified_ensemble == "plasma") |>
      count(sample_id, tissue_system, name = "plasma_n"),
    by = c("sample_id", "tissue_system")
  ) |>
  mutate(
    plasma_n = replace_na(plasma_n, 0L),
    prop_plasma_observed = plasma_n / total_n,
    prop_plasma_observed_arcsin_sqrt = asin(sqrt(prop_plasma_observed))
  )

target_metadata <- metadata_qc |>
  filter(cell_type_unified_ensemble %in% selected_compartments) |>
  left_join(sample_composition, by = c("sample_id", "tissue_system"))

if (!nrow(target_metadata)) {
  stop("No plasma, stromal or epithelial cells remain after metadata QC.")
}

target_metadata |>
  distinct(tissue_system, cell_type_unified_ensemble, sample_id) |>
  count(tissue_system, cell_type_unified_ensemble, name = "n_samples") |>
  knitr::kable(caption = "Samples available before single-cell retrieval")
```

# Retrieve single cells

```{r retrieve-single-cells}
sce <- cellNexus::get_single_cell_experiment(
  target_metadata,
  cache_directory = cellnexus_cache
)

sce_metadata <- as.data.frame(SummarizedExperiment::colData(sce))
required_columns <- c(
  "sample_id",
  "cell_type_unified_ensemble",
  "dataset_id",
  "tissue_groups",
  "tissue",
  "tissue_system",
  "prop_plasma_observed",
  "prop_plasma_observed_arcsin_sqrt"
)
missing_columns <- setdiff(required_columns, names(sce_metadata))
if (length(missing_columns)) {
  stop(
    "Retrieved single-cell metadata is missing: ",
    paste(missing_columns, collapse = ", ")
  )
}

sample_tracking[[3]] <- count_samples(
  sce_metadata,
  "3_cells_retrieved",
  "3. Cells retrieved"
)
```

# Seurat integration and single-cell UMAP

The selected cells are normalized in Seurat, represented by 3,000 variable
features and 30 principal components, and integrated across `sample_id` with
Harmony. If only one sample is selected, the report uses the PCA representation
directly. The integrated object is saved alongside the pseudobulk outputs.

```{r integrate-single-cells}
if (ncol(sce) < 3L) {
  stop("At least three cells are required for Seurat integration and UMAP.")
}

counts_assay <- if ("counts" %in% SummarizedExperiment::assayNames(sce)) {
  "counts"
} else {
  SummarizedExperiment::assayNames(sce)[1]
}

seurat_metadata <- sce_metadata |>
  mutate(across(where(is.factor), as.character))
rownames(seurat_metadata) <- colnames(sce)

seurat_object <- Seurat::CreateSeuratObject(
  counts = SummarizedExperiment::assay(sce, counts_assay),
  meta.data = seurat_metadata,
  project = "plasma_niche"
) |>
  Seurat::NormalizeData(verbose = FALSE) |>
  Seurat::FindVariableFeatures(
    nfeatures = min(3000L, nrow(sce)),
    verbose = FALSE
  )

variable_features <- Seurat::VariableFeatures(seurat_object)
n_pcs <- min(30L, length(variable_features) - 1L, ncol(seurat_object) - 1L)
if (n_pcs < 2L) {
  stop("At least three variable features are required for PCA.")
}

seurat_object <- seurat_object |>
  Seurat::ScaleData(features = variable_features, verbose = FALSE) |>
  Seurat::RunPCA(
    features = variable_features,
    npcs = n_pcs,
    verbose = FALSE
  )

if (dplyr::n_distinct(seurat_object$sample_id) > 1L) {
  seurat_object <- harmony::RunHarmony(
    seurat_object,
    group.by.vars = "sample_id",
    reduction.use = "pca",
    dims.use = seq_len(n_pcs),
    reduction.save = "harmony",
    verbose = FALSE
  )
  integration_reduction <- "harmony"
} else {
  integration_reduction <- "pca"
}

seurat_object <- Seurat::RunUMAP(
  seurat_object,
  reduction = integration_reduction,
  dims = seq_len(n_pcs),
  reduction.name = "umap.integrated",
  n.neighbors = min(50L, ncol(seurat_object) - 1L),
  min.dist = 0.3,
  verbose = FALSE
)

saveRDS(
  seurat_object,
  file.path(output_dir, "plasma_niche_single_cell_seurat_integrated.rds")
)
```

```{r seurat-pca-elbow}
Seurat::ElbowPlot(seurat_object, ndims = n_pcs) +
  labs(title = "Principal components used for single-cell integration")
```

```{r seurat-umap-compartment}
Seurat::DimPlot(
  seurat_object,
  reduction = "umap.integrated",
  group.by = "cell_type_unified_ensemble",
  split.by = "tissue_system",
  raster = TRUE,
  pt.size = 0.1
) +
  labs(title = "Integrated single-cell UMAP by compartment and tissue system")
```

```{r seurat-umap-tissue}
Seurat::DimPlot(
  seurat_object,
  reduction = "umap.integrated",
  group.by = "tissue",
  raster = TRUE,
  pt.size = 0.1
) +
  labs(title = "Integrated single-cell UMAP by tissue")
```

```{r seurat-umap-sample}
Seurat::DimPlot(
  seurat_object,
  reduction = "umap.integrated",
  group.by = "sample_id",
  raster = TRUE,
  pt.size = 0.1
) +
  Seurat::NoLegend() +
  labs(title = "Integrated single-cell UMAP by sample")
```

# Aggregate pseudobulk

```{r aggregate-pseudobulk}
aggregate_ids <- S4Vectors::DataFrame(
  sample_id = as.character(sce_metadata$sample_id),
  cell_compartment = as.character(sce_metadata$cell_type_unified_ensemble)
)

pseudobulk_all <- scuttle::aggregateAcrossCells(
  sce,
  ids = aggregate_ids,
  statistics = "sum",
  BPPARAM = bpparam
)

lookup_columns <- unique(c(
  required_columns,
  "age_bin",
  "age_decade",
  "age_years",
  "age_days_scaled",
  "sex",
  "ethnicity_groups",
  "assay_groups",
  "disease_groups",
  "total_n",
  "plasma_n",
  "prop_plasma_observed",
  "prop_plasma_observed_arcsin_sqrt"
))

sample_lookup <- sce_metadata |>
  select(any_of(lookup_columns)) |>
  rename(cell_compartment = cell_type_unified_ensemble) |>
  distinct(sample_id, cell_compartment, .keep_all = TRUE)

pseudobulk_metadata <- as.data.frame(SummarizedExperiment::colData(pseudobulk_all))
pseudobulk_key <- paste(
  pseudobulk_metadata$sample_id,
  pseudobulk_metadata$cell_compartment,
  sep = "___"
)
lookup_key <- paste(
  sample_lookup$sample_id,
  sample_lookup$cell_compartment,
  sep = "___"
)
lookup_index <- match(pseudobulk_key, lookup_key)
if (anyNA(lookup_index)) {
  stop("Could not match all pseudobulk columns back to sample metadata.")
}

for (column in setdiff(names(sample_lookup), names(pseudobulk_metadata))) {
  SummarizedExperiment::colData(pseudobulk_all)[[column]] <-
    sample_lookup[[column]][lookup_index]
}

pseudobulk_metadata <- as.data.frame(
  SummarizedExperiment::colData(pseudobulk_all)
)
pseudobulk_metadata$cell_type_unified_ensemble <-
  pseudobulk_metadata$cell_compartment
SummarizedExperiment::colData(pseudobulk_all) <-
  S4Vectors::DataFrame(pseudobulk_metadata, row.names = colnames(pseudobulk_all))

sample_tracking[[4]] <- count_samples(
  pseudobulk_metadata,
  "4_pseudobulk_aggregated",
  "4. Pseudobulk aggregated"
)
```

# Separate QC and export

Each tissue-system and cell-compartment combination is filtered and normalized
independently. Pseudobulks require more than the configured minimum number of
contributing cells, low-abundance genes are removed with `edgeR::filterByExpr()`,
and TMMwsp normalization factors and log offsets are stored in `colData()`.

```{r qc-and-export}
combinations <- tidyr::expand_grid(
  tissue_system = selected_systems,
  cell_compartment = selected_compartments
)

counts_assay <- if ("counts" %in% SummarizedExperiment::assayNames(pseudobulk_all)) {
  "counts"
} else {
  SummarizedExperiment::assayNames(pseudobulk_all)[[1]]
}

output_summary_rows <- list()
sample_qc_rows <- list()
minimum_cell_tracking <- list()

for (index in seq_len(nrow(combinations))) {
  tissue_system <- combinations$tissue_system[[index]]
  cell_compartment <- combinations$cell_compartment[[index]]
  column_metadata <- as.data.frame(SummarizedExperiment::colData(pseudobulk_all))
  selected_columns <-
    column_metadata$tissue_system == tissue_system &
    column_metadata$cell_compartment == cell_compartment

  pseudobulk <- pseudobulk_all[, selected_columns, drop = FALSE]
  genes_before <- nrow(pseudobulk)
  samples_before <- ncol(pseudobulk)

  if (ncol(pseudobulk)) {
    keep_columns <- SummarizedExperiment::colData(pseudobulk)$ncells > min_cells
    pseudobulk <- pseudobulk[, keep_columns, drop = FALSE]
  }

  minimum_cell_tracking[[index]] <- tibble(
    step = "5_minimum_cell_qc",
    step_label = paste0("5. More than ", min_cells, " contributing cells"),
    tissue_system = tissue_system,
    cell_compartment = cell_compartment,
    n_samples = ncol(pseudobulk)
  )

  output_file <- file.path(
    output_dir,
    paste0("pseudobulk_qc_", tissue_system, "_", cell_compartment, ".rds")
  )

  if (ncol(pseudobulk) < 2L) {
    output_summary_rows[[index]] <- tibble(
      tissue_system = tissue_system,
      cell_compartment = cell_compartment,
      samples_before_qc = samples_before,
      samples_after_qc = ncol(pseudobulk),
      genes_before_qc = genes_before,
      genes_after_qc = NA_integer_,
      status = "skipped: fewer than two pseudobulks",
      output_file = NA_character_
    )
    next
  }

  counts_matrix <- as.matrix(
    SummarizedExperiment::assay(pseudobulk, counts_assay)
  )
  keep_genes <- edgeR::filterByExpr(counts_matrix)
  if (!any(keep_genes)) {
    output_summary_rows[[index]] <- tibble(
      tissue_system = tissue_system,
      cell_compartment = cell_compartment,
      samples_before_qc = samples_before,
      samples_after_qc = ncol(pseudobulk),
      genes_before_qc = genes_before,
      genes_after_qc = 0L,
      status = "skipped: no genes passed expression QC",
      output_file = NA_character_
    )
    next
  }

  pseudobulk <- pseudobulk[keep_genes, , drop = FALSE]
  filtered_counts <- counts_matrix[keep_genes, , drop = FALSE]
  dge <- edgeR::DGEList(counts = filtered_counts)
  dge <- edgeR::calcNormFactors(dge, method = "TMMwsp")
  effective_library_size <- with(
    dge$samples,
    lib.size * norm.factors
  )

  SummarizedExperiment::colData(pseudobulk)$library_size <-
    dge$samples$lib.size
  SummarizedExperiment::colData(pseudobulk)$normalization_factor <-
    dge$samples$norm.factors
  SummarizedExperiment::colData(pseudobulk)$offset <-
    log(pmax(effective_library_size, 1))

  sample_metadata <- as.data.frame(SummarizedExperiment::colData(pseudobulk))
  sample_qc_rows[[index]] <- tibble(
    tissue_system = tissue_system,
    cell_compartment = cell_compartment,
    sample_id = sample_metadata$sample_id,
    dataset_id = sample_metadata$dataset_id,
    contributing_cells = sample_metadata$ncells,
    library_size = dge$samples$lib.size,
    normalization_factor = dge$samples$norm.factors,
    effective_library_size = effective_library_size,
    zero_fraction = colMeans(filtered_counts == 0)
  )

  saveRDS(pseudobulk, output_file)
  output_summary_rows[[index]] <- tibble(
    tissue_system = tissue_system,
    cell_compartment = cell_compartment,
    samples_before_qc = samples_before,
    samples_after_qc = ncol(pseudobulk),
    genes_before_qc = genes_before,
    genes_after_qc = nrow(pseudobulk),
    status = "written",
    output_file = output_file
  )
}

sample_tracking <- bind_rows(
  bind_rows(sample_tracking),
  bind_rows(minimum_cell_tracking)
)
output_summary <- bind_rows(output_summary_rows)
sample_qc <- bind_rows(sample_qc_rows)

readr::write_csv(
  sample_tracking,
  file.path(output_dir, "pseudobulk_qc_stage_summary.csv")
)
readr::write_csv(
  dataset_entropy,
  file.path(output_dir, "pseudobulk_qc_dataset_entropy.csv")
)
readr::write_csv(
  output_summary,
  file.path(output_dir, "pseudobulk_qc_output_summary.csv")
)
if (nrow(sample_qc)) {
  readr::write_csv(
    sample_qc,
    file.path(output_dir, "pseudobulk_qc_sample_summary.csv")
  )
}

output_summary |>
  knitr::kable(caption = "Generated pseudobulk objects")
```

# Interactive QC

## Sample retention

```{r interactive-sample-retention}
sample_tracking <- sample_tracking |>
  mutate(
    step_number = as.integer(str_extract(step, "^[0-9]+")),
    step_label = factor(
      step_label,
      levels = unique(step_label[order(step_number)])
    )
  )

sample_retention_plot <- sample_tracking |>
  ggplot(
    aes(
      x = step_label,
      y = n_samples,
      colour = cell_compartment,
      group = cell_compartment,
      text = paste0(
        "Tissue system: ", tissue_system,
        "<br>Compartment: ", cell_compartment,
        "<br>QC step: ", step_label,
        "<br>Samples: ", n_samples
      )
    )
  ) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  facet_wrap(~ tissue_system, scales = "free_y") +
  labs(
    title = "Pseudobulk sample retention across QC",
    x = NULL,
    y = "Number of samples",
    colour = "Cell compartment"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

plotly::ggplotly(sample_retention_plot, tooltip = "text")
```

## Library size and sparsity

```{r interactive-library-qc}
if (nrow(sample_qc)) {
  library_qc_plot <- sample_qc |>
    ggplot(
      aes(
        x = effective_library_size,
        y = zero_fraction,
        colour = cell_compartment,
        text = paste0(
          "Sample: ", sample_id,
          "<br>Dataset: ", dataset_id,
          "<br>Tissue system: ", tissue_system,
          "<br>Compartment: ", cell_compartment,
          "<br>Contributing cells: ", contributing_cells,
          "<br>Effective library size: ",
          scales::comma(round(effective_library_size)),
          "<br>Zero fraction: ", scales::percent(zero_fraction, accuracy = 0.1)
        )
      )
    ) +
    geom_point(alpha = 0.75, size = 2) +
    scale_x_log10(labels = scales::label_comma()) +
    scale_y_continuous(labels = scales::label_percent()) +
    facet_wrap(~ tissue_system) +
    labs(
      title = "Library size and sparsity after QC",
      x = "Effective library size (log scale)",
      y = "Zero-expression fraction",
      colour = "Cell compartment"
    ) +
    theme_minimal(base_size = 12)

  plotly::ggplotly(library_qc_plot, tooltip = "text")
} else {
  cat("No tissue-system/cell-compartment combination produced an output.")
}
```

# Output files

The integrated Seurat RDS contains normalized expression, PCA, Harmony and UMAP
representations. The six pseudobulk RDS files contain raw counts after sample
and gene QC. TMMwsp normalization factors and log effective-library-size offsets
are stored in `colData()`; counts are not log-transformed. The accompanying CSV
files provide the complete QC audit trail.

```{r output-files}
list.files(output_dir, full.names = TRUE) |>
  file.info() |>
  rownames_to_column("file") |>
  select(file, size, mtime) |>
  arrange(file) |>
  knitr::kable(caption = "Plasma-niche input files")
```

# Session information

```{r session-information}
sessionInfo()
```
